---
title: "IA006 - QTL - SEC of mutants 31 and 83- Differential Analysis - normalized"
author: "Ilaria"
date: "`r Sys.Date()`"
output:
  # pdf_document:
  #   latex_engine: xelatex
  #   toc: yes
  html_document:
    toc_depth: 3
    number_sections: yes
    theme: united
    toc: yes
    code_folding: hide
    keep_md: true #enable to save all plots in a folder when knitting to html 
fontsize: 11pt
params:
  metabolite: "ATP"   # default; overridden by scripts/run_all_metabolites.R
editor_options: 
  markdown: 
    wrap: 72
---

```{=tex}
letoldShadedShaded
letendoldShadedendShaded
renewenvironment{Shaded}{footnotesizeoldShaded}{endoldShaded}
letoldverbatimverbatim
letendoldverbatimendverbatim
renewenvironment{verbatim}{footnotesizeoldverbatim}{endoldverbatim}
```

<!-- =========================================================================
 CLEANED MAIN PIPELINE (QTL strain31 vs strain83).
 Reorganised from the original on 2026-06-19. What changed vs the original:
   * Inline "fix" functions removed -> sourced from R/ccprofiler_fixes.R.
   * Package chunk uses library() only (installation is handled by setup.R/renv);
     library(CCprofilerDiffAnna) -> library(CCprofiler).
   * Paths use here::here(); inputs read from data/raw/, outputs go to output/.
   * Exploratory / other-project / one-off code moved to analysis/experiments.Rmd.
 The original is preserved verbatim as ORIGINAL_DiffAnalysis_Norm_QTL_31_83_2026.Rmd.
 PROTEIN QUANTIFICATION: the step that builds `protein_traces_list` (missing from
 the original) is now in place in "6. Protein quantification" - it uses the lab
 sibling/replicate peptide-correlation fix (Option A). The report still needs your
 input files in data/raw/ to knit end-to-end.
============================================================================ -->

# Introduction

Yeast 83(15_6_c) and 31(6_3_c) grown and harvested at OD 0.8, 3 biological replicates each.

## CHECKLIST script:

-   [x] give inputfiles path, filename pattern, set conditions names in
    "setup"

-   [x] go through and change accordingly in "analysis SETUP"

-   [x] do you want to rerun protein feature-finding? if yes, delete
    Rdata file from folder.

-   [x] are there outlier fraction? define and select how to handle in
    "analysis SETUP"

-   [ ] check if fraction annotation file chunck is ok

-   [ ] do you want to filter out proteins with less then x precursors?

# 1. SETUP

```{r setup, include=TRUE}

knitr::opts_chunk$set(echo = TRUE)
## language settings
Sys.setenv(LANG="en")

# ==========================================================================
# Ecoli_PCM - metabolite comparisons (reference is ALWAYS ctrl)
# Each metabolite has its own Spectronaut .tsv (that metabolite vs its own ctrl),
# so each binary comparison still has separate raw data.
# To run ONE comparison: set `metabolite` below (or render with params$metabolite).
# To run ALL metabolites in one pass WITHOUT re-knitting by hand, run:
#     source(here::here("scripts", "run_all_metabolites.R"))
# which renders this report once per metabolite. A later cross-metabolite chunk
# can then load the per-metabolite outputs from output/PCM_ctrl_vs_<metabolite>/.
# ==========================================================================
metabolite_files <- list(
  # TODO: one entry per metabolite -> its Spectronaut .tsv (metabolite vs ctrl).
  ATP = "TODO_ctrl_vs_ATP.tsv",
  PEP = "TODO_ctrl_vs_PEP.tsv"
  # , ADP = "TODO_ctrl_vs_ADP.tsv", GTP = "...", NAD = "...", ...  (add all 8)
)
ref_condition_name <- "ctrl"   # reference condition for every comparison

# `metabolite` may be supplied by a parameterized render (params$metabolite);
# otherwise the default below is used (interactive / single knit).
metabolite <- if (exists("params") && !is.null(params$metabolite)) params$metabolite else "ATP"

## ---- everything below is derived (no need to edit) ----
stopifnot(metabolite %in% names(metabolite_files))
comparison_id        <- paste0("PCM_ctrl_vs_", metabolite)
experiment_name      <- comparison_id
data_location        <- metabolite_files[[metabolite]]
condition1           <- ref_condition_name   # ctrl (reference)
condition2           <- metabolite            # treatment / test
calibration_location <- "IA003_01_SEC_calibration_table.xlsx"   # shared by all comparisons
complex_portal_file  <- "Complex_portal_83333.tsv"              # E. coli K-12 (taxid 83333)
trypsin_id           <- "CON_P00761"   # digestion-control protein for the trypsin QC plot
                                   # (yeast DBs: "P00761"; contaminant-prefixed DBs e.g. E. coli: "CON_P00761")
file_name_pattern    <- c("TODO_PCM_filename_prefix_")

# colours: reference vs treatment
color_scheme_treatment <- setNames(c("#caf0f0", "#f6acca"), c(condition1, condition2))
condition_levels       <- c(condition1, condition2)

```

## 1a. Parameters

```{r}
run_qcs_wProtti <- T

perform_normalization_cyclicloess <- F #(now it works :) Benni fixed it)

load_proteinFeatures.RData <- F
load_protein_DiffExprPep_and_Protein <- F
################### ordering the samples?

sample_sort_by_number <- T # if need to reorder samples e.g. based on treatment concentration

################### removing lowest MW in MW calibration curve

remove_lowest_MW <- T

################### defining ref and treatment conditions

ref_condition <- condition1

treatment_conditions <- c(condition2) #,condition3,condition4)

################### aggregate replicate when plotting?

aggregate_rep <- T # choose T if more then 1 rep per condition, F if only 1 rep

################### what to do with the problematic fractions?

run_annotateOutlierFractions <- T # don't want to run this, I'm annotating it manually

problematic_fractions <- c() #c("P2_36", "P2_47", "P2_48") # to annotate manually


################### GRID-SEARCH

perform_prot_grid_search <- T

# With ibmt pval < 5e-3 in the control-control comparison experiment - TODO refine the way we look at this, but good enough for now!
#proteins_to_exclude <- c("P04816", "P07003", "P07118", "P07822", "P08839", "P08997", "P09126", "P0A6C1", "P0A6C5", "P0A6F3", "P0A6H5", "P0A6T5", "P0A6U3", "P0A6W0", "P0A6W9", "P0A725", "P0A759", "P0A780", "P0A7B3", "P0A7G2", "P0A817", "P0A862", "P0A8I1", "P0A8P3", "P0A8V0", "P0A951", "P0A9H1", "P0A9P9", "P0AA16", "P0AAN5", "P0AAR3", "P0AAZ7", "P0AB14", "P0AB43", "P0AB46", "P0AB61", "P0AB65", "P0AB80", "P0ABE2", "P0ABH0", "P0AD53", "P0ADA1", "P0ADE6", "P0ADU5", "P0ADW8", "P0AE08", "P0AE48", "P0AE60", "P0AE63", "P0AEM4", "P0AES6", "P0AES9", "P0AFF6", "P0AFG0", "P0AFH0", "P0AFI7", "P0AFL3", "P0AFM4", "P0AFW0", "P0AFW8", "P0AGG4", "P0C0L9", "P10346", "P10442", "P16700", "P21151", "P21177", "P21889", "P23538", "P23843", "P23878", "P25524", "P26602", "P27278", "P27303", "P28248", "P30176", "P32141", "P32685", "P33219", "P33355", "P36659", "P36979", "P37613", "P37665", "P37691", "P39274", "P39408", "P40191", "P42616", "P45470", "P52119", "P64490", "P64506", "P64519", "P64559", "P68206", "P69222", "P69795", "P69822", "P76141", "P76243", "P76422", "P76423", "P76569", "P77695", "Q2M7X4", "Q46798")

#Additional proteins to keep that upon manual inspection show an opposite effect than what occurs just by time
proteins_to_keep <- c()



```

## 1b. create folders

```{r}

# All outputs for this comparison go under output/<comparison_id>/
out_dir <- here::here("output", comparison_id)

output_files_location <- file.path(out_dir, "output_files")
RData_for_further_plotting_and_analysis_location <- file.path(out_dir, "RData_for_further_plotting_and_analysis")
DistinctAssembly_hits_location <- file.path(out_dir, "DistinctAssembly_hits")
DistinctAssembly_hits_nolog2FC_location <- file.path(out_dir, "DistinctAssembly_hits_nolog2FC")

# Create the comparison's output folders (incl. figures/tables/rdata) if missing
for (d in c(out_dir,
            file.path(out_dir, c("figures", "tables", "rdata")),
            output_files_location,
            RData_for_further_plotting_and_analysis_location,
            DistinctAssembly_hits_location,
            DistinctAssembly_hits_nolog2FC_location)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
cat("Outputs for this run go to: ", out_dir)

```


## 1c. Loading packages and libraries


Bugs on the CCprofiler functions are fixed in CCprofilerDiffAnna, in the PPlabFunctions.R file. 
There the IBMT test, gaussian fitting sliding_window functions and are also saved.

meaning that essential for this analysis are the packages:
** CCprofilerDiffAnna
** PrInCE

```{r, message=FALSE, warning=FALSE, cache=FALSE}
#' # Load CCprofiler package and set working directory
#library(devtools)
library(data.table)
library(stringr)
library(ggplot2)
library(betareg)
library(lmtest)
library(ggrepel)

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("preprocessCore")
# 
# install_github("CCprofiler/CCprofiler", ref = "differential")
###### if the upper part does not work -> use the following: I've downloaded the repository in my downloads folder
#devtools::install_local("/Volumes/biol_bc_picotti_1/Anna/PROJECTS/SEC_metabolites/CCprofiler/CCprofiler-differential", force = TRUE)
#devtools::install_local("W:/PMC_project/CCprofiler/CCprofiler-differential", force = TRUE)
#library(CCprofiler)


#devtools::install_github("AnnaPagotto/CCprofilerDiffAnna", dependencies = TRUE)
#devtools::install_local("W:/PMC_project/CCprofiler/CCprofilerDiffAnna-differential", force = TRUE)
library(CCprofiler)   # was CCprofilerDiffAnna (lab fork); fixes now loaded from R/ccprofiler_fixes.R

library(scales)

########## my usuals

library(protti)
library(tidyverse)
library(DT)

library(readxl) #to read Excel files like .xlsx with the function read_excel()
library(writexl) # to write Excel files

# for overlap plot
#install.packages("UpSetR")
library(UpSetR)

#For Gaussian fitting and probably new complex identification

# if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install("PrInCE")
library(PrInCE)

library(MSnbase)
library(methods)
library(progress)
library(rlist)
library(data.table)
library(ggpmisc)
library(limma)

### additional packages required

# NOTE: package installation is handled once by setup.R / renv (see README.md).
# (The original had install.packages(parallel/foreach/doParallel/igraph) here,
#  which also errored because the package names were unquoted.)

library(parallel)
library(foreach)
library(doParallel)
library(igraph)  # For network-based proteoform detection

```

'here' settings

```{r}
#devtools::install_local("C:/Users/apagotto/Downloads/rprojroot-main/rprojroot-main", dependencies = TRUE)
# install.packages("here")
library(here)
#install.packages("fs")
library(fs)


# List all files and directories recursively

#dir_tree(path = here() , recurse = TRUE)

here()
```


## 1d. Load the CCprofiler fixes

The lab fixes that used to be pasted inline here now live in
`R/ccprofiler_fixes.R` (see that file's header for full provenance). They are
sourced **on top of** the official CCprofiler package, so the corrected versions
take precedence when called from this document.

```{r load-fixes, message=FALSE, warning=FALSE}
source(here::here("R", "ccprofiler_fixes.R"))
```

# 2. DATA preparation

## 2a. Import data

```{r, eval=TRUE}


if(file.exists(file.path(out_dir, "rdata", "data_input.RData"))){
  load(file.path(out_dir, "rdata", "data_input.RData"))
} else {
  data <- read_protti(here("data", "raw", comparison_id, data_location))
  save(data, file = file.path(out_dir, "rdata", "data_input.RData"))
}

head(data, n = 10)
```

## 2b. Fraction annotation table To import the data into the *CCprofiler*
container format, an annotation table mapping the biochemical fraction
numbers to the respective file name is required. It is important that
the fractions start at 1 and consecutively increase by 1 unit until
reaching the total number of fractions. An example annotation file is
available as `exampleFractionAnnotation`.

filename fraction_number heuselm_J130729_001 1

```{r, eval=TRUE}

#unique(data$r_file_name)

fraction_annotation <- data %>%
  separate(r_file_name, c("date", "instrument", "operator", "experiment", "replicate_number", "condition", "fraction_number", "plate_position"), sep="_", remove = FALSE, extra="drop") %>% # get the fraction number
  dplyr::select(r_file_name, condition, fraction_number, replicate_number) %>%
  distinct() %>%
  mutate(r_file_name = str_replace_all(r_file_name, pattern = file_name_pattern, replacement = "")) %>%
  mutate(fraction_number= str_replace_all(fraction_number, pattern = "f", replacement = "")) %>%
  mutate(replicate_number= str_replace_all(replicate_number, pattern = "biolrep", replacement = ""))
  

head(fraction_annotation, n = 20)
```

## 2c. MW calibration In some co-fractionation techniques, such as
size-exclusion chromatography, a log-linear relationship between the
fraction dimension and molecular weight (MW) can be assumed. In such
cases a MW calibration is possible by leveraging the elution information
of standard proteins with known MWs run separately on the fractionation
device. The inferred apparent molecular weight can then be used to judge
a protein’s likely assembly state given its monomeric molecular weight
and observed elution fraction (~ apparent molecular weight). Monomer
weights can be added to the data by the annotation functions (see
below).

```{r, eval=TRUE, fig.show='hold'}

calibrationTable <- read_excel(here("data", "raw", calibration_location)) %>%
  as.data.table() 

if(remove_lowest_MW == T) {calibrationTable <- calibrationTable %>%
  filter(std_weights_kDa != 0.244)}


calibration = calibrateMW(calibrationTable)
```

## 2d. FILTER data

Filtering for intensity > 64 (log2(I)>6)

```{r}
data_filter <- data %>%
  filter(pep_is_proteotypic == T) %>% # remove non proteotypic peptides
   mutate(fg_quantity_log2 = log2(fg_quantity)) %>% 
  filter(fg_quantity >= 50) %>% # only include intensities above log2 value of 6 (= fg_quantity of 64). Lower intensities are likely false assignments. This is for the Astral: we are being less stringent than with the other instruments, where we use a cutoff of fg_quantity_log2>10 
  filter(pg_protein_accessions != "iRT-Kit_WR_fusion") %>%
  dplyr::select(r_file_name, pg_protein_accessions, pep_grouping_key, fg_quantity, pep_stripped_sequence, pep_nr_of_missed_cleavages)%>%
  mutate(r_file_name = str_replace_all(r_file_name, pattern = file_name_pattern, replacement = "")) #%>%
  #rename(pg_precursor_id = pep_grouping_key)
  
```

# 3. QCs with Protti

```{r}
if(run_qcs_wProtti == T){
  qc <- data_filter %>%
  left_join(fraction_annotation)
}
```

## 3a. Intensity distribution before filtering 

```{r}
data %>%
  mutate(fg_quantity_log2 = log2(fg_quantity)) %>%
  qc_intensity_distribution(
    grouping = pep_grouping_key,
    intensity_log2 = fg_quantity_log2,
    plot_style = "histogram"
  ) +
  geom_vline(xintercept = 5.64, color = "red", linewidth = 0.3) + #log2(50) = 5.643856
  labs(subtitle = "Red line = cutoff for removal of low intensities") 
```



## 3b. IDs

```{r ids_per_frac, fig.height=7, fig.width=15, dev='pdf'}
if(run_qcs_wProtti == T){

###### Pep (precursor) Ids
  
  for (i in unique(qc$condition)) {
    
      plot <- qc%>%
        filter(condition==i) %>%
        qc_ids(, 
             sample = r_file_name,
             grouping = pep_grouping_key,
             intensity = fg_quantity, # precursor lvl
             condition = condition,
             title = paste0("Number of peptide IDs per sample_", i),
             interactive = FALSE) +
        theme(legend.position = "bottom") +
        ggplot2::scale_fill_manual(values = color_scheme_treatment) 
      
      print(plot)
    }

pep_ids <- qc_ids(qc, 
       sample = r_file_name,
       grouping = pep_grouping_key,
       intensity = fg_quantity, # precursor lvl
       condition = condition,
       plot = FALSE)

head(pep_ids)


###### Prot ids

    for (i in unique(qc$condition)) {
        
        plot <- qc%>%
          filter(condition==i) %>%
          qc_ids(
             sample = r_file_name,
             grouping = pg_protein_accessions,
             intensity = fg_quantity,
             condition = condition,
             title = paste0("Number of protein IDs per sample_", i),
             interactive = FALSE) +
        theme(legend.position = "bottom") +
        ggplot2::scale_fill_manual(values = color_scheme_treatment)
        
        print(plot)
    }

prot_ids <- qc_ids(qc, 
       sample = r_file_name,
       grouping = pg_protein_accessions,
       intensity = fg_quantity,
       condition = condition,
       plot = FALSE)

head(prot_ids)
}
```

```{r total_prot_IDs}
if(run_qcs_wProtti == T){

###### Protein Ids

      plot <- qc%>%
        qc_ids( 
             sample = condition,
             grouping = pg_protein_accessions,
             intensity = fg_quantity, # precursor lvl
             condition = condition,
             title = "Number of protein IDs per condition",
             interactive = FALSE) +
        theme(legend.position = "bottom") +
        ggplot2::scale_fill_manual(values = color_scheme_treatment) 
      
      print(plot)
      
      prot_ids_all <- qc_ids(qc, 
       sample = condition,
       grouping = pg_protein_accessions,
       intensity = fg_quantity,
       condition = condition,
       plot = FALSE)
      
      prot_ids_all
    }
```

## 3c. missed cleavages

```{r qc_missed_cleavages, fig.height=7, fig.width=15, dev='pdf'}

if(run_qcs_wProtti == T){
  
  for (i in unique(qc$condition)) {
    
    plot <- qc%>%
          filter(condition==i) %>%
      qc_missed_cleavages(
        sample = r_file_name,
        grouping = pep_stripped_sequence,
        missed_cleavages = pep_nr_of_missed_cleavages,
        intensity = fg_quantity,
        method = "intensity",
        plot = T
      ) +
      labs(subtitle = i)
    
    print(plot)}
}
```

# 4. CC PROFILER differential analysis workflow

## 4a. create DESIGN MATRIX (w/ fraction_annotation)

```{r design_matrix}

#fix colnames and as.data.table

fraction_annotation_table <- fraction_annotation %>%
  dplyr::rename("filename" = "r_file_name") %>%
  mutate(fraction_number = as.numeric(fraction_number)) %>%
as.data.table() %>%
  arrange(fraction_number) %>% ################################# it is important to reorder the fraction_annotation file because otehrwise it gives you an error when running the importPCPdata() function
  dplyr::rename("condition_id" = "condition") %>%
  dplyr::rename("replicate_id" = "replicate_number") %>% ##just to have the same names as Isabell...
  mutate(sample= paste0(condition_id, "_", replicate_id)) 
 
  

fraction_annotation_sub <- fraction_annotation_table %>%
  dplyr::select(c(filename,fraction_number,sample)) %>%
  distinct()
  

 
## create design matrx
  
design_matrix <- unique(subset(fraction_annotation_table, select=c("condition_id","replicate_id","sample")))

setnames(design_matrix, c("condition_id","replicate_id","sample"), c("Condition","Replicate","Sample_name"))

design_matrix$Condition <- factor(design_matrix$Condition, levels= condition_levels)

# Safety net: factor() above turns Condition into NA if the registry ref/treat
# don't match the labels parsed from the data. Fail fast with a clear message
# instead (this mismatch is what previously cascaded into the q-value / hist
# errors much further downstream).
if (anyNA(design_matrix$Condition)) {
  stop("design_matrix$Condition has NA: the comparison's ref/treat (",
       paste(condition_levels, collapse = ", "),
       ") do not match the condition labels in the data. Labels found: ",
       paste(sort(unique(as.character(fraction_annotation_table$condition_id))), collapse = ", "),
       ". Edit ref/treat in the SETUP comparison registry to match.")
}


```

## 4b. fix data_filter for CCprofiler

```{r data_filter_ccp_mod}

data_filter_ccp_mod <- data_filter %>%
  filter(!grepl("CON_", pg_protein_accessions) | pg_protein_accessions == trypsin_id) %>% #remove contaminants, except for trypsin (trypsin_id in SETUP)
dplyr::rename ("filename" = "r_file_name", "protein_id" = "pg_protein_accessions", "peptide_id"="pep_grouping_key", "intensity"="fg_quantity") %>%
 left_join(fraction_annotation_table) %>%
  group_by(filename, protein_id) %>% ### FILTER OUT ALL THE PROTEIN WITH LESS THAN 3 eg_precursor_id?
add_count(protein_id, name = "n_precursors") %>%# get the n precursors per protein per sample
   dplyr::select(protein_id, peptide_id, filename, intensity)%>%
as.data.table()

```

## 4c. create TRACE ANNOTATION table

```{r Fetch_uniprot for traceAnnotation df}
uniprot <- fetch_uniprot(unique(data_filter_ccp_mod$protein_id),
                         columns = c("protein_name", "length", "sequence", "mass", "gene_names","organism_name","xref_geneid",
    "xref_string", "go_f", "go_p", "go_c", "cc_interaction", "ft_act_site", "ft_binding",
    "cc_cofactor", "cc_catalytic_activity", "xref_pdb"),
    batchsize = 50) 
```

```{r traceAnnotation_table}
## Trace annotation table

## information from uniprot
traceAnnotation <- as.data.table(uniprot) %>%
  dplyr::rename("Mass" ="mass") # there is a bug in the annotateTraces function, it doesn't add the protein_mw column if the mass column is "mass" and not "Mass". Tried to fix the function but not fully managed: it does add the column, but put NAs as values....

head(traceAnnotation, n = 2)
```

## 4d. Get traces list

```{r pepTraces_raw}

samples <- unique(fraction_annotation_table$sample)


# to order by number in the string
if(sample_sort_by_number == T) {
samples <- str_sort(samples, numeric = TRUE)}


traces_list <- lapply(samples,function(x){
  message(x)
  ann <- subset(fraction_annotation_sub, sample==x) 
  ann <- subset(ann, filename %in% data_filter_ccp_mod$filename) 
  ann <- subset(ann, select = c("filename","fraction_number")) 
  setkey(ann,fraction_number) 
  data_sub_select <- subset(data_filter_ccp_mod,filename %in% ann$filename) 
  data_sub_select <- merge(data_sub_select,ann, by=c("filename")) 
  setkey(data_sub_select,fraction_number) 
  data_sub_select[,fraction_number:=NULL] 
  traces <- importPCPdata(input_data=data_sub_select,fraction_annotation=ann)
  return(traces)
})


names(traces_list) = samples
class(traces_list) <- "tracesList"


#' # Annotate traces with information from uniprot
 
pepTraces_raw <- annotateTraces(traces= traces_list,
                            trace_annotation= traceAnnotation,
                            traces_id_column = "protein_id",
                            trace_annotation_id_column = "input_id",
                            #trace_annotation_mass_column = "Mass",
                            uniprot_mass_format = TRUE,
                            replace_whitespace = TRUE)

#' # Annotate traces with molecular weight calibration
pepTraces_raw <- annotateMolecularWeight(pepTraces_raw,
                                     calibration)

summary(pepTraces_raw)
```


# 5. QC with CCprofiler

## 5a. alignment

```{r align across all samples}

#' # Assess how well SEC traces align across all samples
alignTraces(pepTraces_raw, min_lag = -3, max_lag = 3, plot = T, PDF=F) #to save as PDF, set PDF=T
#' # Assess total intensity as a proxi for extraction efficiency across samples
plotGlobalIntensities(pepTraces_raw, plot = T, PDF=F)
```

## 5b. annotateOutlierFractions & impute

```{r}
if(run_annotateOutlierFractions == T) {
pepTracesOutlFrac <- annotateOutlierFractions(traces=pepTraces_raw,
                                              excludeBoarder = T,
                                              plot = F)

annotateOutlierFractions(traces=pepTraces_raw,
                         excludeBoarder = T,
                         plot = TRUE,
                         PDF = FALSE,
                         name = "OutlierFractions")
}
```

```{r}

if(run_annotateOutlierFractions == F) {
pepTracesOutlFrac <- lapply(pepTraces_raw, function(lst) {
  # Add the isOutlier column to the fraction_annotation data frame
    # Check if 'filename' is in problematic_fractions and create the isOutlier column
  lst$fraction_annotation$isOutlier <- lst$fraction_annotation$filename %in% problematic_fractions
  
  return(lst)
})

# Reassign the class to "tracesList"
class(pepTracesOutlFrac) <- class(pepTraces_raw)

# Check the class of the output
class(pepTracesOutlFrac)
}
  
```

```{r}
### NB NB NB, use mean, not spline, mean is much better, Benni said :) But there is a bug in the code -> even if you set 'mean', it will go for spline anyway -> fix the bug (Benni already did...)

pepTracesOutlFracImp <- imputeOutlierFractions( traces = pepTracesOutlFrac,
                                                bound_left = 2,
                                                bound_right = 2,
                                                consider_borders = TRUE,
                                                method = "spline" # mean is also an option
)
```

## 5c. general imputation

```{r Find missing values}

  
#' Find missing values
#' (defined as having identifications in left and right neigbouring fractions):
pepTracesMV <- findMissingValues(pepTracesOutlFracImp,
                                 bound_left = 1,
                                 bound_right = 1,
                                 consider_borders = TRUE)

#' Impute NA values by fitting a spline:
pepTracesImp <- imputeMissingVals(pepTracesMV, method = "spline")

#' Plot imputation summary:
plotImputationSummary(pepTracesMV, pepTracesImp, PDF = F,
                      plot_traces = T, max_n_traces = 2)

```



## 5d. normalization

```{r Normalize intensity values across samples}

if(perform_normalization_cyclicloess==T) {
# Normalize intensity values across samples
pepTracesNormalized <- normalizeByCyclicLoess(pepTracesImp, window = 3, step = 1, plot = F) 

normalizeByCyclicLoess(pepTracesImp, window = 3, step = 1, plot = T, PDF = F) #, name = "normalizeByCyclicLoess")

#' # Assess total intensity after normalization
plotGlobalIntensities(pepTracesNormalized, plot = T, PDF=F ) #, name = "IntensitySummary_postNormalization")
}


if(perform_normalization_cyclicloess==F) { pepTracesNormalized <- pepTracesImp}

```

### -Trypsin plot for QC before  and after normalization
```{r}
# Trypsin (digestion control) ID differs per dataset (set as trypsin_id in SETUP).
# Resolve robustly (try trypsin_id, then its CON_/non-CON_ variant) and skip with a
# message if absent, instead of crashing on an empty traces subset.
.tryp_ids <- unique(unlist(lapply(pepTraces_raw, function(t) t$trace_annotation$protein_id)))
.tryp <- intersect(c(trypsin_id, sub("^CON_", "", trypsin_id), paste0("CON_", trypsin_id)), .tryp_ids)
if (length(.tryp) == 0) {
  message("Trypsin control '", trypsin_id, "' not found - skipping the trypsin QC plot.")
} else {
  .tryp <- .tryp[1]
  plot(subset(pepTraces_raw,       trace_subset_ids = .tryp, trace_subset_type = "protein_id"),
       legend = F, design_matrix = design_matrix, name = "Trypsin non-normalized")
  plot(subset(pepTracesNormalized, trace_subset_ids = .tryp, trace_subset_type = "protein_id"),
       legend = F, design_matrix = design_matrix, name = "Trypsin normalized")
}
```


> **Moved:** the exploratory *trypsin-ratio normalization* that was here has been
> moved to `analysis/experiments.Rmd`. It was an alternative to the normalization
> above and is **not** used by the main pipeline, which continues from
> `pepTracesNormalized`.

## - Pep-lvl filtering

```{r}
#' # Filter by consecutive IDs --> comment out: and peptide correlation--> work with pepTracesConsIds instead of pepTracesSPC
#'
# (CCprofiler + fixes are already loaded in section 1; no need to reload here)

# Combine all traces for filtering:

traces_combined <- combineTracesMutiCond(pepTracesNormalized)
# Filter by consecutive IDs:
pepTracesConsIds <- filterConsecutiveIdStretches(traces_combined,
                                                 min_stretch_length = 3,
                                                 remove_empty = T)

# Filter by maximum correlation for outlier removal:
pepTracesMaxCorr <- filterByMaxCorr(pepTracesConsIds,
                                cutoff = 0.8,
                                plot = T, PDF = F)

# Filter by consecutive SPC:
# pepTracesSPC <- filterBySibPepCorr(pepTracesMaxCorr,
#                                    absolute_spcCutoff = 0.6,
#                                    plot = T, PDF = F)
```

```{r}
#' # Subset pepTracesNormalized to valid peptides selected 
#' by consecutive filtering and SPC cutoff
validPeps <- unique(pepTracesMaxCorr$trace_annotation$id)
pepTracesList_filtered <- lapply(pepTracesNormalized, function(x){
  sub_trace <- subset(x, trace_subset_ids=validPeps)
  sub_trace$trace_annotation <- sub_trace$trace_annotation# %>% #left_join(select(pepTracesMaxCorr$trace_annotation, id #SibPepCorr), by=join_by(id)) # Inserts the sibling peptide correlation data
 sub_trace
})
class(pepTracesList_filtered) <- "tracesList"

#' Update traces with additional metrics for each fraction:
pepTracesList_filtered <- updateTraces(pepTracesList_filtered)

#' Inspect traces list:
summary(pepTracesList_filtered)

```



## - Global stats

```{r}
#' # Global stats
traces_names <- c(
  "pepTraces_raw",
  "pepTracesList_filtered"
)

countDT <- data.table(evidence=character(), sample=character(), count=integer(), traces=character())

for (traces in traces_names) {
  #traces <- readRDS(paste0(traces_name,".rda"))
  traces_name <- traces
  traces <- eval(as.name(traces))

  getIDs <- lapply(traces, function(x){
    peptides <- unique(x$trace_annotation[grep("DECOY",protein_id, invert = T)]$id)
    proteins <- unique(x$trace_annotation[grep("DECOY",protein_id, invert = T)]$protein_id)
    list(peptides=length(peptides),proteins=length(proteins))
  })

  count_dt <- as.data.table(getIDs)
  count_dt[, evidence := c("peptides", "proteins")]
  count_dt <- melt(count_dt, id.vars = "evidence", variable.name = "sample", value.name = "count")
  count_dt[,traces:=traces_name]
  count_dt[, count := unlist(count)]
  countDT <- rbind(countDT,count_dt)
}

traces_names_dt <- data.table(
  traces=c("pepTraces_raw",
           "pepTracesList_filtered"),
  traces_name=c("raw",
                "filtered")
)

countDT <- merge(countDT, traces_names_dt, by="traces")

countDT$traces_name <- factor(countDT$traces_name, levels=traces_names_dt$traces_name)

# write.table(countDT, "traces_stat_counts.txt", sep="t",quote = F, col.names = T, row.names = F)
# 
# pdf("traces_stat_counts.pdf",height=9,width=12)

g <- ggplot(countDT,aes(x=traces_name,y=count,fill=evidence, group=sample)) +
  geom_bar(stat="identity") +
     geom_label(label = countDT$count) +
  facet_wrap(sample ~ evidence, scales="free",nrow=3) +
  scale_fill_manual(values=c("#E69F00", "#56B4E9"))+
  theme_classic() +
  theme(legend.position="bottom") + theme(legend.title = element_blank()) +
  labs(y = "count") +
  theme(axis.title.x=element_blank()) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

print(g)


```


# 6. Protein quantification

> **Protein quantification.** Builds the protein-level `tracesList` from the
> filtered peptide traces using the lab "sibling/replicate peptide-correlation"
> fix (`proteinQuantification_sibPepCorrFix.tracesList`, in
> `R/ccprofiler_fixes.R`). This is the step that was missing from the original
> `.Rmd`; **Option A** is the confirmed one, **Option B** is a stock-CCprofiler
> alternative kept commented for reference.

```{r protein-quantification, message=FALSE, warning=FALSE}
# Option A (used): lab sibling/replicate peptide-correlation fix
# (defined in R/ccprofiler_fixes.R). It already returns an object of class "tracesList".
# NOTE: CCprofiler prints one BENIGN warning per protein that has >1 annotation row
# ("multiple entries for protein_id ...: . Picking the first") - it only affects
# which metadata row is kept, not the quantification. suppressWarnings() keeps it
# from flooding the console/report.
protein_traces_list <- suppressWarnings(proteinQuantification_sibPepCorrFix.tracesList(
  pepTracesList_filtered,
  topN       = 2,
  keep_less  = FALSE,
  rm_decoys  = TRUE,
  quantLevel = "protein_id"
))

# Option B (alternative): stock CCprofiler protein quantification. Uncomment to use.
# protein_traces_list <- proteinQuantification(pepTracesList_filtered, topN = 2)
# class(protein_traces_list) <- "tracesList"
```

# GLOBAL assembly state

## - Protein Mass distribution

```{r}

# evaluate protein mass distribution
protTraces_assembly <- annotateMassDistribution(protein_traces_list)


uniqueProteins <- Reduce(intersect, lapply(protTraces_assembly, function(x){x$trace_annotation$id}))

protTraces_assembly <- subset(protTraces_assembly, trace_subset_ids = uniqueProteins)

lapply(names(protTraces_assembly), function(x){
  summarizeMassDistribution(protTraces_assembly[[x]], 
                            PDF=F) #,
                            #name=paste0(here("output_files/massDistribution_",x)))
  })

```

## - Get assembly changes

```{r}
n_conditions <- length(unique(design_matrix$Condition))

#create a list of vectors to save the all_proteins_independentOfAssemblyState; proteins_noAssemblyChanges; more_assembled_proteins; less_assembled_proteins; proteins_withAssemblyChanges

all_proteins_independentOfAssemblyState_list <- list()
proteins_noAssemblyChange_list <- list()
more_assembled_proteins_list <- list()
less_assembled_proteins_list <- list()
proteins_withAssemblyChanges_list <- list()
  

# compare the treatmets to the ref (the list is specified in the vectors treatment_conditions and ref_condition in the Analysis SETUP section)

diffAssemblyState <- getMassAssemblyChange_aljazfix(tracesList=protTraces_assembly,    
                                             design_matrix= design_matrix, 
                                             compare_between = "Condition",
                                             quantLevel = "protein_id",
                                             plot=T, 
                                             PDF=F #,
                                             #name = paste0("beta_pvalue_histogram_"ref_condition,"_",i)
                                             )

```

## - look-up traces of proteins WITH global assembly changes

LET'S have a look at these proteins that change in assembly in at least
one of the treatments:

```{r}
 # lookup all the proteins_withAssembly Changes in any of the treatment conditions


for(i in filter(diffAssemblyState, betaPval_BHadj < 0.05 & abs(meanDiff) > 0.1)$protein_id){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i)
} #not saved in the nas

```

```{r filter not by pval, but by the strength of the difference}
for(i in filter(diffAssemblyState, abs(meanDiff) > 0.1)$protein_id){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i)
}
```


actually, the other way around => fix this by inverting the factor
order for ATP and ctrl. Proteins that are LESS assembled

```{r}

#get all the proteins in the list
all_elements_less_assembled_proteinslist <- unique(unlist(less_assembled_proteins_list))

# lookup all the proteins_withAssembly Changes in any of the treatment conditions

for(i in all_elements_less_assembled_proteinslist){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i) #not saved in the nas
}

```

Proteins that are MORE assembled

```{r}

#get all the proteins in the list
all_elements_more_assembled_proteinslist <- unique(unlist(more_assembled_proteins_list))

# lookup all the proteins_withAssembly Changes in any of the treatment conditions

for(i in all_elements_more_assembled_proteinslist){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i) #not saved in the nas
}

```

`

# DISTINCT assembly state ~ PROTEIN level analysis

### - filter for prot with min N pep

```{r}
## subset to keep only the proteins with at least 5 precursors 
pepTracesList_filtered_minN <- subset(pepTracesList_filtered, trace_subset_ids = uniqueProteins, trace_subset_type = "protein_id")
```

## - performProteinGridSearch

```{r perform_prot_grid_search}

if(perform_prot_grid_search == T) {
# Calculate the number of unique protein IDs to select (10% of total)
num_ids <- ceiling(length(unique(pepTracesList_filtered_minN$ctrl_1$trace_annotation$protein_id)) * 0.1)

# Randomly sample 10% of unique protein IDs
subset_ids <- sample(unique(pepTracesList_filtered_minN$ctrl_1$trace_annotation$protein_id), num_ids)

grid_peptideTraces <- subset(pepTracesList_filtered_minN,
                        trace_subset_ids = subset_ids,
                        trace_subset_type = "protein_id")

# performProteinGridSearch doesn't work with Traces List -> only work on the control list:

grid_peptideTraces <- grid_peptideTraces$ctrl_1


# Perform a small grid search for 2 parameter combinations
 # Depending on the computational resources this can take several minutes


   proteinFeatures_grid <- performProteinGridSearch(traces = grid_peptideTraces,
                                      corrs = c(0.75, 0.9, 0.95), #corr_cutoff, 0.9 for Isabel. #tried 0.5, not good 
                                      windows = c(5,7,10,12), #window_size, 7 for Isabel
                                      smoothing = c(7,9,11), #smoothing_length, 7 for Isabel 
                                      rt_heights = c(1,4,5), #rt_height, 1 for Isabel
                                      n_cores = 5) #n_cores = 3, for Isabel


 lapply(proteinFeatures_grid, head, n = 2) }
```

```{r proteinFeatures_bestStats, eval=FALSE}
if(perform_prot_grid_search == T) {
proteinFeatures_scoredData <- lapply(proteinFeatures_grid, calculateCoelutionScore)
proteinFeatures_qvalueData <- lapply(proteinFeatures_scoredData, calculateQvalue, plot=F)
proteinFeatures_stats <- qvaluePositivesPlotGrid(proteinFeatures_qvalueData)
proteinFeatures_bestStats <- getBestQvalueParameters(proteinFeatures_stats, FDR_cutoff = 0.05)
proteinFeatures_bestStats

save(proteinFeatures_grid,proteinFeatures_bestStats, file = "ProtGridSearch.RData")}

if(perform_prot_grid_search == F) {load(file = "ProtGridSearch.RData")}
#if(perform_prot_grid_search == F) {load(file = "ProtGridSearch_fromJP02.RData")} # this one is to use in case you don't want to run the gridsearch at all

```

## - findProteinFeatures

```{r}
#' # Integrate peptide traces across conditions 
# per each oe 
pepTracesIntegrated_sum <- integrateTraceIntensities(pepTracesList_filtered_minN,
                                                     design_matrix = NULL,
                                                     integrate_within = NULL,
                                                     aggr_fun = "sum")

#' # Perform protein-centric analysis
# Suppress messages for clarity

suppressMessages(
  proteinFeatures <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff = 0.85, # 0.9,
                                          window_size = 7,# 7,
                                          parallelized = T,
                                          n_cores = 6,
                                          collapse_method = "apex_only",
                                          perturb_cutoff = "5%",
                                          rt_height = 1, #1,
                                          smoothing_length = 3 #7)

)
)

suppressMessages(
  proteinFeatures_0p7_corr_cutoff <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff = 0.7, # 0.9,
                                          window_size = 7,# 7,
                                          parallelized = T,
                                          n_cores = 6,
                                          collapse_method = "apex_only",
                                          perturb_cutoff = "5%",
                                          rt_height = 1, #1,
                                          smoothing_length = 3 #7)

)
)

suppressMessages(
  proteinFeatures_0p5_corr_cutoff_winsize5 <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff = 0.5, # 0.9,
                                          window_size = 5,# 7,
                                          parallelized = T,
                                          n_cores = 6,
                                          collapse_method = "apex_only",
                                          perturb_cutoff = "5%",
                                          rt_height = 1, #1,
                                          smoothing_length = 3 #7)

)
)

suppressMessages(
  proteinFeatures_0p4_corr_cutoff_winsize5 <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff = 0.4, # 0.9,
                                          window_size = 5,# 7,
                                          parallelized = T,
                                          n_cores = 6,
                                          collapse_method = "apex_only",
                                          perturb_cutoff = "5%",
                                          rt_height = 1, #1,
                                          smoothing_length = 3 #7)

)
)


save(proteinFeatures, file = "proteinFeatures.RData")

#' Score protein features and filter for a detection FDR of 5%
proteinFeaturesFiltered <- scoreFeatures(proteinFeatures, FDR=0.05, PDF=TRUE, name="output_files/qvalueStats_proteinFeatures")

#' Evaluate protein feature stats
summarizeFeatures(proteinFeaturesFiltered,
                  plot=TRUE,
                  PDF = F)
                  # PDF=TRUE,
                  # name="output_files/feature_summary_proteinFeaturesFiltered")

#' ## Extract feature values
proteinFeaturesFiltered_featureVals <- extractFeatureVals(traces = pepTracesList_filtered_minN,
                                             features = proteinFeaturesFiltered,
                                             design_matrix = design_matrix,
                                             extract = "subunits_detected",
                                             imputeZero = T,
                                             verbose = F)


#' ## Fill feature values
proteinFeaturesFiltered_featureValsFilled <- fillFeatureVals(featureVals = proteinFeaturesFiltered_featureVals,
                                                tracesList = pepTracesList_filtered_minN,
                                                design_matrix = design_matrix)

proteinFeaturesFiltered_0p15featurefdr <- scoreFeatures(proteinFeatures, FDR=0.15, PDF=TRUE, name="output_files/qvalueStats_proteinFeatures")

#' Evaluate protein feature stats
summarizeFeatures(proteinFeaturesFiltered_0p15featurefdr,
                  plot=TRUE,
                  PDF = F)
                  # PDF=TRUE,
                  # name="output_files/feature_summary_proteinFeaturesFiltered")

#' ## Extract feature values
proteinFeaturesFiltered_featureVals_0p15featurefdr <- extractFeatureVals(traces = pepTracesList_filtered_minN,
                                             features = proteinFeaturesFiltered_0p15featurefdr,
                                             design_matrix = design_matrix,
                                             extract = "subunits_detected",
                                             imputeZero = T,
                                             verbose = F)


#' ## Fill feature values
proteinFeaturesFiltered_featureValsFilled_0p15featurefdr <- fillFeatureVals(featureVals = proteinFeaturesFiltered_featureVals_0p15featurefdr,
                                                tracesList = pepTracesList_filtered_minN,
                                                design_matrix = design_matrix)

# save(proteinFeaturesFiltered_featureVals, proteinFeaturesFiltered_featureValsFilled, file = "Anna_proteinFeaturesFiltered_featureValsFilled.RData")


```

## - differential analysis

2 cond, 1 rep function

" Differential analysis is performed in 5 steps: (i)
Peptide-levelintensities are computed for each protein feature and
sample.Missing values in single fractions, replicates, or conditions
areimputed by uniformly sampling values between zero and theminimum
detected signal of a peptide. The peptide intensity ofone feature is
then calculated by summing the intensities of allfractions across the
corresponding protein feature range. (ii)The mean intensity across all
replicates within a condition(specified by the design matrix) is
calculated. (iii) The log2-fold-change between conditions is calculated
based on themean feature intensities. (iv) If replicates are available,
p-values are estimated by comparing the summed intensities across
conditions by a non paired t test. If no replicates are
available,p-values are estimated by comparing each fraction within a
feature by a paired t test across the conditions. (v) To subsequently
derive protein-level information, the peptide-level tests are aggregated
as follows: (1) protein log2-fold-changesare derived from the median
log-2-fold change across alldetected peptides of the protein, and (2)
protein p-values are estimated by determining the fold-change adjusted
median p-value and applying a beta distribution as described by Teo
etal.43 and Suomi et al.44 (for details see the methods section). (vi)
Multiple testing correction is performed by Benjamini-Hochberg
adjustment of the protein-level p-values

```{r diff}
          
            #' ## Perform peptide-level differential expression testing for all features
          
            protein_DiffExprPep <- testDifferentialExpression_beniFix(featureVals = proteinFeaturesFiltered_featureValsFilled,
                                                              compare_between = "Condition",
                                                              level = "peptide",
                                                              measuredOnly = FALSE) #also considers the imputed fractions
            # Append the df to the list

          
            
            #' ## Aggregate differential expression results to the protein level
            protein_DiffExprProtein <- aggregatePeptideTests(protein_DiffExprPep) 
            
            protein_DiffExprPep_0p15featurefdr <- testDifferentialExpression_beniFix(featureVals = proteinFeaturesFiltered_featureValsFilled_0p15featurefdr,
                                                              compare_between = "Condition",
                                                              level = "peptide",
                                                              measuredOnly = FALSE) #also considers the imputed fractions
            # Append the df to the list

          
            
            #' ## Aggregate differential expression results to the protein level
            protein_DiffExprProtein_0p15featurefdr <- aggregatePeptideTests(protein_DiffExprPep_0p15featurefdr) 
            
protein_DiffExprProtein <- protein_DiffExprProtein %>% left_join(select(uniprot, accession, protein_name, mass, gene_names, go_f, go_p, go_c, ft_act_site, ft_binding, cc_cofactor), by=join_by(feature_id == accession))

protein_DiffExprProtein_0p15featurefdr <- protein_DiffExprProtein_0p15featurefdr %>% left_join(select(uniprot, accession, protein_name, mass, gene_names, go_f, go_p, go_c, ft_act_site, ft_binding, cc_cofactor), by=join_by(feature_id == accession))
          
save(protein_DiffExprPep, file = "protein_DiffExprPep_list.RData")
save(protein_DiffExprProtein, file = "protein_DiffExprProtein_list.RData")
save.image()

```

Global differential assessment:

In addition to the feature-specific differential analysis, global
differential assessment is performed by comparing integrated intensities
across the entire fractionation dimension, instead of restricting the
analysis to a feature-specific range. The same strategies as for
feature-specific estimation of log2-fold-changes and p-values are
performed. To assess whether the signal within a protein feature is
changing because of a global change in the protein’s expression or due
to a rearrangement of the proteins relative distribution across
different assembly states,an additional analysis step is available in
CCprofiler. Here, the relative feature-specific mass fraction (FMF) is
estimated by the fraction of the detected MS signal in the
feature-specific mass range relative to the total detected signal: FMF =
sum(intensity features)/sum(intensity global) A change in FMF is
subsequently estimated by the difference in the mean FMF across
conditions diffFMF= mean(FMF condA) - mean (FMF condB)

... check the paper for full explanation of the steps
<https://pubs.acs.org/doi/epdf/10.1021/acs.jproteome.3c00125?ref=article_openPDF>

A significant change in the FMF across conditions indicates that the
protein’s relative contribution to different distinct assembly states
has changed across the conditions, thus providing insights into protein
rewiring, which is not observable by global proteome analyses. In
contrast to complex-centric analyses, described in the following
section, protein-centric differential analysis enables the assessment of
changes in distinct protein assembly states independent of the actual
knowledge of the protein’s exact interaction partners

## - volcano plots and save lists

```{r volcano plots}
library(data.table)

#create unique df - keep as data.table
combined_protein_DiffExprPep_list <- unique(protein_DiffExprPep)

combined_protein_DiffExprProtein_list <- unique(protein_DiffExprProtein)

comparison_level <- unique(combined_protein_DiffExprPep_list$testOrder) 

# create vector list to append to
diffProteins_list <- list()
diffProteins_onlypadj_list <- list()
diffProteins_global_list <- list()

#' # Make volcano plots
for (i in comparison_level) {
  
  # Keep as data.table for plotVolcano compatibility
  protein_DiffExprPep_i <- combined_protein_DiffExprPep_list[testOrder == i]
  protein_DiffExprProtein_i <- combined_protein_DiffExprProtein_list[testOrder == i]
  
  #' General volcano plots - local
  plotVolcano(protein_DiffExprPep_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2,  
              name = paste0("output_files/pep_", i))
  plotVolcano(protein_DiffExprProtein_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              name = paste0("output_files/prot_", i))
  plotVolcano(protein_DiffExprProtein_i, PDF = F, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              name = paste0("prot_", i)) #to see here
  
  #' global
  plotVolcano(protein_DiffExprPep_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              level = "global", name = paste0("output_files/global_pep_", i))
  plotVolcano(protein_DiffExprProtein_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              level = "global", name = paste0("output_files/global_prot_", i))
  
  #' Get different proteins - using data.table syntax
  diffProteins_i <- unique(protein_DiffExprProtein_i[pBHadj < 0.05 & abs(medianLog2FC) > 1]$feature_id)
  
  diffProteins_i_onlypadj <- unique(protein_DiffExprProtein_i[pBHadj < 0.05]$feature_id)
  
  #save data
  write.table(diffProteins_i, paste0("output_files/diffProteins_", i, ".txt"),
              sep = "\t", quote = F, row.names = F, col.names = F)
  write.table(diffProteins_i_onlypadj, paste0("output_files/diffProteins_onlypadj_", i, ".txt"),
              sep = "\t", quote = F, row.names = F, col.names = F)
  
  diffProteins_global_i <- unique(protein_DiffExprProtein_i[global_pBHadj < 0.05 & abs(global_medianLog2FC) > 1]$feature_id)
  
  #save data
  write.table(diffProteins_global_i, paste0("output_files/diffProteins_global_", i, ".txt"),
              sep = "\t", quote = F, row.names = F, col.names = F)
  
  #' Volcano plots highlighting different Proteins          
  highlight_protein <- diffProteins_i
  plotVolcano(protein_DiffExprProtein_i, highlight = highlight_protein, 
              PDF = F, pBHadj_cutoff = 0.05)
  
  #append to the list
  diffProteins_list[[i]] <- diffProteins_i
  diffProteins_onlypadj_list[[i]] <- diffProteins_i_onlypadj
  diffProteins_global_list[[i]] <- diffProteins_global_i
}

```


```{r compare local vs global}

#### local vs global
for(i in Sample_id_treat) {
            
          sample_vector <- c(i, ref_condition) 
          
          #filter for just 0- and condition x
          proteinFeaturesFiltered_featureValsFilled_i <- proteinFeaturesFiltered_featureValsFilled %>%
            filter(Sample %in% sample_vector) %>%
            mutate(Sample = factor(Sample, levels = sample_vector)) %>%
            arrange(Sample)

  for(cond in diffProteins_list){
  
   testFilled_norm <- proteinFeaturesFiltered_featureValsFilled_i[feature_id %in% cond]

  local_vs_global_test <- testLocalVsGlobal(featureVals = testFilled_norm,
                                            compare_between = "Condition",
                                            plot = TRUE)
  
  # local_vs_global_test_protein <- aggregateLocalVsGlobal(local_vs_global_test)
  # 
  # local_vs_global_test_sig <- unique(local_vs_global_test_protein[pBHadj<0.05][abs(medianDiff) > 0.3]$feature_id)
  # 
  # 
  # #plot
  # for (id in local_vs_global_test_sig) {
  #   plotFeatures(feature_table = proteinFeaturesFiltered,
  #                traces = pepTracesList_filtered_minN,
  #                feature_id = id,
  #                design_matrix=sub_design_matrix,
  #                calibration=calibration,
  #                peak_area = T,
  #                legend = F,
  #                onlyBest = F,
  #                monomer_MW=T,
  #                aggregateReplicates=F)
  # }
  }
}
```


# COMPLEX level analysis

```{r}
# Complex Portal reference table for this organism. The filename is set once in
# the SETUP section via `complex_portal_file` (this branch: S. cerevisiae,
# taxid 559292). Place that file in data/raw/. (Original sep was 't', a stray
# character; corrected to "\t".)
complex_portal <- read.csv(here("data", "raw", complex_portal_file), sep = "\t", header = TRUE) %>%
  janitor::clean_names() 
```

## - set hypothesis

```{r}
# set hypothesis

complexHypotheses <- complex_portal %>%
  dplyr::select(x_complex_ac, recommended_name, aliases_for_complex, description, complex_properties, complex_assembly, expanded_participant_list ) %>% #only keep needed columns 
separate_rows(expanded_participant_list, sep = "|") %>% #separate each partecipant in a different row
separate(expanded_participant_list, c("protein_id", "stoichiometry"), sep = "(", convert = TRUE, remove = FALSE) %>% # separate the stoichiometry in a different column
mutate(stoichiometry = gsub(")", "", stoichiometry)) %>% #remove second braket from the stoichiometry number
 dplyr::rename("complex_id"="x_complex_ac", "complex_name"="recommended_name") # fix col names

complexHypotheses_minimal <- complexHypotheses %>%
   dplyr::select(complex_id, complex_name, protein_id ) %>%
  mutate(complex_id = str_replace(complex_id, "CPX-", "")) %>%
  as.data.table()

head(complexHypotheses_minimal)


binary_hyp <- complexHypotheses_minimal %>%
  dplyr::select(complex_id, protein_id) %>%
  mutate(protein_id = gsub("-PRO_.*", "", protein_id))#remove -PRO_

#### plus decoy

pathLength <- calculatePathlength(binary_hyp)

complexTargetsPlusDecoys <- generateComplexDecoys(target_hypotheses=complexHypotheses_minimal,
                                                dist_info=pathLength,
                                                min_distance = 3,
                                                append=TRUE)


```

## - get Complex features

```{r}
#' ## Summarize protein traces across conditions
protein_traces_sum <- integrateTraceIntensities(protein_traces_list,
                                                design_matrix = NULL,
                                                integrate_within = NULL,
                                                aggr_fun = "sum")
```

```{r}

plotSummarizedMScoverage(hypotheses = complexHypotheses_minimal,
                         protTraces = protein_traces_sum,
                         PDF = F)


#' ## Complex feature finding
complexFeatures <-  findComplexFeatures(traces = protein_traces_sum,
                                        complex_hypothesis = complexTargetsPlusDecoys,
                                        corr_cutoff = 0.9,
                                        window_size = 7,
                                        parallelized = F,
                                        n_cores = 1,
                                        collapse_method = "apex_network",
                                        perturb_cutoff = "5%",
                                        rt_height = 1,
                                        smoothing_length = 7)
```

## - filter Complex features

```{r, eval=TRUE, message=FALSE, warning=FALSE, cache=FALSE}

complexFeaturesBest <- getBestFeatures(complexFeatures)
```

```{r}

complexFeaturesBestFilterd <- filterFeatures(complexFeaturesBest,
                                  complex_ids = NULL,
                                  protein_ids = NULL,
                                  min_feature_completeness = NULL,
                                  min_hypothesis_completeness = NULL,
                                  min_subunits = NULL,
                                  min_peak_corr = NULL,
                                  min_monomer_distance_factor = 2)





complexFeaturesBestFilterdScored <- scoreFeatures(complexFeaturesBestFilterd, FDR=0.1, PDF=F) #,  ## at FDR 0.05 anything passes name="qvalueStats_complexFeatures")

scoredDataAll <- appendSecondaryComplexFeatures(scoredPrimaryFeatures = complexFeaturesBestFilterdScored, allFeatures = complexFeatures, peakCorr_cutoff = 0.7)


plotSummarizedComplexes(scoredDataAll, complexHypotheses_minimal, protein_traces_sum, PDF=F) # ,name="complex_completeness_pie")

summarizeFeatures(scoredDataAll,
                plot=TRUE,
                PDF=F ) #,name="feature_summary_scoredDataAll")

summarizeFeatures(complexFeaturesBestFilterdScored,
                  plot=TRUE,
                  PDF= F) #,  name="feature_summary_complexFeaturesBestFiltered")

```

## - extract Complex features from each pep traces

extract complex features from pep traces pepTracesList_filtered_minN

by condition

```{r}


complexFeaturesUnique <- getUniqueFeatureGroups(
                            feature_table = scoredDataAll,
                            rt_height = 0,
                            distance_cutoff = 1.25)

complexFeaturesCollapsed <- callapseByUniqueFeatureGroups(
                              feature_table = complexFeaturesUnique,
                              rm_decoys = TRUE)

################' ## Extract feature values
if(!file.exists("complex_featureVals.RData")){

complex_featureVals <- extractFeatureVals(traces = pepTracesList_filtered_minN,
                                          features = complexFeaturesCollapsed,
                                          design_matrix = design_matrix,
                                          extract = "subunits",
                                          imputeZero = T,
                                          verbose = F)

save(complex_featureVals, file = "complex_featureVals.RData")}

if(file.exists("complex_featureVals.RData")){
  load("complex_featureVals.RData")}

######################################################################################

complex_featureValsFilled <- fillFeatureVals(featureVals = complex_featureVals,
                                             tracesList = pepTracesList_filtered_minN,
                                             design_matrix = design_matrix)
```

## - differential

```{r}
# Reordering the vector


# create list to append to
complex_DiffExprPep_list <- list()

for(i in Sample_id_treat) {
            
          sample_vector <- c(i, ref_condition) 
          
          #filter for just 0- and condition x
          complex_featureValsFilled_i <- complex_featureValsFilled %>%
            filter(Sample %in% sample_vector) %>%
            mutate(Sample = factor(Sample, levels = sample_vector)) %>%
            arrange(Sample)


          #' ## Perform peptide-level differential expression testing for all features
          complex_DiffExprPep_i <- testDifferentialExpression_1repfix_chatgpt(featureVals = complex_featureValsFilled_i,
                                                            compare_between = "Condition",
                                                            level = "peptide",
                                                            measuredOnly = FALSE)
          
          # Append to the list
                    complex_DiffExprPep_list[[i]] <- complex_DiffExprPep_i
}
```

```{r}
for (i in complex_DiffExprPep_list) {
  
#plots some numbers
hist(i$pVal, breaks = 100)
    hist(i$global_pVal, breaks = 100)
    hist(log(i$qint1), breaks = 100)
    hist(log(i$qint2), breaks = 100)
    hist(log(i$global_int2_imp), breaks = 100)
    hist(log(i$global_int1_imp), breaks = 100)
    hist(i$log2FC, breaks = 50)
    hist(i$global_log2FC, breaks = 50)
}
```

```{r}
complex_DiffExprProtein_list <- list()

for (i in seq_along(complex_DiffExprPep_list)) {
  
  # Get the current dataframe
  df <- complex_DiffExprPep_list[[i]] # need to do this, because otherwise the indexing is wrong when I want to append the results to complex_DiffExprProtein_list
  
#' ## Aggregate differential expression results to the protein level
  complex_DiffExprProtein <- aggregatePeptideTests(df) # need to do this, because otherwise the indexing is wrong when I want to append the results to complex_DiffExprProtein_list


# and plot numbers
hist(complex_DiffExprProtein$pVal, breaks = 100)
    hist(complex_DiffExprProtein$global_pVal, breaks = 100)
    hist(complex_DiffExprProtein$medianLog2FC, breaks = 50)
    hist(complex_DiffExprProtein$global_medianLog2FC, breaks = 50)
    
#append to vector
    # Append to the list with name
  name <- names(complex_DiffExprPep_list)[i]
  complex_DiffExprProtein_list[[name]] <- complex_DiffExprProtein
    }
```

```{r}
complex_DiffExprComplex_list <- list()

for (i in seq_along(complex_DiffExprProtein_list)) {
  
  # Get the current dataframe
  df <- complex_DiffExprProtein_list[[i]] # need to do this, because otherwise the indexing is wrong when I want to append the results to complex_DiffExprComplex_list
  
 #' ## Aggregate differential expression results to the complex level
 complex_DiffExprComplex <- aggregateProteinTests(df)

# plot some numbers
 hist(complex_DiffExprComplex$pVal, breaks = 100)
    hist(complex_DiffExprComplex$global_pVal, breaks = 100)
    hist(complex_DiffExprComplex$medianLog2FC, breaks = 50)
    hist(complex_DiffExprComplex$global_medianLog2FC, breaks = 50)

#append to vector keeping names
  
  name <- names(complex_DiffExprProtein_list)[i]
  complex_DiffExprComplex_list[[name]] <- complex_DiffExprComplex
}
    
```

## - general volcano plots

```{r}
for(i in complex_DiffExprPep_list) {
  plotVolcano(i, PDF = F, pBHadj_cutoff = 0.05)}
```

```{r}
for(i in complex_DiffExprProtein_list) {
  plotVolcano(i, PDF = F, pBHadj_cutoff = 0.05)}
```

```{r}
for(i in complex_DiffExprComplex_list) {
  plotVolcano(i, PDF = F, pBHadj_cutoff = 0.05)}
```

## - Get differential complexes

```{r}

diffComplexes_list <- list()

for (j in seq_along(complex_DiffExprComplex_list)) {
  
  # Extract the current dataframe
  i <- complex_DiffExprComplex_list[[j]]
  
  # Identify unique complexes and differential complexes
  allComplexes <- unique(i$complex_id)
  diffComplexes <- unique(i[i$pBHadj < 0.05 & abs(i$medianLog2FC) > 1, "complex_id"])
  
  #   Append diffComplexes to the list with the name from the original list
  name <- names(complex_DiffExprComplex_list)[j]
  diffComplexes_list[[name]] <- diffComplexes
  
  # Volcano plot highlighting differential complexes
  highlight_complex <- diffComplexes
  plotVolcano(i, highlight = highlight_complex, PDF = FALSE, pBHadj_cutoff = 0.05)
}
```

## - plot diffComplexes

```{r}
  #' # Plot some differential protein features
  #' 
for (i in diffComplexes_list) {
  
    for (id in i) {
        plotFeatures(feature_table = scoredDataAll,
                     traces = protein_traces_list,
                     feature_id = id,
                     annotation_label = "input_id",
                     design_matrix=design_matrix,
                     calibration=calibration,
                     peak_area = T,
                     legend = F,
                     onlyBest = F,
                     monomer_MW=T)
    }
  }

```

## - plot all_complexes

```{r}

all_complexes <- unique(complex_featureValsFilled$complex_id)

for (id in all_complexes) {
  plotFeatures(feature_table = scoredDataAll,
               traces = protein_traces_list,
               feature_id = id,
               annotation_label = "input_id",
               design_matrix=design_matrix,
               calibration=calibration,
               peak_area = F,
               legend = T,
               onlyBest = F,
               monomer_MW=T,
               aggregateReplicates=F) #turn T if n_replicates >1
}
dev.off()
```

## save RData for plotting

```{r}

save(pepTracesList_filtered, pepTracesList_filtered_minN,protein_traces_list,protein_DiffExprPep_list,protein_DiffExprProtein_list,  file = paste0(RData_for_further_plotting_and_analysis_location,"/",experiment_name, "_for_plotting.RData"))

```

# 7. Cross-correlation differential analysis (alternative / complement)

The statistical test above compares **intensity / assembly state** between
conditions. This section is a complementary, shape-based view: for each protein it
compares the **elution profile shape and position** between the two conditions
using the **cross-correlation function (CCF)**.

For every protein we cross-correlate its mean profile in the reference condition
(`strain31`) against the treatment (`strain83`) over a range of fraction lags and
record:

* `max_ccf` — the highest cross-correlation (how similar the two profiles are in
  shape, scale-independent);
* `best_lag` — the lag (in fractions) at which that maximum occurs, i.e. the
  **peak shift** between conditions (0 = no shift);
* `zero_lag_cor` — the correlation with no shift (similarity "as is").

A non-zero `best_lag` (a peak that moves to a different fraction / apparent MW) or
a low similarity flags proteins whose elution behaviour changes between strains —
e.g. complex (dis)assembly — which an abundance-only test can miss.

## 7a. Parameters

```{r ccf-params}
ccf_lag_max         <- 5                       # max shift (fractions) to scan
ccf_shift_threshold <- 1                       # |best_lag| >= this  -> flagged "shifted"
ccf_min_intensity   <- 0                       # require this summed intensity in BOTH conditions
ccf_ref_condition   <- ref_condition           # strain31 (reference)
ccf_treat_condition <- treatment_conditions[1] # strain83 (treatment)
```

## 7b. Per-condition mean elution profiles

```{r ccf-condition-traces}
# Average replicate protein traces within each condition. CCF is scale-invariant,
# so aggr_fun = "sum" (used elsewhere in this pipeline) and "mean" are equivalent
# for the correlation/shift; "sum" is used here because it is known to work with
# this CCprofiler build.
cond_traces <- integrateTraceIntensities(
  protein_traces_list,
  design_matrix    = design_matrix,
  integrate_within = "Condition",
  aggr_fun         = "sum"
)

# Robust extractor: traces$traces -> numeric matrix (proteins x fractions).
.get_trace_matrix <- function(traces_obj) {
  dt <- traces_obj$traces
  frac_cols <- grep("^[0-9]+$", colnames(dt), value = TRUE)
  frac_cols <- frac_cols[order(as.numeric(frac_cols))]
  if (data.table::is.data.table(dt)) {
    m <- as.matrix(dt[, ..frac_cols])
  } else {
    m <- as.matrix(dt[, frac_cols, drop = FALSE])
  }
  rownames(m) <- dt$id
  m
}

ref_mat    <- .get_trace_matrix(cond_traces[[ccf_ref_condition]])
treat_mat  <- .get_trace_matrix(cond_traces[[ccf_treat_condition]])
common_ids <- intersect(rownames(ref_mat), rownames(treat_mat))
message("Cross-correlation over ", length(common_ids), " proteins present in both conditions.")
```

## 7c. Compute the cross-correlation metrics per protein

```{r ccf-compute}
compute_ccf_metrics <- function(x, y, lag_max = 5) {
  x[is.na(x)] <- 0
  y[is.na(y)] <- 0
  if (stats::sd(x) == 0 || stats::sd(y) == 0) {
    return(c(max_ccf = NA_real_, best_lag = NA_real_, zero_lag_cor = NA_real_))
  }
  cc       <- stats::ccf(x, y, lag.max = lag_max, plot = FALSE)
  acf_vals <- as.numeric(cc$acf)
  lags     <- as.numeric(cc$lag)
  i_max    <- which.max(acf_vals)
  zero_i   <- which(lags == 0)
  c(max_ccf      = acf_vals[i_max],
    best_lag     = lags[i_max],
    zero_lag_cor = if (length(zero_i)) acf_vals[zero_i] else NA_real_)
}

ccf_results <- data.table::rbindlist(lapply(common_ids, function(id) {
  m <- compute_ccf_metrics(ref_mat[id, ], treat_mat[id, ], lag_max = ccf_lag_max)
  data.table::data.table(
    protein_id   = id,
    max_ccf      = m[["max_ccf"]],
    best_lag     = m[["best_lag"]],
    zero_lag_cor = m[["zero_lag_cor"]],
    int_ref      = sum(ref_mat[id, ],   na.rm = TRUE),
    int_treat    = sum(treat_mat[id, ], na.rm = TRUE)
  )
}))

ccf_results[, shifted := (!is.na(best_lag)) &
              (abs(best_lag) >= ccf_shift_threshold) &
              (int_ref   >= ccf_min_intensity) &
              (int_treat >= ccf_min_intensity)]
ccf_results[, log2_int_ratio := log2((int_treat + 1) / (int_ref + 1))]
data.table::setorder(ccf_results, -shifted, -best_lag)

message("Proteins flagged as shifted (|lag| >= ", ccf_shift_threshold, "): ",
        sum(ccf_results$shifted, na.rm = TRUE))
DT::datatable(ccf_results[order(-abs(best_lag))][1:min(.N, 100)],
              caption = "Per-protein cross-correlation metrics (top by |shift|)")
```

## 7d. Diagnostic plots

```{r ccf-plots, fig.width=7, fig.height=4}
# Distribution of optimal lags (peak shifts)
print(
  ggplot(ccf_results[!is.na(best_lag)], aes(x = best_lag)) +
    geom_bar(fill = "#4C72B0") + theme_classic() +
    labs(title = "Optimal cross-correlation lag (peak shift) per protein",
         x = "Optimal lag (fractions)   [0 = no shift]", y = "proteins")
)

# Profile-shape similarity (max CCF)
print(
  ggplot(ccf_results[!is.na(max_ccf)], aes(x = max_ccf)) +
    geom_histogram(bins = 50, fill = "#55A868") + theme_classic() +
    labs(title = "Maximum cross-correlation (profile-shape similarity)",
         x = "max CCF", y = "proteins")
)

# Shift vs zero-lag similarity, hits highlighted
print(
  ggplot(ccf_results[!is.na(best_lag)],
         aes(x = best_lag, y = zero_lag_cor, color = shifted)) +
    geom_jitter(width = 0.12, height = 0, alpha = 0.6) +
    scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick")) +
    theme_classic() +
    labs(title = "Elution shift vs zero-lag similarity",
         x = "Optimal lag (fractions)", y = "zero-lag correlation", color = "shifted")
)
```

## 7e. Inspect top shifted proteins (overlaid profiles)

```{r ccf-profile-plots, fig.width=7, fig.height=4}
plot_ccf_protein <- function(id) {
  df <- rbind(
    data.frame(fraction = seq_len(ncol(ref_mat)),   intensity = ref_mat[id, ],
               condition = ccf_ref_condition),
    data.frame(fraction = seq_len(ncol(treat_mat)), intensity = treat_mat[id, ],
               condition = ccf_treat_condition)
  )
  mrow <- ccf_results[protein_id == id]
  ggplot(df, aes(x = fraction, y = intensity, color = condition)) +
    geom_line(linewidth = 1) + geom_point(size = 1) + theme_classic() +
    labs(title = paste0(id, "   (lag = ", mrow$best_lag,
                        ", maxCCF = ", round(mrow$max_ccf, 2), ")"),
         x = "Fraction", y = "Mean intensity")
}

top_shifted <- head(ccf_results[shifted == TRUE][order(-abs(best_lag))]$protein_id, 12)
for (id in top_shifted) print(plot_ccf_protein(id))
```

## 7f. Complement to the statistical test

```{r ccf-save-and-compare}
# Save the cross-correlation table
write.table(
  ccf_results,
  file      = file.path(out_dir, "tables", "crosscorrelation_results.txt"),
  sep       = "\t", quote = FALSE, row.names = FALSE
)

# If the statistical protein-level differential results exist, cross-tabulate
# which proteins each approach flags (the "complement" view). Adjust the p-value /
# log2FC column names to match your stats table if they differ.
if (exists("protein_DiffExprProtein_list")) {
  stat_dt <- data.table::rbindlist(protein_DiffExprProtein_list, fill = TRUE)
  pcol <- intersect(c("pBHadj", "qVal", "pVal", "global_pBHadj"), names(stat_dt))[1]
  if (!is.na(pcol) && "protein_id" %in% names(stat_dt)) {
    stat_best <- stat_dt[, .(stat_p = min(get(pcol), na.rm = TRUE)), by = protein_id]
    merged    <- merge(ccf_results, stat_best, by = "protein_id", all = TRUE)
    merged[, stat_sig := !is.na(stat_p) & stat_p < 0.05]
    print(table(shifted = merged$shifted, stat_significant = merged$stat_sig, useNA = "ifany"))
    message("Saved comparison-ready table; 'shifted' (cross-correlation) vs ",
            "'stat_significant' (", pcol, " < 0.05) shown above.")
  } else {
    message("Found protein_DiffExprProtein_list but no recognised p-value/protein_id ",
            "column - join ccf_results to it manually on protein_id.")
  }
} else {
  message("Statistical protein-level results not in memory; run section 6's ",
          "differential test, then re-run this chunk to compare the two approaches.")
}
```
