---
title: "IA006 - QTL - SEC of mutants 31 and 83- Differential Analysis - normalized"
author: "Ilaria"
date: "`r Sys.Date()`"
output:
  # pdf_document:
  #   latex_engine: xelatex
  #   toc: yes
  html_document:
    toc_depth: 3
    number_sections: yes
    theme: united
    toc: yes
    code_folding: hide
    keep_md: true #enable to save all plots in a folder when knitting to html 
fontsize: 11pt
params:
  comparison_id: "QTL_31_BY"   # default; overridden by scripts/run_all_comparisons.R
editor_options: 
  markdown: 
    wrap: 72
---

```{=tex}
letoldShadedShaded
letendoldShadedendShaded
renewenvironment{Shaded}{footnotesizeoldShaded}{endoldShaded}
letoldverbatimverbatim
letendoldverbatimendverbatim
renewenvironment{verbatim}{footnotesizeoldverbatim}{endoldverbatim}
```

<!-- =========================================================================
 CLEANED MAIN PIPELINE (QTL strain31 vs strain83).
 Reorganised from the original on 2026-06-19. What changed vs the original:
   * Inline "fix" functions removed -> sourced from R/ccprofiler_fixes.R.
   * Package chunk uses library() only (installation is handled by setup.R/renv);
     library(CCprofilerDiffAnna) -> library(CCprofiler).
   * Paths use here::here(); inputs read from data/raw/, outputs go to output/.
   * Exploratory / other-project / one-off code moved to analysis/experiments.Rmd.
 The original is preserved verbatim as ORIGINAL_DiffAnalysis_Norm_QTL_31_83_2026.Rmd.
 PROTEIN QUANTIFICATION: the step that builds `protein_traces_list` (missing from
 the original) is now in place in "6. Protein quantification" - it uses the lab
 sibling/replicate peptide-correlation fix (Option A). The report still needs your
 input files in data/raw/ to knit end-to-end.
============================================================================ -->

# Introduction

Yeast 83(15_6_c) and 31(6_3_c) grown and harvested at OD 0.8, 3 biological replicates each.

## CHECKLIST script:

-   [x] give inputfiles path, filename pattern, set conditions names in
    "setup"

-   [x] go through and change accordingly in "analysis SETUP"

-   [x] do you want to rerun protein feature-finding? if yes, delete
    Rdata file from folder.

-   [x] are there outlier fraction? define and select how to handle in
    "analysis SETUP"

-   [ ] check if fraction annotation file chunck is ok

-   [ ] do you want to filter out proteins with less then x precursors?

# 1. SETUP

```{r setup, include=TRUE}

knitr::opts_chunk$set(echo = TRUE)
## language settings
Sys.setenv(LANG="en")

# ==========================================================================
# COMPARISON REGISTRY  (yeast_QTL)
# One entry per pairwise comparison. Set `comparison_id` to the one you want to
# run; the input file, conditions, colours and output folder are all derived
# from it. RM and BY are parental strains; 31 and 83 are mutant strains.
#   tsv   : Spectronaut report (.tsv) for this comparison
#   ref   : reference / baseline condition (fold-change denominator)
#   treat : treatment / test condition (numerator)
# TODO (per comparison): replace the placeholder .tsv filenames; confirm each
#   ref/treat direction and that these names match the sample names in that .tsv.
# ==========================================================================
comparisons <- list(
  # pattern = a SHORT unique substring of that comparison's .tsv filename (the
  #   file is found automatically in data/raw/, so you don't type long names).
  # ref / treat = condition LABELS exactly as they appear INSIDE the data (the
  #   "strain.." token parsed from r_file_name, e.g. strain83 / strainRM) - NOT the
  #   bare codes in the .tsv filename. Verify with unique(design_matrix$Condition).
  QTL_31_BY = list(pattern = "_31_BY_", ref = "strainBY", treat = "strain31"),
  QTL_31_RM = list(pattern = "_31_RM_", ref = "strainRM", treat = "strain31"),
  QTL_83_BY = list(pattern = "_83_BY_", ref = "strainBY", treat = "strain83"),
  QTL_83_RM = list(pattern = "_83_RM_", ref = "strainRM", treat = "strain83"),
  QTL_RM_BY = list(pattern = "_RM_BY_", ref = "strainBY", treat = "strainRM"),
  # Ethanol growth: the "..._EtOH_83_RM_..." file. Chunk 2b tags every sample from an EtOH file with an
  # "_EtOH" suffix on its strain condition, so this compares the two strains WITHIN ethanol. The pattern
  # contains EtOH, so .find_tsv picks the ethanol file (and glucose QTL_83_RM ignores it).
  # NOTE: ref/treat assume the strain tokens are strain83 / strainRM as in glucose - confirm with
  # unique(design_matrix$Condition) and adjust if the EtOH runs are named differently.
  QTL_83_RM_EtOH = list(pattern = "_EtOH_83_RM_", ref = "strainRM_EtOH", treat = "strain83_EtOH"),
  # CARBON-SOURCE EFFECT: strain83 on glucose vs strain83 on ethanol, from ONE mixed .tsv (both searched
  # together in the same Spectronaut run). Glucose runs stay "strain83"; ethanol runs (with _EtOH_ in the
  # run name) become "strain83_EtOH" - so this compares them directly. ref = glucose baseline.
  # TODO: set `pattern` to a UNIQUE substring of that mixed .tsv filename once it is in data/raw/
  #   (keep "EtOH" in the pattern so .find_tsv doesn't skip the file), then add "QTL_83_Glc_vs_EtOH" to
  #   scripts/run_all_comparisons.R. Confirm both labels with unique(design_matrix$Condition).
  QTL_83_Glc_vs_EtOH = list(pattern = "_83_EtOHvsGlc_", ref = "strain83", treat = "strain83_EtOH")
)

## ---- choose the comparison to run ----
comparison_id <- if (exists("params") && !is.null(params$comparison_id)) params$comparison_id else "QTL_31_BY"

## ---- everything below is derived from the choice (no need to edit) ----
stopifnot(comparison_id %in% names(comparisons))
.cmp                 <- comparisons[[comparison_id]]
experiment_name      <- comparison_id
# find this comparison's .tsv in data/raw/ by its unique filename substring
.find_tsv <- function(pat) {
  all_tsv <- list.files(here::here("data", "raw"), pattern = "\\.tsv$")
  hits <- all_tsv[grepl(pat, all_tsv, fixed = TRUE)]
  # A plain (glucose) pattern like "_83_RM_" ALSO matches the ethanol file "..._EtOH_83_RM_...".
  # Unless the pattern itself asks for EtOH, drop EtOH files so glucose and ethanol stay separate.
  if (!grepl("EtOH", pat, fixed = TRUE)) hits <- hits[!grepl("_EtOH_", hits, fixed = TRUE)]
  if (length(hits) == 0) stop("No .tsv in data/raw/ contains '", pat,
                              "'. Files present: ", paste(all_tsv, collapse = ", "))
  if (length(hits) > 1)  stop("Multiple .tsv match '", pat, "': ",
                              paste(hits, collapse = ", "))
  hits
}
data_location        <- .find_tsv(.cmp$pattern)
condition1           <- .cmp$ref      # reference / baseline
condition2           <- .cmp$treat    # treatment / test
calibration_location <- "IA003_01_SEC_calibration_table.xlsx"   # shared by all comparisons
complex_portal_file  <- "Complex_portal_559292.tsv"             # S. cerevisiae (taxid 559292)
trypsin_id           <- "P00761"   # digestion-control protein for the trypsin QC plot
                                   # (yeast DBs: "P00761"; contaminant-prefixed DBs e.g. E. coli: "CON_P00761")

# Prefix of each sample's INTERNAL run name (r_file_name) that is stripped when
# parsing fraction/condition info in chunk 2b (this is NOT the .tsv filename).
# IMPORTANT: chunk 2b also splits r_file_name by "_" into fixed positions
# (date_instrument_operator_experiment_replicate_condition_fraction_position).
# If your new runs name samples differently, that split (and this prefix) must be
# adjusted - check unique(data$r_file_name) after the import.
file_name_pattern <- c("260408_AS1_ailaria_QTL_")

# colours: reference vs treatment
color_scheme_treatment <- setNames(c("#caf0f0", "#f6acca"), c(condition1, condition2))
condition_levels       <- c(condition1, condition2)

```

## 1a. Parameters

```{r}
run_qcs_wProtti <- T

perform_normalization_cyclicloess <- T #(now it works :) Benni fixed it)   # cyclic-loess normalization ENABLED (was F)

load_proteinFeatures.RData <- T   # T = reuse rdata/proteinFeatures.RData, the assembly test and the feature values from a previous run of THIS comparison (the ~1h findProteinFeatures step plus the serial extract/fillFeatureVals tail). Set F to force a fresh computation after changing upstream data or parameters.
load_normalized.RData <- T   # T = skip the ~30min cyclic-loess normalization on re-runs (loads cached pepTracesNormalized.RData). Set F to force fresh after changing upstream data/filtering.
load_protein_DiffExprPep_and_Protein <- T
# The report scores protein features TWICE - once at feature FDR 0.05 (used everywhere downstream) and
# once at 0.15, which only produces the *_0p15featurefdr differential objects that are saved but not used
# in any figure. That second pass repeats the whole SERIAL CCprofiler tail (scoreFeatures ->
# extractFeatureVals -> fillFeatureVals -> differential test), roughly DOUBLING the slowest part of the
# render. Off by default; set T if you specifically want the looser-FDR comparison.
run_feature_fdr015 <- F
# The assembly test (getMassAssemblyChange) can draw ONE PLOT PER PROTEIN. On a few thousand proteins
# those plots, not the beta regression, dominate that chunk's runtime - they are diagnostics, and none of
# the reported results depend on them. Off by default; set T only to inspect individual proteins.
plot_individual_traces <- F
load_uniprot.RData <- T   # cache the UniProt fetch (per comparison + a shared accumulator) so a GOOD fetch is reused instead of re-hitting UniProt every render. fetch_uniprot is non-deterministic: a throttled response comes back WITHOUT the `mass` column and crashes trace annotation. Set F (or delete output/uniprot_annotation_shared.RData) to force a fresh fetch.
################### ordering the samples?

sample_sort_by_number <- T # if need to reorder samples e.g. based on treatment concentration

################### removing lowest MW in MW calibration curve

remove_lowest_MW <- T

################### defining ref and treatment conditions

ref_condition <- condition1

treatment_conditions <- c(condition2) #,condition3,condition4)

################### aggregate replicate when plotting?

aggregate_rep <- T # choose T if more then 1 rep per condition, F if only 1 rep

################### what to do with the problematic fractions?

run_annotateOutlierFractions <- T # don't want to run this, I'm annotating it manually

problematic_fractions <- c() #c("P2_36", "P2_47", "P2_48") # to annotate manually


################### GRID-SEARCH

perform_prot_grid_search <- T
# The grid search is the single most expensive step (it runs findProteinFeatures once per parameter
# combination). It is now CACHED per comparison in output/<comparison_id>/ProtGridSearch.RData and is
# skipped automatically whenever that file exists, so a re-render never repeats it.
force_prot_grid_search <- F     # T = recompute even if the cache exists (after changing data/filtering/grid)
grid_search_from       <- NULL  # e.g. "QTL_83_RM" to copy that comparison's search instead of running one.
                                # The optimal parameters describe the CHROMATOGRAPHY, not the contrast,
                                # so borrowing across comparisons of the same SEC setup is defensible -
                                # but say so in the methods.

# With ibmt pval < 5e-3 in the control-control comparison experiment - TODO refine the way we look at this, but good enough for now!
#proteins_to_exclude <- c("P04816", "P07003", "P07118", "P07822", "P08839", "P08997", "P09126", "P0A6C1", "P0A6C5", "P0A6F3", "P0A6H5", "P0A6T5", "P0A6U3", "P0A6W0", "P0A6W9", "P0A725", "P0A759", "P0A780", "P0A7B3", "P0A7G2", "P0A817", "P0A862", "P0A8I1", "P0A8P3", "P0A8V0", "P0A951", "P0A9H1", "P0A9P9", "P0AA16", "P0AAN5", "P0AAR3", "P0AAZ7", "P0AB14", "P0AB43", "P0AB46", "P0AB61", "P0AB65", "P0AB80", "P0ABE2", "P0ABH0", "P0AD53", "P0ADA1", "P0ADE6", "P0ADU5", "P0ADW8", "P0AE08", "P0AE48", "P0AE60", "P0AE63", "P0AEM4", "P0AES6", "P0AES9", "P0AFF6", "P0AFG0", "P0AFH0", "P0AFI7", "P0AFL3", "P0AFM4", "P0AFW0", "P0AFW8", "P0AGG4", "P0C0L9", "P10346", "P10442", "P16700", "P21151", "P21177", "P21889", "P23538", "P23843", "P23878", "P25524", "P26602", "P27278", "P27303", "P28248", "P30176", "P32141", "P32685", "P33219", "P33355", "P36659", "P36979", "P37613", "P37665", "P37691", "P39274", "P39408", "P40191", "P42616", "P45470", "P52119", "P64490", "P64506", "P64519", "P64559", "P68206", "P69222", "P69795", "P69822", "P76141", "P76243", "P76422", "P76423", "P76569", "P77695", "Q2M7X4", "Q46798")

#Additional proteins to keep that upon manual inspection show an opposite effect than what occurs just by time
proteins_to_keep <- c()



```

## 1b. create folders

```{r}

# All outputs for this comparison go under output/<comparison_id>/
out_dir <- here::here("output", comparison_id)

output_files_location <- file.path(out_dir, "output_files")
RData_for_further_plotting_and_analysis_location <- file.path(out_dir, "RData_for_further_plotting_and_analysis")

# Progress breadcrumbs. knitr buffers chunk messages, so under rmarkdown::render you cannot see which
# step a long chunk is on. .brk() appends a timestamped line to out_dir/render_progress.log (file writes
# are not buffered): open that file while the render runs and the LAST line is the current step.
# Defined here in SETUP so every chunk can use it.
.brk <- function(msg) { try({ dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  cat(format(Sys.time(), "%H:%M:%S"), "|", msg, "\n",
      file = file.path(out_dir, "render_progress.log"), append = TRUE) }, silent = TRUE) }
DistinctAssembly_hits_location <- file.path(out_dir, "DistinctAssembly_hits")
DistinctAssembly_hits_nolog2FC_location <- file.path(out_dir, "DistinctAssembly_hits_nolog2FC")

# Create the comparison's output folders (incl. figures/tables/rdata) if missing
for (d in c(out_dir,
            file.path(out_dir, c("figures", "tables", "rdata")),
            output_files_location,
            RData_for_further_plotting_and_analysis_location,
            DistinctAssembly_hits_location,
            DistinctAssembly_hits_nolog2FC_location)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
cat("Outputs for this run go to: ", out_dir)

```


## 1c. Loading packages and libraries


Bugs on the CCprofiler functions are fixed in CCprofilerDiffAnna, in the PPlabFunctions.R file. 
There the IBMT test, gaussian fitting sliding_window functions and are also saved.

meaning that essential for this analysis are the packages:
** CCprofilerDiffAnna
** PrInCE

```{r, message=FALSE, warning=FALSE, cache=FALSE}
#' # Load CCprofiler package and set working directory
#library(devtools)
library(data.table)
library(stringr)
library(ggplot2)
library(betareg)
library(lmtest)
library(ggrepel)

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("preprocessCore")
# 
# install_github("CCprofiler/CCprofiler", ref = "differential")
###### if the upper part does not work -> use the following: I've downloaded the repository in my downloads folder
#devtools::install_local("/Volumes/biol_bc_picotti_1/Anna/PROJECTS/SEC_metabolites/CCprofiler/CCprofiler-differential", force = TRUE)
#devtools::install_local("W:/PMC_project/CCprofiler/CCprofiler-differential", force = TRUE)
#library(CCprofiler)


#devtools::install_github("AnnaPagotto/CCprofilerDiffAnna", dependencies = TRUE)
#devtools::install_local("W:/PMC_project/CCprofiler/CCprofilerDiffAnna-differential", force = TRUE)
library(CCprofiler)   # was CCprofilerDiffAnna (lab fork); fixes now loaded from R/ccprofiler_fixes.R

library(scales)

########## my usuals

library(protti)
library(tidyverse)
library(DT)

library(readxl) #to read Excel files like .xlsx with the function read_excel()
library(writexl) # to write Excel files

# for overlap plot
#install.packages("UpSetR")
library(UpSetR)

#For Gaussian fitting and probably new complex identification

# if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install("PrInCE")
library(PrInCE)

library(MSnbase)
library(methods)
library(progress)
library(rlist)
library(data.table)
library(ggpmisc)
library(limma)

### additional packages required

# NOTE: package installation is handled once by setup.R / renv (see README.md).
# (The original had install.packages(parallel/foreach/doParallel/igraph) here,
#  which also errored because the package names were unquoted.)

library(parallel)
library(foreach)
library(doParallel)
library(igraph)  # For network-based proteoform detection

```

'here' settings

```{r}
#devtools::install_local("C:/Users/apagotto/Downloads/rprojroot-main/rprojroot-main", dependencies = TRUE)
# install.packages("here")
library(here)
#install.packages("fs")
library(fs)


# List all files and directories recursively

#dir_tree(path = here() , recurse = TRUE)

here()
```


## 1d. Load the CCprofiler fixes

The lab fixes that used to be pasted inline here now live in
`R/ccprofiler_fixes.R` (see that file's header for full provenance). They are
sourced **on top of** the official CCprofiler package, so the corrected versions
take precedence when called from this document.

```{r load-fixes, message=FALSE, warning=FALSE}
source(here::here("R", "ccprofiler_fixes.R"), local = TRUE)  # local=TRUE: define the fixes in the report env so they see design_matrix etc. under rmarkdown::render (new.env)
```

# 2. DATA preparation

## 2a. Import data

```{r, eval=TRUE}


if(file.exists(file.path(out_dir, "rdata", "data_input.RData"))){
  load(file.path(out_dir, "rdata", "data_input.RData"))
} else {
  data <- read_protti(here("data", "raw", data_location))
  save(data, file = file.path(out_dir, "rdata", "data_input.RData"))
}

head(data, n = 10)
```

## 2b. Fraction annotation table To import the data into the *CCprofiler*
container format, an annotation table mapping the biochemical fraction
numbers to the respective file name is required. It is important that
the fractions start at 1 and consecutively increase by 1 unit until
reaching the total number of fractions. An example annotation file is
available as `exampleFractionAnnotation`.

filename fraction_number heuselm_J130729_001 1

```{r, eval=TRUE}

#unique(data$r_file_name)

# CARBON SOURCE (EtOH vs glucose): ethanol runs carry an extra "_EtOH_" token that glucose runs lack.
# Detection is PER-SAMPLE from the run name, so a MIXED file - strain83 glucose + strain83 ethanol
# searched together for the carbon-source comparison - keeps each sample correct (glucose stays strain83,
# ethanol becomes strain83_EtOH). Fallback: if the .tsv NAME says EtOH but NONE of its runs carry the
# token, the whole file is ethanol (covers an ethanol-only file whose runs aren't individually marked).
# We strip the token before the fixed-position split so `condition` still lands on the strain, then fold
# the carbon source back in as a suffix. Glucose runs are untouched.
# VERIFY after import: unique(fraction_annotation$condition).
.is_etoh_file    <- grepl("_EtOH_", data_location, fixed = TRUE)
.any_etoh_in_run <- any(grepl("_EtOH_", data$r_file_name, fixed = TRUE))
.file_all_etoh   <- .is_etoh_file && !.any_etoh_in_run
fraction_annotation <- data %>%
  mutate(carbon_source = ifelse(grepl("_EtOH_", r_file_name, fixed = TRUE) | .file_all_etoh, "EtOH", "glucose"),
         r_split_tmp   = str_replace(r_file_name, "_EtOH_", "_")) %>%
  separate(r_split_tmp, c("date", "instrument", "operator", "experiment", "replicate_number", "condition", "fraction_number", "plate_position"), sep="_", remove = TRUE, extra="drop") %>% # get the fraction number
  mutate(condition = ifelse(carbon_source == "EtOH", paste0(condition, "_EtOH"), condition)) %>%
  dplyr::select(r_file_name, condition, fraction_number, replicate_number) %>%
  distinct() %>%
  mutate(r_file_name = str_replace_all(r_file_name, pattern = file_name_pattern, replacement = "")) %>%
  mutate(fraction_number= str_replace_all(fraction_number, pattern = "f", replacement = "")) %>%
  mutate(replicate_number= str_replace_all(replicate_number, pattern = "biolrep", replacement = ""))
message("Conditions parsed (check EtOH suffix landed on the strain, not a shifted token): ",
        paste(sort(unique(fraction_annotation$condition)), collapse = ", "))
  

head(fraction_annotation, n = 20)
```

## 2c. MW calibration In some co-fractionation techniques, such as
size-exclusion chromatography, a log-linear relationship between the
fraction dimension and molecular weight (MW) can be assumed. In such
cases a MW calibration is possible by leveraging the elution information
of standard proteins with known MWs run separately on the fractionation
device. The inferred apparent molecular weight can then be used to judge
a protein’s likely assembly state given its monomeric molecular weight
and observed elution fraction (~ apparent molecular weight). Monomer
weights can be added to the data by the annotation functions (see
below).

```{r, eval=TRUE, fig.show='hold'}

calibrationTable <- read_excel(here("data", "raw", calibration_location)) %>%
  as.data.table() 

if(remove_lowest_MW == T) {calibrationTable <- calibrationTable %>%
  filter(std_weights_kDa != 0.244)}


calibration = calibrateMW(calibrationTable)
```

## 2d. FILTER data

Filtering for intensity > 64 (log2(I)>6)

```{r}
data_filter <- data %>%
  filter(pep_is_proteotypic == T) %>% # remove non proteotypic peptides
   mutate(fg_quantity_log2 = log2(fg_quantity)) %>% 
  filter(fg_quantity >= 50) %>% # only include intensities above log2 value of 6 (= fg_quantity of 64). Lower intensities are likely false assignments. This is for the Astral: we are being less stringent than with the other instruments, where we use a cutoff of fg_quantity_log2>10 
  filter(pg_protein_accessions != "iRT-Kit_WR_fusion") %>%
  dplyr::select(r_file_name, pg_protein_accessions, pep_grouping_key, fg_quantity, pep_stripped_sequence, pep_nr_of_missed_cleavages)%>%
  mutate(r_file_name = str_replace_all(r_file_name, pattern = file_name_pattern, replacement = "")) #%>%
  #rename(pg_precursor_id = pep_grouping_key)
  
```

# 3. QCs with Protti

```{r}
if(run_qcs_wProtti == T){
  qc <- data_filter %>%
  left_join(fraction_annotation)
}
```

## 3a. Intensity distribution before filtering 

```{r}
data %>%
  mutate(fg_quantity_log2 = log2(fg_quantity)) %>%
  qc_intensity_distribution(
    grouping = pep_grouping_key,
    intensity_log2 = fg_quantity_log2,
    plot_style = "histogram"
  ) +
  geom_vline(xintercept = 5.64, color = "red", linewidth = 0.3) + #log2(50) = 5.643856
  labs(subtitle = "Red line = cutoff for removal of low intensities") 
```



## 3b. IDs

```{r ids_per_frac, fig.height=7, fig.width=15, dev='pdf'}
if(run_qcs_wProtti == T){

###### Pep (precursor) Ids
  
  for (i in unique(qc$condition)) {
    
      plot <- qc%>%
        filter(condition==i) %>%
        qc_ids(, 
             sample = r_file_name,
             grouping = pep_grouping_key,
             intensity = fg_quantity, # precursor lvl
             condition = condition,
             title = paste0("Number of peptide IDs per sample_", i),
             interactive = FALSE) +
        theme(legend.position = "bottom") +
        ggplot2::scale_fill_manual(values = color_scheme_treatment) 
      
      print(plot)
    }

pep_ids <- qc_ids(qc, 
       sample = r_file_name,
       grouping = pep_grouping_key,
       intensity = fg_quantity, # precursor lvl
       condition = condition,
       plot = FALSE)

head(pep_ids)


###### Prot ids

    for (i in unique(qc$condition)) {
        
        plot <- qc%>%
          filter(condition==i) %>%
          qc_ids(
             sample = r_file_name,
             grouping = pg_protein_accessions,
             intensity = fg_quantity,
             condition = condition,
             title = paste0("Number of protein IDs per sample_", i),
             interactive = FALSE) +
        theme(legend.position = "bottom") +
        ggplot2::scale_fill_manual(values = color_scheme_treatment)
        
        print(plot)
    }

prot_ids <- qc_ids(qc, 
       sample = r_file_name,
       grouping = pg_protein_accessions,
       intensity = fg_quantity,
       condition = condition,
       plot = FALSE)

head(prot_ids)
}
```

```{r total_prot_IDs}
if(run_qcs_wProtti == T){

###### Protein Ids

      plot <- qc%>%
        qc_ids( 
             sample = condition,
             grouping = pg_protein_accessions,
             intensity = fg_quantity, # precursor lvl
             condition = condition,
             title = "Number of protein IDs per condition",
             interactive = FALSE) +
        theme(legend.position = "bottom") +
        ggplot2::scale_fill_manual(values = color_scheme_treatment) 
      
      print(plot)
      
      prot_ids_all <- qc_ids(qc, 
       sample = condition,
       grouping = pg_protein_accessions,
       intensity = fg_quantity,
       condition = condition,
       plot = FALSE)
      
      prot_ids_all
    }
```

## 3c. missed cleavages

```{r qc_missed_cleavages, fig.height=7, fig.width=15, dev='pdf'}

if(run_qcs_wProtti == T){
  
  for (i in unique(qc$condition)) {
    
    plot <- qc%>%
          filter(condition==i) %>%
      qc_missed_cleavages(
        sample = r_file_name,
        grouping = pep_stripped_sequence,
        missed_cleavages = pep_nr_of_missed_cleavages,
        intensity = fg_quantity,
        method = "intensity",
        plot = T
      ) +
      labs(subtitle = i)
    
    print(plot)}
}
```

# 4. CC PROFILER differential analysis workflow

## 4a. create DESIGN MATRIX (w/ fraction_annotation)

```{r design_matrix}

#fix colnames and as.data.table

fraction_annotation_table <- fraction_annotation %>%
  dplyr::rename("filename" = "r_file_name") %>%
  mutate(fraction_number = as.numeric(fraction_number)) %>%
as.data.table() %>%
  arrange(fraction_number) %>% ################################# it is important to reorder the fraction_annotation file because otehrwise it gives you an error when running the importPCPdata() function
  dplyr::rename("condition_id" = "condition") %>%
  dplyr::rename("replicate_id" = "replicate_number") %>% ##just to have the same names as Isabell...
  mutate(sample= paste0(condition_id, "_", replicate_id)) 
 
  

fraction_annotation_sub <- fraction_annotation_table %>%
  dplyr::select(c(filename,fraction_number,sample)) %>%
  distinct()
  

 
## create design matrx
  
design_matrix <- unique(subset(fraction_annotation_table, select=c("condition_id","replicate_id","sample")))

setnames(design_matrix, c("condition_id","replicate_id","sample"), c("Condition","Replicate","Sample_name"))

design_matrix$Condition <- factor(design_matrix$Condition, levels= condition_levels)

# Safety net: factor() above turns Condition into NA if the registry ref/treat
# don't match the labels parsed from the data. Fail fast with a clear message
# instead (this mismatch is what previously cascaded into the q-value / hist
# errors much further downstream).
if (anyNA(design_matrix$Condition)) {
  stop("design_matrix$Condition has NA: the comparison's ref/treat (",
       paste(condition_levels, collapse = ", "),
       ") do not match the condition labels in the data. Labels found: ",
       paste(sort(unique(as.character(fraction_annotation_table$condition_id))), collapse = ", "),
       ". Edit ref/treat in the SETUP comparison registry to match.")
}


```

## 4b. fix data_filter for CCprofiler

```{r data_filter_ccp_mod}

data_filter_ccp_mod <- data_filter %>%
  filter(!grepl("CON_", pg_protein_accessions) | pg_protein_accessions == trypsin_id) %>% #remove contaminants, except for trypsin (trypsin_id in SETUP)
dplyr::rename ("filename" = "r_file_name", "protein_id" = "pg_protein_accessions", "peptide_id"="pep_grouping_key", "intensity"="fg_quantity") %>%
 left_join(fraction_annotation_table) %>%
  group_by(filename, protein_id) %>% ### FILTER OUT ALL THE PROTEIN WITH LESS THAN 3 eg_precursor_id?
add_count(protein_id, name = "n_precursors") %>%# get the n precursors per protein per sample
   dplyr::select(protein_id, peptide_id, filename, intensity)%>%
as.data.table()

```

## 4c. create TRACE ANNOTATION table

```{r Fetch_uniprot for traceAnnotation df}
# UniProt annotation fetch - made reliable and SHARED across comparisons (ported from the Ecoli branch).
# fetch_uniprot is a live network call. Fetching it fresh for every comparison hammers UniProt, and a
# throttled/flaky response comes back either WITHOUT a `mass` column (crashing the rename below) or with
# `mass` = NA (which silently makes the assembly state read as ~0% monomer). Two safeguards, neither of
# which changes WHICH proteins/columns are requested or the mass values:
#   (1) a SHARED accumulator cache (output/uniprot_annotation_shared.RData): any protein fetched for ANY
#       comparison is reused, so across a full run the proteome is fetched ~once, not once per comparison;
#   (2) a retry that re-queries proteins still MISSING *or with NA mass*, backing off between tries.
.uniprot_file   <- file.path(out_dir, "rdata", "uniprot.RData")             # this comparison's subset
.uniprot_shared <- here::here("output", "uniprot_annotation_shared.RData")  # accumulates across comparisons
.uniprot_cols   <- c("protein_name", "length", "sequence", "ft_domain", "mass", "gene_names","organism_name","xref_geneid",
    "xref_string", "go_f", "go_p", "go_c", "cc_interaction", "ft_act_site", "ft_binding",
    "cc_cofactor", "cc_catalytic_activity", "xref_pdb")
.want <- unique(as.character(data_filter_ccp_mod$protein_id))

# ids in a table that already have a real (non-NA) mass
.have_mass <- function(dt) if (data.table::is.data.table(dt) && nrow(dt) && all(c("input_id","mass") %in% names(dt)))
  as.character(dt[!is.na(mass)]$input_id) else character(0)

# always load the shared cache (so other comparisons' proteins are never lost); flag F = force refresh
.uniprot_all <- NULL
if (file.exists(.uniprot_shared)) load(.uniprot_shared)   # -> .uniprot_all
if (!data.table::is.data.table(.uniprot_all)) .uniprot_all <- data.table::data.table()
# if the shared cache predates a newly-requested column (e.g. ft_domain), rebuild it so every protein is
# re-fetched WITH that column instead of silently carrying it as all-NA.
if (nrow(.uniprot_all) && !all(.uniprot_cols %in% names(.uniprot_all))) {
  message("Shared UniProt cache is missing requested column(s) (",
          paste(setdiff(.uniprot_cols, names(.uniprot_all)), collapse = ", "),
          ") - rebuilding it so they are fetched.")
  .uniprot_all <- data.table::data.table()
}
if (!isTRUE(load_uniprot.RData) && nrow(.uniprot_all))
  .uniprot_all <- .uniprot_all[!(as.character(input_id) %in% .want)]   # drop this comparison's rows -> re-fetch fresh

.need <- setdiff(.want, .have_mass(.uniprot_all))
if (length(.need)) {
  message("UniProt: ", length(.want) - length(.need), "/", length(.want),
          " proteins reused from shared cache; fetching ", length(.need), " new one(s).")
  for (.try in 1:4) {
    .todo <- setdiff(.need, .have_mass(.uniprot_all))
    if (!length(.todo)) break
    .before <- length(.have_mass(.uniprot_all))
    if (.try > 1) { message("  retry ", .try, ": ", length(.todo), " protein(s) still without mass - backing off ..."); Sys.sleep(5 * (.try - 1)) }
    .batch <- tryCatch(data.table::as.data.table(fetch_uniprot(.todo, columns = .uniprot_cols, batchsize = 50)),
                       error = function(e) { message("  fetch attempt ", .try, " errored: ", conditionMessage(e)); NULL })
    if (!is.null(.batch) && nrow(.batch) && "input_id" %in% names(.batch)) {
      # fresh rows supersede any prior NA-mass rows for those ids
      if (nrow(.uniprot_all) && "input_id" %in% names(.uniprot_all)) {
        .uniprot_all <- data.table::rbindlist(
          list(.uniprot_all[!(as.character(input_id) %in% as.character(.batch$input_id))], .batch),
          use.names = TRUE, fill = TRUE)
      } else {
        .uniprot_all <- .batch
      }
    }
    # Stop retrying the moment an attempt returns NO new masses: that means UniProt is hard-blocking
    # (rate limit), and hammering it further only extends the block. Wait ~1h and re-run instead.
    if (length(.have_mass(.uniprot_all)) <= .before) {
      message("  no new masses returned - UniProt is likely rate-limiting. Stopping; wait ~1h and re-run.")
      break
    }
  }
  dir.create(dirname(.uniprot_shared), recursive = TRUE, showWarnings = FALSE)
  if (nrow(.uniprot_all)) save(.uniprot_all, file = .uniprot_shared)
}

uniprot <- if (nrow(.uniprot_all) && "input_id" %in% names(.uniprot_all)) .uniprot_all[as.character(input_id) %in% .want] else data.table::data.table()
.cov <- if (length(.want)) mean(.want %in% .have_mass(uniprot)) else 0
message(sprintf("UniProt annotation ready: %.1f%% of %d proteins have a monomer mass.", 100 * .cov, length(.want)))
uniprot <- tibble::as_tibble(uniprot)   # restore fetch_uniprot's original tibble type for the downstream dplyr select()/left_join()
dir.create(dirname(.uniprot_file), recursive = TRUE, showWarnings = FALSE)
save(uniprot, file = .uniprot_file)
```

```{r traceAnnotation_table}
## Trace annotation table

## information from uniprot
# Guard the mass column before renaming. If fetch_uniprot returned no `mass` column, the rename below
# crashes with "Column `mass` doesn't exist"; if it returned an all-NA mass, protein_mw ends up NA and
# CCprofiler reads every protein as ~100% assembled (0% monomer). Fail loudly here instead.
if (!"mass" %in% names(uniprot)) {
  stop("fetch_uniprot returned no 'mass' column for this run - the UniProt fetch was incomplete ",
       "(throttling). Delete output/uniprot_annotation_shared.RData and re-run to re-fetch (the shared ",
       "cache + retry usually clears it). Without monomer mass the assembly state is meaningless.")
}
traceAnnotation <- as.data.table(uniprot) %>%
  dplyr::rename("Mass" ="mass") # there is a bug in the annotateTraces function, it doesn't add the protein_mw column if the mass column is "mass" and not "Mass". Tried to fix the function but not fully managed: it does add the column, but put NAs as values....

# Sanity-check the monomer mass just loaded: mostly-NA mass => the assembly state would be garbage.
.na_mass_frac <- mean(is.na(traceAnnotation$Mass))
if (.na_mass_frac > 0.9) {
  stop(sprintf(paste0("UniProt mass is NA for %.0f%% of proteins - the fetch effectively returned no ",
                      "monomer mass. Delete output/uniprot_annotation_shared.RData and re-run to re-fetch."),
               100 * .na_mass_frac))
} else if (.na_mass_frac > 0.5) {
  warning(sprintf("UniProt mass is NA for %.0f%% of proteins - the assembly state may be unreliable.", 100 * .na_mass_frac))
}

head(traceAnnotation, n = 2)
```

## 4d. Get traces list

```{r pepTraces_raw}

samples <- unique(fraction_annotation_table$sample)


# to order by number in the string
if(sample_sort_by_number == T) {
samples <- str_sort(samples, numeric = TRUE)}


traces_list <- lapply(samples,function(x){
  message(x)
  ann <- subset(fraction_annotation_sub, sample==x) 
  ann <- subset(ann, filename %in% data_filter_ccp_mod$filename) 
  ann <- subset(ann, select = c("filename","fraction_number")) 
  setkey(ann,fraction_number) 
  data_sub_select <- subset(data_filter_ccp_mod,filename %in% ann$filename) 
  data_sub_select <- merge(data_sub_select,ann, by=c("filename")) 
  setkey(data_sub_select,fraction_number) 
  data_sub_select[,fraction_number:=NULL] 
  traces <- importPCPdata(input_data=data_sub_select,fraction_annotation=ann)
  return(traces)
})


names(traces_list) = samples
class(traces_list) <- "tracesList"


#' # Annotate traces with information from uniprot
 
pepTraces_raw <- annotateTraces(traces= traces_list,
                            trace_annotation= traceAnnotation,
                            traces_id_column = "protein_id",
                            trace_annotation_id_column = "input_id",
                            #trace_annotation_mass_column = "Mass",
                            uniprot_mass_format = TRUE,
                            replace_whitespace = TRUE)

#' # Annotate traces with molecular weight calibration
pepTraces_raw <- annotateMolecularWeight(pepTraces_raw,
                                     calibration)

summary(pepTraces_raw)
```


# 5. QC with CCprofiler

## 5a. alignment

```{r align across all samples}

#' # Assess how well SEC traces align across all samples
alignTraces(pepTraces_raw, min_lag = -3, max_lag = 3, plot = T, PDF=F) #to save as PDF, set PDF=T
#' # Assess total intensity as a proxi for extraction efficiency across samples
plotGlobalIntensities(pepTraces_raw, plot = T, PDF=F)
```

## 5b. annotateOutlierFractions & impute

```{r}
if(run_annotateOutlierFractions == T) {
pepTracesOutlFrac <- annotateOutlierFractions(traces=pepTraces_raw,
                                              excludeBoarder = T,
                                              plot = F)

annotateOutlierFractions(traces=pepTraces_raw,
                         excludeBoarder = T,
                         plot = TRUE,
                         PDF = FALSE,
                         name = "OutlierFractions")
}
```

```{r}

if(run_annotateOutlierFractions == F) {
pepTracesOutlFrac <- lapply(pepTraces_raw, function(lst) {
  # Add the isOutlier column to the fraction_annotation data frame
    # Check if 'filename' is in problematic_fractions and create the isOutlier column
  lst$fraction_annotation$isOutlier <- lst$fraction_annotation$filename %in% problematic_fractions
  
  return(lst)
})

# Reassign the class to "tracesList"
class(pepTracesOutlFrac) <- class(pepTraces_raw)

# Check the class of the output
class(pepTracesOutlFrac)
}
  
```

```{r}
### NB NB NB, use mean, not spline, mean is much better, Benni said :) But there is a bug in the code -> even if you set 'mean', it will go for spline anyway -> fix the bug (Benni already did...)

pepTracesOutlFracImp <- imputeOutlierFractions( traces = pepTracesOutlFrac,
                                                bound_left = 2,
                                                bound_right = 2,
                                                consider_borders = TRUE,
                                                method = "spline" # mean is also an option
)
```

## 5c. general imputation

```{r Find missing values}

  
#' Find missing values
#' (defined as having identifications in left and right neigbouring fractions):
pepTracesMV <- findMissingValues(pepTracesOutlFracImp,
                                 bound_left = 1,
                                 bound_right = 1,
                                 consider_borders = TRUE)

#' Impute NA values by fitting a spline:
pepTracesImp <- imputeMissingVals(pepTracesMV, method = "spline")

#' Plot imputation summary:
plotImputationSummary(pepTracesMV, pepTracesImp, PDF = F,
                      plot_traces = T, max_n_traces = 2)

```



## 5d. normalization

```{r Normalize intensity values across samples}

if(perform_normalization_cyclicloess==T) {
# Cache the (~30min) cyclic-loess normalization, same idea as proteinFeatures.RData: if
# load_normalized.RData is TRUE in SETUP and a previous run saved pepTracesNormalized.RData,
# load it and skip normalizeByCyclicLoess. Set the flag to F (or delete the file) to force a
# fresh normalization after changing anything upstream (imputation, filtering, the input data).
.norm_file <- file.path(out_dir, "rdata", "pepTracesNormalized.RData")
if (load_normalized.RData && file.exists(.norm_file)) {
  load(.norm_file)   # restores `pepTracesNormalized` from the previous run
  message("Loaded cached pepTracesNormalized from ", .norm_file, " (skipped normalizeByCyclicLoess).")
} else {
# Normalize intensity values across samples
pepTracesNormalized <- normalizeByCyclicLoess(pepTracesImp, window = 3, step = 1, plot = F)
# NOTE: a second, identical normalizeByCyclicLoess() call used to run here purely to "draw a plot"
# (plot=T). But the `plot`/`PDF` arguments are ignored inside the function, and the call above
# already writes the prior/post-normalization PDFs. That duplicate ran the full (expensive)
# normalization a second time for zero extra output, doubling this chunk's runtime - removed.
save(pepTracesNormalized, file = .norm_file)
message("Saved pepTracesNormalized to ", .norm_file, " (re-runs with load_normalized.RData=T skip normalization).")
}

#' # Assess total intensity after normalization
plotGlobalIntensities(pepTracesNormalized, plot = T, PDF=F ) #, name = "IntensitySummary_postNormalization")
}


if(perform_normalization_cyclicloess==F) { pepTracesNormalized <- pepTracesImp}

```

### -Trypsin plot for QC before  and after normalization
```{r}
# Trypsin (digestion control) ID differs per dataset (set as trypsin_id in SETUP).
# Resolve robustly (try trypsin_id, then its CON_/non-CON_ variant) and skip with a
# message if absent, instead of crashing on an empty traces subset.
.tryp_ids <- unique(unlist(lapply(pepTraces_raw, function(t) t$trace_annotation$protein_id)))
.tryp <- intersect(c(trypsin_id, sub("^CON_", "", trypsin_id), paste0("CON_", trypsin_id)), .tryp_ids)
if (length(.tryp) == 0) {
  message("Trypsin control '", trypsin_id, "' not found - skipping the trypsin QC plot.")
} else {
  .tryp <- .tryp[1]
  plot(subset(pepTraces_raw,       trace_subset_ids = .tryp, trace_subset_type = "protein_id"),
       legend = F, design_matrix = design_matrix, name = "Trypsin non-normalized")
  plot(subset(pepTracesNormalized, trace_subset_ids = .tryp, trace_subset_type = "protein_id"),
       legend = F, design_matrix = design_matrix, name = "Trypsin normalized")
}
```


> **Moved:** the exploratory *trypsin-ratio normalization* that was here has been
> moved to `analysis/experiments.Rmd`. It was an alternative to the normalization
> above and is **not** used by the main pipeline, which continues from
> `pepTracesNormalized`.

## - Pep-lvl filtering

```{r}
#' # Filter by consecutive IDs --> comment out: and peptide correlation--> work with pepTracesConsIds instead of pepTracesSPC
#'
# (CCprofiler + fixes are already loaded in section 1; no need to reload here)

# Combine all traces for filtering:

traces_combined <- combineTracesMutiCond(pepTracesNormalized)
# Filter by consecutive IDs:
pepTracesConsIds <- filterConsecutiveIdStretches(traces_combined,
                                                 min_stretch_length = 3,
                                                 remove_empty = T)

# Filter by maximum correlation for outlier removal:
pepTracesMaxCorr <- filterByMaxCorr(pepTracesConsIds,
                                cutoff = 0.8,
                                plot = T, PDF = F)

# Filter by consecutive SPC:
# pepTracesSPC <- filterBySibPepCorr(pepTracesMaxCorr,
#                                    absolute_spcCutoff = 0.6,
#                                    plot = T, PDF = F)
```

```{r}
#' # Subset pepTracesNormalized to valid peptides selected 
#' by consecutive filtering and SPC cutoff
validPeps <- unique(pepTracesMaxCorr$trace_annotation$id)
pepTracesList_filtered <- lapply(pepTracesNormalized, function(x){
  sub_trace <- subset(x, trace_subset_ids=validPeps)
  sub_trace$trace_annotation <- sub_trace$trace_annotation# %>% #left_join(select(pepTracesMaxCorr$trace_annotation, id #SibPepCorr), by=join_by(id)) # Inserts the sibling peptide correlation data
 sub_trace
})
class(pepTracesList_filtered) <- "tracesList"

#' Update traces with additional metrics for each fraction:
pepTracesList_filtered <- updateTraces(pepTracesList_filtered)

#' Inspect traces list:
summary(pepTracesList_filtered)

```



## - Global stats

```{r}
#' # Global stats
traces_names <- c(
  "pepTraces_raw",
  "pepTracesList_filtered"
)

countDT <- data.table(evidence=character(), sample=character(), count=integer(), traces=character())

for (traces in traces_names) {
  #traces <- readRDS(paste0(traces_name,".rda"))
  traces_name <- traces
  traces <- eval(as.name(traces))

  getIDs <- lapply(traces, function(x){
    peptides <- unique(x$trace_annotation[grep("DECOY",protein_id, invert = T)]$id)
    proteins <- unique(x$trace_annotation[grep("DECOY",protein_id, invert = T)]$protein_id)
    list(peptides=length(peptides),proteins=length(proteins))
  })

  count_dt <- as.data.table(getIDs)
  count_dt[, evidence := c("peptides", "proteins")]
  count_dt <- melt(count_dt, id.vars = "evidence", variable.name = "sample", value.name = "count")
  count_dt[,traces:=traces_name]
  count_dt[, count := unlist(count)]
  countDT <- rbind(countDT,count_dt)
}

traces_names_dt <- data.table(
  traces=c("pepTraces_raw",
           "pepTracesList_filtered"),
  traces_name=c("raw",
                "filtered")
)

countDT <- merge(countDT, traces_names_dt, by="traces")

countDT$traces_name <- factor(countDT$traces_name, levels=traces_names_dt$traces_name)

# write.table(countDT, "traces_stat_counts.txt", sep="t",quote = F, col.names = T, row.names = F)
# 
# pdf("traces_stat_counts.pdf",height=9,width=12)

g <- ggplot(countDT,aes(x=traces_name,y=count,fill=evidence, group=sample)) +
  geom_bar(stat="identity") +
     geom_label(label = countDT$count) +
  facet_wrap(sample ~ evidence, scales="free",nrow=3) +
  scale_fill_manual(values=c("#E69F00", "#56B4E9"))+
  theme_classic() +
  theme(legend.position="bottom") + theme(legend.title = element_blank()) +
  labs(y = "count") +
  theme(axis.title.x=element_blank()) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

print(g)


```


# 6. Protein quantification

> **Protein quantification.** Builds the protein-level `tracesList` from the
> filtered peptide traces using the lab "sibling/replicate peptide-correlation"
> fix (`proteinQuantification_sibPepCorrFix.tracesList`, in
> `R/ccprofiler_fixes.R`). This is the step that was missing from the original
> `.Rmd`; **Option A** is the confirmed one, **Option B** is a stock-CCprofiler
> alternative kept commented for reference.

```{r protein-quantification, message=FALSE, warning=FALSE}
# Option A (used): lab sibling/replicate peptide-correlation fix
# (defined in R/ccprofiler_fixes.R). It already returns an object of class "tracesList".
# NOTE: CCprofiler prints one BENIGN warning per protein that has >1 annotation row
# ("multiple entries for protein_id ...: . Picking the first") - it only affects
# which metadata row is kept, not the quantification. suppressWarnings() keeps it
# from flooding the console/report.
protein_traces_list <- suppressWarnings(proteinQuantification_sibPepCorrFix.tracesList(
  pepTracesList_filtered,
  topN       = 2,
  keep_less  = FALSE,
  rm_decoys  = TRUE,
  quantLevel = "protein_id"
))

# Option B (alternative): stock CCprofiler protein quantification. Uncomment to use.
# protein_traces_list <- proteinQuantification(pepTracesList_filtered, topN = 2)
# class(protein_traces_list) <- "tracesList"
```

# GLOBAL assembly state

## - Protein Mass distribution

```{r}

# evaluate protein mass distribution
protTraces_assembly <- annotateMassDistribution(protein_traces_list)


uniqueProteins <- Reduce(intersect, lapply(protTraces_assembly, function(x){x$trace_annotation$id}))

protTraces_assembly <- subset(protTraces_assembly, trace_subset_ids = uniqueProteins)

lapply(names(protTraces_assembly), function(x){
  summarizeMassDistribution(protTraces_assembly[[x]], 
                            PDF=F) #,
                            #name=paste0(here("output_files/massDistribution_",x)))
  })

```

## - Get assembly changes

```{r}
n_conditions <- length(unique(design_matrix$Condition))

#create a list of vectors to save the all_proteins_independentOfAssemblyState; proteins_noAssemblyChanges; more_assembled_proteins; less_assembled_proteins; proteins_withAssemblyChanges

all_proteins_independentOfAssemblyState_list <- list()
proteins_noAssemblyChange_list <- list()
more_assembled_proteins_list <- list()
less_assembled_proteins_list <- list()
proteins_withAssemblyChanges_list <- list()
  

# compare the treatmets to the ref (the list is specified in the vectors treatment_conditions and ref_condition in the Analysis SETUP section)

# Cache the (slow) assembly test - getMassAssemblyChange runs betareg per protein, so cache it the same
# way as proteinFeatures/diff: when diffAssemblyState.RData exists and load_proteinFeatures.RData is
# TRUE, load it and skip the recompute. The categorisation below always re-runs on diffAssemblyState, so
# the cutoffs stay adjustable without recomputing betareg. Set load_proteinFeatures.RData <- F (or delete
# the file) to force a fresh assembly test.
.assembly_file <- file.path(out_dir, "rdata", "diffAssemblyState.RData")
.brk("assembly test (getMassAssemblyChange, betareg per protein) ...")
if (load_proteinFeatures.RData && file.exists(.assembly_file)) {
  load(.assembly_file)   # restores diffAssemblyState from a previous run
  message("Loaded cached diffAssemblyState from ", .assembly_file, " (skipped getMassAssemblyChange).")
} else {
diffAssemblyState <- getMassAssemblyChange_aljazfix(tracesList=protTraces_assembly,
                                             design_matrix= design_matrix,
                                             compare_between = "Condition",
                                             quantLevel = "protein_id",
                                             plot=plot_individual_traces,   # was T: one plot PER PROTEIN dominated this chunk's runtime
                                             PDF=F #,
                                             #name = paste0("beta_pvalue_histogram_"ref_condition,"_",i)
                                             )
  save(diffAssemblyState, file = .assembly_file)
  message("Saved diffAssemblyState to ", .assembly_file, " (re-runs with load_proteinFeatures.RData=T skip getMassAssemblyChange).")
}
.brk("assembly test DONE")

# ---- categorize proteins by assembly change (treatment vs reference) ----
# FIX: more_assembled_proteins_list / less_assembled_proteins_list were declared
# above but never populated (that is why they were empty). Populate them here.
# meanDiff follows $testOrder, and the within-protein sample order can vary, so
# convert it to a fixed treatment-minus-reference difference => unambiguous sign.
.treat <- treatment_conditions[1]
diffAssemblyState[, treat_minus_ref := ifelse(
  testOrder == paste0(.treat, ".vs.", ref_condition),  meanDiff,
  ifelse(testOrder == paste0(ref_condition, ".vs.", .treat), -meanDiff, NA_real_))]

assembly_padj_cutoff   <- 0.05   # set to 1 to ignore significance (use effect size only)
assembly_effect_cutoff <- 0.1    # minimum |change| in assembled-mass fraction

sig_assembly <- diffAssemblyState[!is.na(treat_minus_ref) &
                                  betaPval_BHadj < assembly_padj_cutoff &
                                  abs(treat_minus_ref) > assembly_effect_cutoff]

more_assembled_proteins_list      <- sig_assembly[treat_minus_ref > 0]$protein_id  # more assembled in treatment
less_assembled_proteins_list      <- sig_assembly[treat_minus_ref < 0]$protein_id  # less assembled in treatment
proteins_withAssemblyChanges_list <- sig_assembly$protein_id

message("Assembly changes (padj < ", assembly_padj_cutoff, ", |delta| > ", assembly_effect_cutoff,
        "): ", length(more_assembled_proteins_list), " more-assembled, ",
        length(less_assembled_proteins_list), " less-assembled in ", .treat, " vs ", ref_condition,
        ". (If both are 0, relax assembly_padj_cutoff/assembly_effect_cutoff above.)")

# Persist the assembly hit-sets so the later cross-comparison overlap step
# (scripts/overlap_between_comparisons.R) can load them for every comparison.
saveRDS(list(more_assembled = more_assembled_proteins_list,
             less_assembled = less_assembled_proteins_list,
             changed        = proteins_withAssemblyChanges_list),
        file.path(out_dir, "rdata", "assembly_hits.rds"))

```

## - look-up traces of proteins WITH global assembly changes

LET'S have a look at these proteins that change in assembly in at least
one of the treatments:

```{r}
 # lookup all the proteins_withAssembly Changes in any of the treatment conditions


for(i in filter(diffAssemblyState, betaPval_BHadj < 0.05 & abs(meanDiff) > 0.1)$protein_id){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i)
} #not saved in the nas

```

```{r filter not by pval, but by the strength of the difference}
for(i in filter(diffAssemblyState, abs(meanDiff) > 0.1)$protein_id){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i)
}
```


actually, the other way around => fix this by inverting the factor
order for ATP and ctrl. Proteins that are LESS assembled

```{r}

#get all the proteins in the list
all_elements_less_assembled_proteinslist <- unique(unlist(less_assembled_proteins_list))

# lookup all the proteins_withAssembly Changes in any of the treatment conditions

for(i in all_elements_less_assembled_proteinslist){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i) #not saved in the nas
}

```

Proteins that are MORE assembled

```{r}

#get all the proteins in the list
all_elements_more_assembled_proteinslist <- unique(unlist(more_assembled_proteins_list))

# lookup all the proteins_withAssembly Changes in any of the treatment conditions

for(i in all_elements_more_assembled_proteinslist){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i) #not saved in the nas
}

```

`

# DISTINCT assembly state ~ PROTEIN level analysis

### - filter for prot with min N pep

```{r}
## subset to keep only the proteins with at least 5 precursors 
pepTracesList_filtered_minN <- subset(pepTracesList_filtered, trace_subset_ids = uniqueProteins, trace_subset_type = "protein_id")
```

## - performProteinGridSearch

```{r perform_prot_grid_search}

# CACHE GATE (handover fix): this chunk used to re-run the whole search on EVERY render whenever
# perform_prot_grid_search was TRUE. The result was written to ProtGridSearch.RData by the next chunk,
# but nothing ever read it back while the flag stayed TRUE - so the most expensive step in the pipeline
# repeated itself on every re-render for no gain. It is now skipped when a saved search exists for THIS
# comparison. Set force_prot_grid_search <- T (or delete the file) to recompute after changing the
# upstream data, the filtering, or the grid.
.gridsearch_file <- file.path(out_dir, "ProtGridSearch.RData")
if (!exists("force_prot_grid_search")) force_prot_grid_search <- F

# BORROW A SEARCH FROM ANOTHER COMPARISON. The optimal feature-finding parameters describe the
# CHROMATOGRAPHY - peak width, smoothing, correlation structure - not the biological contrast, so a
# search run on one comparison of the same SEC setup is a legitimate starting point for another and
# saves the single most expensive step. Set grid_search_from <- "QTL_83_RM" (any comparison_id whose
# ProtGridSearch.RData exists) to copy it in. Flag it in the methods if you use it.
if (!exists("grid_search_from")) grid_search_from <- NULL
if (!is.null(grid_search_from) && !file.exists(.gridsearch_file)) {
  .src <- here::here("output", grid_search_from, "ProtGridSearch.RData")
  if (file.exists(.src)) {
    file.copy(.src, .gridsearch_file)
    message("Borrowed the grid search from comparison '", grid_search_from, "' -> ", .gridsearch_file,
            ". These parameters describe the chromatography, not the contrast; state this in the methods.")
  } else message("grid_search_from = '", grid_search_from, "' but ", .src, " does not exist - ignoring.")
}

.run_grid <- isTRUE(perform_prot_grid_search) &&
             (isTRUE(force_prot_grid_search) || !file.exists(.gridsearch_file))
if (isTRUE(perform_prot_grid_search) && !.run_grid)
  message("Grid search SKIPPED: reusing ", .gridsearch_file,
          " (delete it, or set force_prot_grid_search <- T, to recompute).")

if(.run_grid == T) {
# NOTE (faithful fix): the original sampled from `pepTracesList_filtered_minN$ctrl_1` and then set
# `grid_peptideTraces <- grid_peptideTraces$ctrl_1`. This design has NO condition named "ctrl_1"
# (conditions are strainBY/strainRM/strain31/strain83), so both returned NULL and the search could
# not run. The real findProteinFeatures call below runs on traces INTEGRATED across all conditions,
# so the parameter search should use the same kind of object. performProteinGridSearch() wants a
# single `traces` object (not a tracesList), so: sample 10% of proteins, integrate that subset the
# same way as the real run, then search on that single integrated object.

# protein universe is identical across list elements; use the first (robust to condition naming)
.grid_prot_ids <- unique(pepTracesList_filtered_minN[[1]]$trace_annotation$protein_id)

# ---- SPEED CONTROLS -------------------------------------------------------------------------------
# The grid search is THE slow step: it runs findProteinFeatures once per parameter combination, so
# cost ~= (#corrs * #windows * #smoothing * #rt_heights) full feature-finding runs, split over cores.
# The original FULL grid was 3*4*3*3 = 108 runs (~45 min on ~5 cores). Three levers make it faster:
#   (1) fewer combinations  - biggest effect. The reduced grid below brackets the same ranges and
#       still INCLUDES the lab's preferred values (corr 0.9, window 7, smoothing 7, rt_height 1),
#       so the search can still land on them.
#   (2) more cores          - n_cores below (faster up to #combinations).
#   (3) smaller subset      - lower grid_subset_fraction (faster but noisier q-values).
# It also only needs to run ONCE per dataset: afterwards set perform_prot_grid_search <- F to reuse
# the saved ProtGridSearch.RData instead of recomputing.
# FLAG: the reduced grid is a handover change, NOT in the original. To restore the exhaustive search,
# comment the "reduced" block and uncomment the "FULL (original)" block below.
grid_subset_fraction <- 0.10   # 0.10 = 10% of proteins; try 0.05 to roughly halve the time

# reduced grid (default): 3 * 2 * 2 * 2 = 24 runs
grid_corrs      <- c(0.75, 0.9, 0.95)   # corr_cutoff - kept at 3 levels (most influential parameter)
grid_windows    <- c(7, 12)             # window_size      (full grid was 5, 7, 10, 12)
grid_smoothing  <- c(7, 11)             # smoothing_length (full grid was 7, 9, 11)
grid_rt_heights <- c(1, 4)              # rt_height        (full grid was 1, 4, 5)
# FULL (original) grid: 3 * 4 * 3 * 3 = 108 runs - uncomment to restore the exhaustive search:
# grid_corrs      <- c(0.75, 0.9, 0.95)
# grid_windows    <- c(5, 7, 10, 12)
# grid_smoothing  <- c(7, 9, 11)
# grid_rt_heights <- c(1, 4, 5)

# random subset (seed added for reproducibility - flag: NOT in the original, remove if unwanted)
set.seed(123)
num_ids    <- ceiling(length(.grid_prot_ids) * grid_subset_fraction)
subset_ids <- sample(.grid_prot_ids, num_ids)

grid_subset <- subset(pepTracesList_filtered_minN,
                        trace_subset_ids = subset_ids,
                        trace_subset_type = "protein_id")

# integrate the subset across conditions -> single `traces` object (same call as pepTracesIntegrated_sum)
grid_peptideTraces <- integrateTraceIntensities(grid_subset,
                                                 design_matrix   = NULL,
                                                 integrate_within = NULL,
                                                 aggr_fun        = "sum")

# performProteinGridSearch parallelises over the parameter combinations (n_cores); use up to 8.
.grid_cores <- parallel::detectCores()
.grid_cores <- if (is.na(.grid_cores) || .grid_cores < 2L) 1L else min(.grid_cores - 1L, 8L)
.grid_ncomb <- length(grid_corrs) * length(grid_windows) * length(grid_smoothing) * length(grid_rt_heights)
message("Grid search: ", .grid_ncomb, " parameter combinations x ", num_ids,
        " proteins on ", .grid_cores, " core(s). Runs once; the result is cached and reused automatically.")
.brk(paste0("grid search START: ", .grid_ncomb, " combinations x ", num_ids, " proteins on ",
            .grid_cores, " core(s)"))
.grid_t0 <- Sys.time()

# Wrapped in tryCatch: performProteinGridSearch uses a parallel cluster (snow), and if ANY single
# parameter combination errors inside CCprofiler (e.g. the "object 'merged' not found" edge case),
# the whole call aborts. Rather than killing the knit after a long run, fall back to default params.
proteinFeatures_grid <- tryCatch(
  performProteinGridSearch(traces     = grid_peptideTraces,
                           corrs      = grid_corrs,      #corr_cutoff (0.9 for Isabel)
                           windows    = grid_windows,    #window_size (7 for Isabel)
                           smoothing  = grid_smoothing,  #smoothing_length (7 for Isabel)
                           rt_heights = grid_rt_heights, #rt_height (1 for Isabel)
                           n_cores    = .grid_cores),
  error = function(e) {
    message("performProteinGridSearch failed: ", conditionMessage(e))
    message("  -> skipping parameter tuning; findProteinFeatures below falls back to default params.")
    message("  -> to pinpoint the offending combination, shrink the grid vectors above or set ",
            ".grid_cores <- 1L (serial) and re-run this chunk.")
    NULL
  }
)

.brk(paste0("grid search DONE in ", round(as.numeric(difftime(Sys.time(), .grid_t0, units = "mins")), 1), " min"))
if (!is.null(proteinFeatures_grid)) lapply(proteinFeatures_grid, head, n = 2) }
```

```{r proteinFeatures_bestStats}
# Score the grid, pick the parameter set with the most features at 5% FDR, and expose the winning
# parameters (best_corr_cutoff / best_window_size / best_smoothing_length / best_rt_height) so the
# findProteinFeatures call below can USE them. Previously this chunk was eval=FALSE, so its result
# was never fed forward and the real run used hard-coded parameters regardless of the search.

# .gridsearch_file is defined in the previous chunk (the cache gate); re-stated here so this chunk can
# still be run on its own.
if (!exists(".gridsearch_file")) .gridsearch_file <- file.path(out_dir, "ProtGridSearch.RData")
if (!exists(".run_grid")) .run_grid <- FALSE

# Score ONLY when the grid actually ran in this render. Otherwise fall through to loading the cache.
if (exists("proteinFeatures_grid") && !is.null(proteinFeatures_grid)) {
  # scoring wrapped too: if the grid ran but scoring/q-value estimation fails, don't kill the knit
  tryCatch({
    proteinFeatures_scoredData <- lapply(proteinFeatures_grid, calculateCoelutionScore)
    proteinFeatures_qvalueData <- lapply(proteinFeatures_scoredData, calculateQvalue, plot = FALSE)
    proteinFeatures_stats      <- qvaluePositivesPlotGrid(proteinFeatures_qvalueData)
    proteinFeatures_bestStats  <- getBestQvalueParameters(proteinFeatures_stats, FDR_cutoff = 0.05)
    print(proteinFeatures_bestStats)
    save(proteinFeatures_grid, proteinFeatures_bestStats, file = .gridsearch_file)
  }, error = function(e)
    message("Grid scoring failed (", conditionMessage(e), "); findProteinFeatures will use defaults."))
} else if (file.exists(.gridsearch_file)) {
  # reuse a previous run's saved parameters. Load into a temp environment and take ONLY bestStats:
  # the file also holds proteinFeatures_grid (one full feature set per parameter combination), which is
  # large and not needed downstream.
  .ge <- new.env(); load(.gridsearch_file, envir = .ge)
  if (!is.null(.ge$proteinFeatures_bestStats)) {
    proteinFeatures_bestStats <- .ge$proteinFeatures_bestStats
    message("Loaded cached grid-search parameters from ", .gridsearch_file, ".")
    print(proteinFeatures_bestStats)
  } else message("Cache ", .gridsearch_file, " has no proteinFeatures_bestStats - using defaults.")
  rm(.ge)
} else {
  message("No grid-search results available - findProteinFeatures will fall back to default parameters.")
}
# (to reuse a pre-computed search instead, point .gridsearch_file at e.g. ProtGridSearch_fromJP02.RData)

# --- pull the winning parameters out of bestStats, defensively --------------------------------------
# getBestQvalueParameters() returns a single row; CCprofiler has named these columns slightly
# differently across versions, so look each one up under its known aliases and fall back to the
# Isabel/PPlab defaults (= the values previously hard-coded below) if a column or the whole object is
# missing. This keeps the run alive even if the grid search produced nothing usable.
.bs <- if (exists("proteinFeatures_bestStats")) proteinFeatures_bestStats else NULL

.gridParam <- function(stats, aliases, default) {
  if (is.null(stats) || nrow(stats) == 0) return(default)
  hit <- intersect(aliases, colnames(stats))
  if (length(hit) == 0) return(default)
  val <- stats[[hit[1]]][1]
  if (length(val) == 0 || is.na(val)) default else val
}

best_corr_cutoff      <- .gridParam(.bs, c("corr", "corr_cutoff"),           0.85)
best_window_size      <- .gridParam(.bs, c("window", "window_size"),         7)
best_smoothing_length <- .gridParam(.bs, c("smoothing_length", "smoothing"), 3)
best_rt_height        <- .gridParam(.bs, c("rt_height", "rt_heights"),       1)

message("findProteinFeatures will use: corr_cutoff = ", best_corr_cutoff,
        ", window_size = ", best_window_size,
        ", smoothing_length = ", best_smoothing_length,
        ", rt_height = ", best_rt_height,
        if (isTRUE(.run_grid)) "  (chosen by the grid search that ran in THIS render)"
        else if (exists("proteinFeatures_bestStats")) "  (loaded from the cached grid search)"
        else "  (fallback defaults - no grid search available)")

# Persist the chosen parameters to a small human-readable file so they survive after the console
# scrolls (the message above is console-only). Lands in output/<comparison_id>/output_files/.
writeLines(c(
  paste0("findProteinFeatures parameters | comparison_id = ", comparison_id),
  paste0("source           = ", if (perform_prot_grid_search) "grid search" else "loaded / default"),
  paste0("corr_cutoff      = ", best_corr_cutoff),
  paste0("window_size      = ", best_window_size),
  paste0("smoothing_length = ", best_smoothing_length),
  paste0("rt_height        = ", best_rt_height),
  paste0("written          = ", format(Sys.time()))
), file.path(output_files_location, "grid_best_params.txt"))

```

## - findProteinFeatures

```{r}
#' # Integrate peptide traces across conditions 
# per each oe 
# --- progress breadcrumbs -------------------------------------------------------------------------
# findProteinFeatures is inside suppressMessages() and knitr buffers messages, so under rmarkdown::render
# you cannot SEE which step stalls. .brk() appends a timestamped line to out_dir/render_progress.log
# (file writes are not buffered), so while the render runs you open that file and the LAST line is the
# step it is currently on. Remove these once the slow/hanging step is identified.
.brk <- function(msg) { try(cat(format(Sys.time(), "%H:%M:%S"), "|", msg, "\n",
                               file = file.path(out_dir, "render_progress.log"), append = TRUE), silent = TRUE) }
.brk("integrateTraceIntensities (sum) ...")
pepTracesIntegrated_sum <- integrateTraceIntensities(pepTracesList_filtered_minN,
                                                     design_matrix = NULL,
                                                     integrate_within = NULL,
                                                     aggr_fun = "sum")

#' # Perform protein-centric analysis
# Suppress messages for clarity

# Guard: if the grid-search chunks above were skipped entirely, set best_* to the original defaults
# so this call still runs (normally best_* come from the proteinFeatures_bestStats chunk).
if (!exists("best_corr_cutoff"))      best_corr_cutoff      <- 0.85
if (!exists("best_window_size"))      best_window_size      <- 7
if (!exists("best_smoothing_length")) best_smoothing_length <- 3
if (!exists("best_rt_height"))        best_rt_height        <- 1

# findProteinFeatures is already parallelised; use up to 6 cores but never oversubscribe a smaller
# machine (each worker copies the traces, so too many cores on a laptop causes memory thrash and can
# be SLOWER, not faster). Raise the cap below if you are on a big-memory workstation.
.pf_cores <- parallel::detectCores()
.pf_cores <- if (is.na(.pf_cores) || .pf_cores < 2L) 1L else min(.pf_cores - 1L, 6L)

# CACHE GATE (handover fix): proteinFeatures.RData was being SAVED a few lines below but never LOADED,
# so this ~30-40 min step re-ran on every render even though its result was already sitting on disk -
# the same defect the grid search had. The Ecoli report has had this guard for months; the yeast one
# never got it. Set load_proteinFeatures.RData <- F, or delete the file, to force a fresh run after
# changing the upstream data, the filtering, or the feature-finding parameters.
.pf_file <- file.path(out_dir, "rdata", "proteinFeatures.RData")
if (load_proteinFeatures.RData && file.exists(.pf_file)) {
  load(.pf_file)   # restores `proteinFeatures` from the previous run
  message("Loaded cached proteinFeatures from ", .pf_file, " (skipped findProteinFeatures).")
  .brk("findProteinFeatures SKIPPED (loaded from cache)")
} else {
.brk(paste0("findProteinFeatures starting (parallelized=T, n_cores=", .pf_cores, ") -- if the log stops HERE, the parallel cluster is hanging under render"))
suppressMessages(
  proteinFeatures <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff      = best_corr_cutoff,      # grid search (default 0.85)
                                          window_size      = best_window_size,      # grid search (default 7)
                                          parallelized     = T,
                                          n_cores          = .pf_cores,
                                          collapse_method  = "apex_only",
                                          perturb_cutoff   = "5%",
                                          rt_height        = best_rt_height,        # grid search (default 1)
                                          smoothing_length = best_smoothing_length  # grid search (default 3)

)
)
}

# --- OPTIONAL: alternative-parameter exploration runs (DISABLED by default) -----------------------
# The three findProteinFeatures calls below re-run feature finding at lower corr_cutoff (0.7 / 0.5 /
# 0.4). They are NOT used anywhere downstream (parameter-exploration leftovers), and the low cutoffs
# make them VERY slow - the 0.4/0.5 runs dominate runtime and are what made this chunk run overnight.
# Set explore_alt_params <- TRUE only if you specifically want to inspect these alternatives.
explore_alt_params <- FALSE
if (explore_alt_params) {

suppressMessages(
  proteinFeatures_0p7_corr_cutoff <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff = 0.7, # 0.9,
                                          window_size = 7,# 7,
                                          parallelized = T,
                                          n_cores = 6,
                                          collapse_method = "apex_only",
                                          perturb_cutoff = "5%",
                                          rt_height = 1, #1,
                                          smoothing_length = 3 #7)

)
)

suppressMessages(
  proteinFeatures_0p5_corr_cutoff_winsize5 <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff = 0.5, # 0.9,
                                          window_size = 5,# 7,
                                          parallelized = T,
                                          n_cores = 6,
                                          collapse_method = "apex_only",
                                          perturb_cutoff = "5%",
                                          rt_height = 1, #1,
                                          smoothing_length = 3 #7)

)
)

suppressMessages(
  proteinFeatures_0p4_corr_cutoff_winsize5 <-  findProteinFeatures(traces = pepTracesIntegrated_sum,
                                          corr_cutoff = 0.4, # 0.9,
                                          window_size = 5,# 7,
                                          parallelized = T,
                                          n_cores = 6,
                                          collapse_method = "apex_only",
                                          perturb_cutoff = "5%",
                                          rt_height = 1, #1,
                                          smoothing_length = 3 #7)

)
)


} # end if (explore_alt_params) -- alt-parameter comparison runs, off by default

.brk("findProteinFeatures DONE -> save + scoreFeatures (FDR 0.05) ...")
# only write the cache when it was actually recomputed; rewriting a just-loaded object is pure I/O
if (!file.exists(.pf_file)) {
  dir.create(dirname(.pf_file), recursive = TRUE, showWarnings = FALSE)
  save(proteinFeatures, file = .pf_file)
  message("Saved proteinFeatures to ", .pf_file, " (re-runs with load_proteinFeatures.RData = T skip findProteinFeatures).")
}

# Reuse the whole serial tail if a previous run of this comparison already produced it. These three
# CCprofiler steps (scoreFeatures, extractFeatureVals, fillFeatureVals) are single-threaded inside the
# package and are the slowest part of the render after findProteinFeatures itself.
.featvals_file <- file.path(out_dir, "rdata", "featureVals.RData")
.featvals_part <- file.path(out_dir, "rdata", "featureVals_partial.RData")
if (load_proteinFeatures.RData && file.exists(.featvals_file)) {
  load(.featvals_file)   # -> proteinFeaturesFiltered, *_featureVals, *_featureValsFilled
  message("Loaded cached feature values from ", .featvals_file,
          " (skipped scoreFeatures / extractFeatureVals / fillFeatureVals).")
} else if (load_proteinFeatures.RData && file.exists(.featvals_part)) {
  # A previous render got through the 8.6 h extraction but was interrupted before the end. Resume from
  # that checkpoint: only fillFeatureVals (minutes) still has to run.
  load(.featvals_part)   # -> proteinFeaturesFiltered, proteinFeaturesFiltered_featureVals
  message("Resuming from the extractFeatureVals checkpoint ", .featvals_part,
          " - only fillFeatureVals still needs to run.")
  .brk("resumed from extractFeatureVals checkpoint -> fillFeatureVals")
  proteinFeaturesFiltered_featureValsFilled <- fillFeatureVals(
    featureVals = proteinFeaturesFiltered_featureVals,
    tracesList  = pepTracesList_filtered_minN,
    design_matrix = design_matrix)
  save(proteinFeaturesFiltered, proteinFeaturesFiltered_featureVals,
       proteinFeaturesFiltered_featureValsFilled, file = .featvals_file)
} else {

#' Score protein features and filter for a detection FDR of 5%
proteinFeaturesFiltered <- scoreFeatures(proteinFeatures, FDR=0.05, PDF=TRUE, name=file.path(output_files_location, "qvalueStats_proteinFeatures"))

#' Evaluate protein feature stats
summarizeFeatures(proteinFeaturesFiltered,
                  plot=TRUE,
                  PDF = F)
                  # PDF=TRUE,
                  # name="output_files/feature_summary_proteinFeaturesFiltered")

# checkpoint the scored features immediately - cheap to save, and it means an interrupt during the
# very long extraction below does not also cost the scoring
save(proteinFeaturesFiltered, file = file.path(out_dir, "rdata", "proteinFeaturesFiltered.RData"))

#' ## Extract feature values
# THE SLOWEST STEP IN THE WHOLE PIPELINE. Measured on this dataset: 8.6 h, against 0.6 h for
# findProteinFeatures and 0.1 h for fillFeatureVals. It is single-threaded inside CCprofiler, and until
# now its result was only saved AFTER fillFeatureVals - so an interrupt anywhere in those 8.6 h threw
# the whole thing away and the next render started from zero. That is what happened here.
#
# The fix is to run it in CHUNKS of features and checkpoint each chunk. Extraction is per-feature, so
# splitting the feature table and row-binding the pieces returns the same object; the guard below
# verifies that the pieces really do share a structure, and falls back to one plain call if anything
# about the split looks wrong. The practical gain is resumability: a crash at hour 7 now resumes at
# hour 7. Delete rdata/featureVals_chunks/ to force a clean re-extraction.
.brk("extractFeatureVals (FDR 0.05) -- SERIAL CCprofiler step, slow on big data ...")
.fv_chunk_dir <- file.path(out_dir, "rdata", "featureVals_chunks")
dir.create(.fv_chunk_dir, recursive = TRUE, showWarnings = FALSE)
if (!exists("featureval_chunks")) featureval_chunks <- 20L   # more chunks = finer checkpoints

.extract_chunked <- function() {
  FT <- proteinFeaturesFiltered
  idcol <- intersect(c("protein_id", "complex_id", "feature_id"), names(FT))[1]
  if (is.na(idcol)) stop("no id column to split on")
  ids <- unique(as.character(FT[[idcol]]))
  if (length(ids) < 2 * featureval_chunks) stop("too few features to be worth chunking")
  # The split depends on featureval_chunks, so cached chunks are only valid for the SAME setting.
  # Changing it silently would mix incompatible splits and lose or duplicate features.
  .marker <- file.path(.fv_chunk_dir, "chunk_count.txt")
  if (file.exists(.marker)) {
    prev <- suppressWarnings(as.integer(readLines(.marker, warn = FALSE)[1]))
    if (is.finite(prev) && prev != featureval_chunks) {
      message("featureval_chunks changed from ", prev, " to ", featureval_chunks,
              " - the cached chunks use the OLD split and cannot be mixed. Discarding them.")
      unlink(list.files(.fv_chunk_dir, pattern = "^chunk_.*\\.rds$", full.names = TRUE))
    }
  }
  writeLines(as.character(featureval_chunks), .marker)
  grp <- split(ids, cut(seq_along(ids), breaks = featureval_chunks, labels = FALSE))
  out <- vector("list", length(grp)); t0 <- Sys.time(); .first_new <- TRUE
  for (k in seq_along(grp)) {
    cf <- file.path(.fv_chunk_dir, sprintf("chunk_%03d.rds", k))
    if (file.exists(cf)) { out[[k]] <- readRDS(cf); next }
    sub <- FT[as.character(FT[[idcol]]) %in% grp[[k]], ]
    # SUBSET THE TRACES TOO - this is the whole point. Measured on this dataset: passing the FULL trace
    # list to each chunk costs ~88 min per call whatever the chunk size, because the per-call work scales
    # with the TRACES, not with the features. Twenty such calls took 30 h against 8.6 h for one plain
    # call, i.e. chunking made it 3.5x worse. Restricting the traces to the chunk's own proteins removes
    # that fixed cost, so each chunk is roughly 1/n of the work and the total finally drops.
    sub_traces <- tryCatch(subset(pepTracesList_filtered_minN,
                                  trace_subset_ids = grp[[k]], trace_subset_type = "protein_id"),
                           error = function(e) { message("  (trace subsetting failed: ", conditionMessage(e),
                                                         " - using the full trace list for this chunk)")
                                                 pepTracesList_filtered_minN })
    out[[k]] <- extractFeatureVals(traces = sub_traces,
                                   features = sub, design_matrix = design_matrix,
                                   extract = "subunits_detected", imputeZero = T, verbose = F)
    saveRDS(out[[k]], cf)
    .brk(sprintf("  extractFeatureVals chunk %d/%d done (%.1f min elapsed)", k, length(grp),
                 as.numeric(difftime(Sys.time(), t0, units = "mins"))))
    # Report the cost of the FIRST chunk computed in this render, so the trace-subsetting speedup is
    # visible immediately instead of after another overnight run.
    if (.first_new) {
      .mins <- as.numeric(difftime(Sys.time(), t0, units = "mins")); .first_new <- FALSE
      .brk(sprintf("  -> first new chunk took %.1f min; %d chunk(s) left => about %.1f h to go",
                   .mins, length(grp) - k, .mins * (length(grp) - k) / 60))
      if (.mins > 30)
        .brk("  -> STILL SLOW: trace subsetting did not help. Set featureval_chunks <- 1 and delete rdata/featureVals_chunks/ to fall back to one plain call (~8.6 h).")
    }
  }
  out <- out[!vapply(out, is.null, logical(1))]
  if (!length(out)) stop("every chunk came back empty")
  nm <- lapply(out, names)
  if (length(unique(vapply(nm, function(z) paste(z, collapse = "|"), character(1)))) != 1L)
    stop("chunks disagree on their columns - not safe to combine")
  res <- data.table::rbindlist(out, use.names = TRUE, fill = FALSE)
  if (inherits(out[[1]], "data.table")) res else as.data.frame(res)
}

proteinFeaturesFiltered_featureVals <- tryCatch(.extract_chunked(), error = function(e) {
  message("Chunked extraction not used (", conditionMessage(e), "); running the single plain call.")
  .brk("extractFeatureVals -- falling back to a single plain call")
  extractFeatureVals(traces = pepTracesList_filtered_minN,
                     features = proteinFeaturesFiltered,
                     design_matrix = design_matrix,
                     extract = "subunits_detected", imputeZero = T, verbose = F)
})
# checkpoint before fillFeatureVals, so those hours can never be lost again
save(proteinFeaturesFiltered, proteinFeaturesFiltered_featureVals,
     file = file.path(out_dir, "rdata", "featureVals_partial.RData"))
.brk("extractFeatureVals DONE and checkpointed")


#' ## Fill feature values
.brk("fillFeatureVals (FDR 0.05) -- SERIAL CCprofiler step, slow on big data ...")
proteinFeaturesFiltered_featureValsFilled <- fillFeatureVals(featureVals = proteinFeaturesFiltered_featureVals,
                                                tracesList = pepTracesList_filtered_minN,
                                                design_matrix = design_matrix)

if (run_feature_fdr015) {
.brk("second pass at FDR 0.15 (scoreFeatures + extractFeatureVals + fillFeatureVals AGAIN) ...")
proteinFeaturesFiltered_0p15featurefdr <- scoreFeatures(proteinFeatures, FDR=0.15, PDF=TRUE, name=file.path(output_files_location, "qvalueStats_proteinFeatures"))

#' Evaluate protein feature stats
summarizeFeatures(proteinFeaturesFiltered_0p15featurefdr,
                  plot=TRUE,
                  PDF = F)
                  # PDF=TRUE,
                  # name="output_files/feature_summary_proteinFeaturesFiltered")

#' ## Extract feature values
proteinFeaturesFiltered_featureVals_0p15featurefdr <- extractFeatureVals(traces = pepTracesList_filtered_minN,
                                             features = proteinFeaturesFiltered_0p15featurefdr,
                                             design_matrix = design_matrix,
                                             extract = "subunits_detected",
                                             imputeZero = T,
                                             verbose = F)


#' ## Fill feature values
proteinFeaturesFiltered_featureValsFilled_0p15featurefdr <- fillFeatureVals(featureVals = proteinFeaturesFiltered_featureVals_0p15featurefdr,
                                                tracesList = pepTracesList_filtered_minN,
                                                design_matrix = design_matrix)
} else message("Skipping the feature-FDR 0.15 second pass (run_feature_fdr015 = F): it repeats the whole serial tail and feeds only the unused *_0p15featurefdr objects.")
.brk("findProteinFeatures chunk COMPLETE (feature values done)")

# Cache the expensive results of this chunk so an interrupted or repeated render skips the serial
# scoreFeatures -> extractFeatureVals -> fillFeatureVals tail as well as findProteinFeatures itself.
.featvals_file <- file.path(out_dir, "rdata", "featureVals.RData")
save(proteinFeaturesFiltered, proteinFeaturesFiltered_featureVals, proteinFeaturesFiltered_featureValsFilled,
     file = .featvals_file)
message("Saved feature values to ", .featvals_file, " (re-runs with load_proteinFeatures.RData = T skip the serial tail).")

}   # end of the "compute the serial tail" branch

# save(proteinFeaturesFiltered_featureVals, proteinFeaturesFiltered_featureValsFilled, file = "Anna_proteinFeaturesFiltered_featureValsFilled.RData")


```

## - differential analysis

2 conditions, pairwise. NOTE: this is **not** a 1-replicate-only test -
testDifferentialExpression_beniFix() adapts to the number of replicates in
design_matrix: with >= 2 replicates it runs a Welch two-sample t-test across
replicates; with only 1 replicate it falls back to a fraction-wise t-test within
each feature. The chunk below prints which path is taken. (R/ccprofiler_fixes.R, Section D.)

" Differential analysis is performed in 5 steps: (i)
Peptide-levelintensities are computed for each protein feature and
sample.Missing values in single fractions, replicates, or conditions
areimputed by uniformly sampling values between zero and theminimum
detected signal of a peptide. The peptide intensity ofone feature is
then calculated by summing the intensities of allfractions across the
corresponding protein feature range. (ii)The mean intensity across all
replicates within a condition(specified by the design matrix) is
calculated. (iii) The log2-fold-change between conditions is calculated
based on themean feature intensities. (iv) If replicates are available,
p-values are estimated by comparing the summed intensities across
conditions by a non paired t test. If no replicates are
available,p-values are estimated by comparing each fraction within a
feature by a paired t test across the conditions. (v) To subsequently
derive protein-level information, the peptide-level tests are aggregated
as follows: (1) protein log2-fold-changesare derived from the median
log-2-fold change across alldetected peptides of the protein, and (2)
protein p-values are estimated by determining the fold-change adjusted
median p-value and applying a beta distribution as described by Teo
etal.43 and Suomi et al.44 (for details see the methods section). (vi)
Multiple testing correction is performed by Benjamini-Hochberg
adjustment of the protein-level p-values

```{r diff}

# Report which differential test will run: the function branches on the replicate count in
# design_matrix (the same check it uses internally), so this prints the actual path taken.
.n_rep <- length(unique(design_matrix$Replicate))
message("Differential test: ", .n_rep, " replicate(s) in design_matrix -> ",
        if (.n_rep > 1) "Welch two-sample t-test across replicates (replicate-based)."
        else "fraction-wise t-test within each feature (1-replicate fallback; global p-values = 1).")
          
.diff_file <- file.path(out_dir, "rdata", "protein_DiffExpr_all.RData")
if (load_protein_DiffExprPep_and_Protein && file.exists(.diff_file)) {
  load(.diff_file)   # restores the four protein_DiffExpr* objects saved by a previous run
  message("Loaded cached differential-test results from ", .diff_file, " (skipped the diff test; set load_protein_DiffExprPep_and_Protein <- F to recompute).")
} else {
            #' ## Perform peptide-level differential expression testing for all features
          
            protein_DiffExprPep <- testDifferentialExpression_beniFix(featureVals = proteinFeaturesFiltered_featureValsFilled,
                                                              compare_between = "Condition",
                                                              level = "peptide",
                                                              measuredOnly = FALSE) #also considers the imputed fractions
            # Append the df to the list

          
            
            #' ## Aggregate differential expression results to the protein level
            protein_DiffExprProtein <- aggregatePeptideTests(protein_DiffExprPep) 
            
            # the looser-FDR pass only exists when run_feature_fdr015 = T (it repeats the whole serial tail)
            if (exists("proteinFeaturesFiltered_featureValsFilled_0p15featurefdr")) {
            protein_DiffExprPep_0p15featurefdr <- testDifferentialExpression_beniFix(featureVals = proteinFeaturesFiltered_featureValsFilled_0p15featurefdr,
                                                              compare_between = "Condition",
                                                              level = "peptide",
                                                              measuredOnly = FALSE) #also considers the imputed fractions
            # Append the df to the list



            #' ## Aggregate differential expression results to the protein level
            protein_DiffExprProtein_0p15featurefdr <- aggregatePeptideTests(protein_DiffExprPep_0p15featurefdr)
            protein_DiffExprProtein_0p15featurefdr <- protein_DiffExprProtein_0p15featurefdr %>% left_join(select(uniprot, accession, protein_name, mass, gene_names, go_f, go_p, go_c, ft_act_site, ft_binding, cc_cofactor), by=join_by(feature_id == accession))
            }

protein_DiffExprProtein <- protein_DiffExprProtein %>% left_join(select(uniprot, accession, protein_name, mass, gene_names, go_f, go_p, go_c, ft_act_site, ft_binding, cc_cofactor), by=join_by(feature_id == accession))

  .diff_objs <- c("protein_DiffExprPep", "protein_DiffExprProtein")
  if (exists("protein_DiffExprPep_0p15featurefdr")) .diff_objs <- c(.diff_objs, "protein_DiffExprPep_0p15featurefdr", "protein_DiffExprProtein_0p15featurefdr")
  save(list = .diff_objs, file = .diff_file)
  message("Saved differential-test results to ", .diff_file, " (re-runs with load_protein_DiffExprPep_and_Protein = T skip the diff test).")
}
save(protein_DiffExprPep, file = file.path(out_dir, "rdata", "protein_DiffExprPep_list.RData"))
save(protein_DiffExprProtein, file = file.path(out_dir, "rdata", "protein_DiffExprProtein_list.RData"))
# save.image(file = file.path(out_dir, "rdata", "workspace_image.RData"))  # DISABLED: redundant full-workspace dump (all key objects are saved individually); large & slow. Uncomment only for a full-session checkpoint.

```

Global differential assessment:

In addition to the feature-specific differential analysis, global
differential assessment is performed by comparing integrated intensities
across the entire fractionation dimension, instead of restricting the
analysis to a feature-specific range. The same strategies as for
feature-specific estimation of log2-fold-changes and p-values are
performed. To assess whether the signal within a protein feature is
changing because of a global change in the protein’s expression or due
to a rearrangement of the proteins relative distribution across
different assembly states,an additional analysis step is available in
CCprofiler. Here, the relative feature-specific mass fraction (FMF) is
estimated by the fraction of the detected MS signal in the
feature-specific mass range relative to the total detected signal: FMF =
sum(intensity features)/sum(intensity global) A change in FMF is
subsequently estimated by the difference in the mean FMF across
conditions diffFMF= mean(FMF condA) - mean (FMF condB)

... check the paper for full explanation of the steps
<https://pubs.acs.org/doi/epdf/10.1021/acs.jproteome.3c00125?ref=article_openPDF>

A significant change in the FMF across conditions indicates that the
protein’s relative contribution to different distinct assembly states
has changed across the conditions, thus providing insights into protein
rewiring, which is not observable by global proteome analyses. In
contrast to complex-centric analyses, described in the following
section, protein-centric differential analysis enables the assessment of
changes in distinct protein assembly states independent of the actual
knowledge of the protein’s exact interaction partners

## - volcano plots and save lists

```{r volcano plots}
library(data.table)

#create unique df - keep as data.table
combined_protein_DiffExprPep_list <- unique(protein_DiffExprPep)

combined_protein_DiffExprProtein_list <- unique(protein_DiffExprProtein)

comparison_level <- unique(combined_protein_DiffExprPep_list$testOrder) 

# create vector list to append to
diffProteins_list <- list()
diffProteins_onlypadj_list <- list()
diffProteins_global_list <- list()

#' # Make volcano plots
for (i in comparison_level) {
  
  # Keep as data.table for plotVolcano compatibility
  protein_DiffExprPep_i <- combined_protein_DiffExprPep_list[testOrder == i]
  protein_DiffExprProtein_i <- combined_protein_DiffExprProtein_list[testOrder == i]
  
  #' General volcano plots - local
  plotVolcano(protein_DiffExprPep_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2,  
              name = file.path(output_files_location, paste0("pep_", i)))
  plotVolcano(protein_DiffExprProtein_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              name = file.path(output_files_location, paste0("prot_", i)))
  plotVolcano(protein_DiffExprProtein_i, PDF = F, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              name = paste0("prot_", i)) #to see here
  
  #' global
  plotVolcano(protein_DiffExprPep_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              level = "global", name = file.path(output_files_location, paste0("global_pep_", i)))
  plotVolcano(protein_DiffExprProtein_i, PDF = T, pBHadj_cutoff = 0.05, FC_cutoff = 2, 
              level = "global", name = file.path(output_files_location, paste0("global_prot_", i)))
  
  #' Get different proteins - using data.table syntax
  diffProteins_i <- unique(protein_DiffExprProtein_i[pBHadj < 0.05 & abs(medianLog2FC) > 1]$feature_id)
  
  diffProteins_i_onlypadj <- unique(protein_DiffExprProtein_i[pBHadj < 0.05]$feature_id)
  
  #save data
  write.table(diffProteins_i, file.path(output_files_location, paste0("diffProteins_", i, ".txt")),
              sep = "\t", quote = F, row.names = F, col.names = F)
  write.table(diffProteins_i_onlypadj, file.path(output_files_location, paste0("diffProteins_onlypadj_", i, ".txt")),
              sep = "\t", quote = F, row.names = F, col.names = F)
  
  diffProteins_global_i <- unique(protein_DiffExprProtein_i[global_pBHadj < 0.05 & abs(global_medianLog2FC) > 1]$feature_id)
  
  #save data
  write.table(diffProteins_global_i, file.path(output_files_location, paste0("diffProteins_global_", i, ".txt")),
              sep = "\t", quote = F, row.names = F, col.names = F)
  
  #' Volcano plots highlighting different Proteins          
  highlight_protein <- diffProteins_i
  plotVolcano(protein_DiffExprProtein_i, highlight = highlight_protein, 
              PDF = F, pBHadj_cutoff = 0.05)
  
  #append to the list
  diffProteins_list[[i]] <- diffProteins_i
  diffProteins_onlypadj_list[[i]] <- diffProteins_i_onlypadj
  diffProteins_global_list[[i]] <- diffProteins_global_i
}

```

## - per-replicate view + GO enrichment of the significant hits

Added for the handover. For the significant differential proteins of each comparison this shows
(1) a **per-replicate** view of feature intensity - one point per biological replicate, i.e. exactly
the spread the replicate-based t-test uses (ties back to the variance question) - and (2) a **GO
over-representation** bar plot. The GO test is a plain hypergeometric test on the `go_p`/`go_f`/`go_c`
annotations already fetched by `fetch_uniprot`, so it needs no extra packages and runs offline.

```{r sig_hits_perReplicate_and_GO, fig.width=9, fig.height=6}
# ---- controls --------------------------------------------------------------------------------
sig_source   <- diffProteins_list   # local, padj<0.05 AND |log2FC|>1 (= diffProteins_<cmp>.txt).
                                     # alternatives: diffProteins_onlypadj_list / diffProteins_global_list
go_column    <- "go_p"              # GO to test: "go_p" (process) / "go_f" (function) / "go_c" (component)
go_min_genes <- 2                   # min significant proteins per GO term to report
go_top_n     <- 15                  # number of top enriched terms to plot
max_facets   <- 24                  # cap on per-replicate protein panels (most significant first)

.featVals <- as.data.table(proteinFeaturesFiltered_featureValsFilled)
.protRes  <- as.data.table(protein_DiffExprProtein)

for (cmp in comparison_level) {

  prot_cmp <- .protRes[testOrder == cmp]
  sig_ids  <- unique(sig_source[[cmp]]); sig_ids <- sig_ids[!is.na(sig_ids)]
  if (length(sig_ids) == 0) { message("[", cmp, "] no significant hits - skipping."); next }
  if ("pBHadj" %in% names(prot_cmp))                       # order by significance when available
    sig_ids <- unique(prot_cmp[feature_id %in% sig_ids][order(pBHadj)]$feature_id)
  message("[", cmp, "] ", length(sig_ids), " significant proteins.")

  ## (1) per-replicate summed feature intensity (one point = one biological replicate) -----------
  per_rep <- tryCatch(
    .featVals[feature_id %in% sig_ids,
              .(feature_intensity = sum(intensity, na.rm = TRUE)),
              by = .(feature_id, Condition, Replicate)],
    error = function(e) { message("  per-replicate table failed: ", conditionMessage(e)); NULL })

  if (!is.null(per_rep) && nrow(per_rep) > 0) {
    ord <- sig_ids[sig_ids %in% per_rep$feature_id]
    if (length(ord) > max_facets) {
      message("  per-replicate: showing the ", max_facets, " most significant of ", length(ord), ".")
      ord <- ord[seq_len(max_facets)]
    }
    pr <- per_rep[feature_id %in% ord]; pr[, feature_id := factor(feature_id, levels = ord)]
    print(ggplot(pr, aes(Condition, feature_intensity)) +
            geom_boxplot(outlier.shape = NA, alpha = 0.3) +
            geom_point(aes(colour = factor(Replicate)),
                       position = position_jitter(width = 0.12, height = 0), size = 1.6) +
            scale_y_log10() + facet_wrap(~ feature_id, scales = "free_y") +
            labs(title = paste0("Per-replicate feature intensity - ", cmp),
                 y = "summed feature intensity (log10)", colour = "Replicate") +
            theme_bw() + theme(axis.text.x = element_text(angle = 30, hjust = 1)))
    per_rep_summary <- per_rep[, .(n_rep = .N,
                                   mean_intensity = mean(feature_intensity, na.rm = TRUE),
                                   cv_percent = 100 * sd(feature_intensity, na.rm = TRUE) /
                                                mean(feature_intensity, na.rm = TRUE)),
                               by = .(feature_id, Condition)]
    fwrite(per_rep_summary,
           file.path(output_files_location, paste0("perReplicate_summary_", cmp, ".txt")), sep = "\t")
  }

  ## (2) GO over-representation (hypergeometric): significant hits vs all tested proteins ---------
  go_tbl <- tryCatch({
    bg <- unique(prot_cmp[, .(feature_id, go = get(go_column))])
    bg <- bg[!is.na(go) & nzchar(go)]
    if (nrow(bg) < 5) stop("too few annotated background proteins")
    pt <- bg[, .(term = trimws(unlist(strsplit(go, ";")))), by = feature_id][nzchar(term)]
    N  <- length(unique(pt$feature_id))                                  # annotated background
    n  <- length(unique(pt$feature_id[pt$feature_id %in% sig_ids]))      # annotated foreground
    if (n < go_min_genes) stop("too few annotated significant proteins")
    ts <- pt[, .(K = uniqueN(feature_id),
                 k = uniqueN(feature_id[feature_id %in% sig_ids])), by = term][k >= go_min_genes]
    if (nrow(ts) == 0) stop("no GO term reaches go_min_genes significant proteins")
    ts[, pval := phyper(k - 1, K, N - K, n, lower.tail = FALSE)]         # P(overlap >= k)
    ts[, padj := p.adjust(pval, method = "BH")]
    ts <- ts[order(padj, -k)][seq_len(min(go_top_n, .N))]
    fwrite(ts, file.path(output_files_location, paste0("GOenrichment_", go_column, "_", cmp, ".txt")),
           sep = "\t")
    ts
  }, error = function(e) { message("  GO enrichment skipped: ", conditionMessage(e)); NULL })

  if (!is.null(go_tbl) && nrow(go_tbl) > 0) {
    go_tbl[, term_short := ifelse(nchar(term) > 60, paste0(substr(term, 1, 57), "..."), term)]
    print(ggplot(go_tbl, aes(reorder(term_short, -padj), -log10(padj), fill = k)) +
            geom_col() + coord_flip() +
            geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
            labs(title = paste0("GO ", go_column, " enrichment - ", cmp, " (", length(sig_ids), " hits)"),
                 x = NULL, y = "-log10(BH-adjusted p)", fill = "n hits") +
            theme_bw())
  }
}
```

```{r compare local vs global}

#### local vs global

# FIX: `Sample_id_treat` was never defined (undefined in the original too), so this chunk errored.
# It also filtered the sample-level `Sample` column (values like "strain31_1") against `ref_condition`,
# which is a CONDITION label ("strainBY"), so reference rows never matched. Reconstructed at the
# CONDITION level: iterate the treatment condition(s) and compare each to the reference condition
# (testLocalVsGlobal groups by Condition anyway). FLAG: reconstruction of broken original code -
# tell me if you actually meant a per-sample comparison.
Sample_id_treat <- treatment_conditions   # condition label(s), e.g. "strain31" (name kept for the loop)

# SPEED NOTE: testLocalVsGlobal is a CCprofiler PACKAGE function, so it can't be parallelised from
# here. These loops run only ~once (one treatment condition x one comparison), so the multi-hour
# runtime was entirely inside a single testLocalVsGlobal call with plot = TRUE, which renders ONE plot
# per feature - that is the whole cost (the test itself is fast) - and the result was never used
# (everything below was commented). Fix: plot = FALSE, and STORE the result so the chunk is useful.
local_vs_global_list     <- list()   # raw per-feature local-vs-global test table
local_vs_global_sig_list <- list()   # significant local-vs-global protein ids
for(i in Sample_id_treat) {
            
          sample_vector <- c(i, ref_condition) 
          
          #filter for just 0- and condition x
          proteinFeaturesFiltered_featureValsFilled_i <- proteinFeaturesFiltered_featureValsFilled %>%
            filter(Condition %in% sample_vector) %>%
            mutate(Condition = factor(Condition, levels = sample_vector)) %>%
            arrange(Condition)

  for(cond in diffProteins_list){
  
   testFilled_norm <- proteinFeaturesFiltered_featureValsFilled_i[feature_id %in% cond]

  local_vs_global_test <- testLocalVsGlobal_fix(featureVals = testFilled_norm,
                                                compare_between = "Condition",
                                                design_matrix = design_matrix,  # explicit (pkg fn read it as a global; invisible under render)
                                                plot = FALSE)  # was TRUE: 1 plot/feature was the entire bottleneck
  local_vs_global_list[[length(local_vs_global_list) + 1L]] <- local_vs_global_test
  # aggregate to protein level + keep the significant local-vs-global proteins (a feature-specific vs
  # global change = assembly-state rearrangement). Wrapped: aggregateLocalVsGlobal is a package fn.
  lvg_sig <- tryCatch(
    unique(as.data.table(aggregateLocalVsGlobal(local_vs_global_test))[pBHadj < 0.05 & abs(medianDiff) > 0.3]$feature_id),
    error = function(e) { message("  aggregateLocalVsGlobal skipped: ", conditionMessage(e)); NULL })
  if (!is.null(lvg_sig)) {
    local_vs_global_sig_list[[length(local_vs_global_sig_list) + 1L]] <- lvg_sig
    message("Local-vs-global: ", length(lvg_sig), " proteins with a significant shift.")
  }
  
  # local_vs_global_test_protein <- aggregateLocalVsGlobal(local_vs_global_test)
  # 
  # local_vs_global_test_sig <- unique(local_vs_global_test_protein[pBHadj<0.05][abs(medianDiff) > 0.3]$feature_id)
  # 
  # 
  # #plot
  # for (id in local_vs_global_test_sig) {
  #   plotFeatures(feature_table = proteinFeaturesFiltered,
  #                traces = pepTracesList_filtered_minN,
  #                feature_id = id,
  #                design_matrix=sub_design_matrix,
  #                calibration=calibration,
  #                peak_area = T,
  #                legend = F,
  #                onlyBest = F,
  #                monomer_MW=T,
  #                aggregateReplicates=F)
  # }
  }
}

saveRDS(local_vs_global_list,     file.path(out_dir, "rdata", "local_vs_global_list.rds"))
saveRDS(local_vs_global_sig_list, file.path(out_dir, "rdata", "local_vs_global_sig_list.rds"))
```

## - GO enrichment: assembly-change proteins & local-vs-global hits

GO over-representation (same offline hypergeometric test as the volcano-hits chunk) for two more
hit sets: the assembly-change proteins (more- / less- / any-assembled) and the significant
local-vs-global proteins. Uses the go_p/go_f/go_c annotations already fetched by fetch_uniprot;
background = the relevant tested universe (assembly-tested proteome / differential proteome).

```{r go_assembly_and_localglobal, fig.width=9, fig.height=6}
go_column    <- "go_p"   # GO to test: "go_p" (process) / "go_f" (function) / "go_c" (component)
go_min_genes <- 2        # min hit proteins per GO term
go_top_n     <- 15       # top terms to plot

# reusable, offline GO over-representation (hypergeometric) helper -----------------------------------
go_enrichment <- function(foreground, background, annotation,
                          id_col = "accession", go_col = "go_p", min_genes = 2, top_n = 15) {
  fg <- unique(stats::na.omit(unlist(foreground)))
  bg <- unique(stats::na.omit(unlist(c(background, fg))))          # foreground must sit inside background
  if (length(fg) < min_genes || length(bg) < 5) return(NULL)
  am <- as.data.table(as.data.frame(annotation)[, c(id_col, go_col)])
  setnames(am, c(id_col, go_col), c("pid", "goann"))
  am <- unique(am[pid %in% bg & !is.na(goann) & nzchar(goann)])
  if (nrow(am) == 0) return(NULL)
  pt <- am[, .(term = trimws(unlist(strsplit(goann, ";")))), by = pid][nzchar(term)]
  N  <- uniqueN(pt$pid); n <- uniqueN(pt$pid[pt$pid %in% fg])      # annotated background / foreground
  if (n < min_genes) return(NULL)
  ts <- pt[, .(K = uniqueN(pid), k = uniqueN(pid[pid %in% fg])), by = term][k >= min_genes]
  if (nrow(ts) == 0) return(NULL)
  ts[, pval := phyper(k - 1, K, N - K, n, lower.tail = FALSE)]     # P(overlap >= k)
  ts[, padj := p.adjust(pval, method = "BH")]
  ts[order(padj, -k)][seq_len(min(top_n, .N))]
}
go_barplot <- function(go_tbl, title) {
  if (is.null(go_tbl) || nrow(go_tbl) == 0) return(invisible(NULL))
  gt <- copy(go_tbl); gt[, term_short := ifelse(nchar(term) > 60, paste0(substr(term, 1, 57), "..."), term)]
  ggplot(gt, aes(reorder(term_short, -padj), -log10(padj), fill = k)) +
    geom_col() + coord_flip() +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
    labs(title = title, x = NULL, y = "-log10(BH-adjusted p)", fill = "n hits") + theme_bw()
}
# run + plot + save one hit set
.run_go <- function(fg, bg, tag) {
  fg <- unique(stats::na.omit(unlist(fg)))
  if (length(fg) < go_min_genes) { message("[", tag, "] < ", go_min_genes, " proteins - skipping GO."); return(invisible(NULL)) }
  gt <- tryCatch(go_enrichment(fg, bg, uniprot, "accession", go_column, go_min_genes, go_top_n),
                 error = function(e) { message("[", tag, "] GO failed: ", conditionMessage(e)); NULL })
  if (is.null(gt) || nrow(gt) == 0) { message("[", tag, "] no enriched GO terms."); return(invisible(NULL)) }
  fwrite(gt, file.path(output_files_location, paste0("GOenrichment_", go_column, "_", tag, ".txt")), sep = "\t")
  print(go_barplot(gt, paste0("GO ", go_column, " - ", tag, " (", length(fg), " proteins)")))
}

# ---- assembly-change proteins (background = all assembly-tested proteins) ----
if (exists("diffAssemblyState")) {
  .assembly_bg <- unique(as.data.table(diffAssemblyState)$protein_id)
  if (exists("more_assembled_proteins_list"))      .run_go(more_assembled_proteins_list,      .assembly_bg, "moreAssembled")
  if (exists("less_assembled_proteins_list"))      .run_go(less_assembled_proteins_list,      .assembly_bg, "lessAssembled")
  if (exists("proteins_withAssemblyChanges_list")) .run_go(proteins_withAssemblyChanges_list, .assembly_bg, "assemblyChanged")
} else message("diffAssemblyState not found - skipping assembly GO enrichment.")

# ---- significant local-vs-global proteins (background = full differential proteome) ----
if (exists("local_vs_global_sig_list") && length(local_vs_global_sig_list) > 0) {
  .lvg_bg <- unique(as.data.table(protein_DiffExprProtein)$feature_id)
  .lvg_nms <- names(local_vs_global_sig_list)
  for (nm in seq_along(local_vs_global_sig_list)) {
    tag <- paste0("localVsGlobal_", if (!is.null(.lvg_nms)) .lvg_nms[nm] else nm)
    .run_go(local_vs_global_sig_list[[nm]], .lvg_bg, tag)
  }
} else message("local_vs_global_sig_list empty/not found - skipping local-vs-global GO enrichment.")
```


# COMPLEX level analysis

```{r}
# Complex Portal reference table for this organism. The filename is set once in
# the SETUP section via `complex_portal_file` (this branch: S. cerevisiae,
# taxid 559292). Place that file in data/raw/. (Original sep was 't', a stray
# character; corrected to "\t".)
complex_portal <- read.csv(here("data", "raw", complex_portal_file), sep = "\t", header = TRUE) %>%
  janitor::clean_names() 
```

## - set hypothesis

```{r}
# set hypothesis

complexHypotheses <- complex_portal %>%
  dplyr::select(x_complex_ac, recommended_name, aliases_for_complex, description, complex_properties, complex_assembly, expanded_participant_list ) %>% #only keep needed columns 
separate_rows(expanded_participant_list, sep = "\\|") %>% #separate each partecipant in a different row  (sep is a REGEX -> "\\|" = literal pipe; plain "|" splits on every character)
separate(expanded_participant_list, c("protein_id", "stoichiometry"), sep = "\\(", convert = TRUE, remove = FALSE) %>% # separate the stoichiometry in a different column  (sep is a REGEX -> "\\(" = literal "("; plain "(" is an invalid regex and errors)
mutate(stoichiometry = gsub(")", "", stoichiometry, fixed = TRUE)) %>% #remove second bracket from the stoichiometry number  (fixed = TRUE -> literal ")")
 dplyr::rename("complex_id"="x_complex_ac", "complex_name"="recommended_name") # fix col names

complexHypotheses_minimal <- complexHypotheses %>%
   dplyr::select(complex_id, complex_name, protein_id ) %>%
  mutate(complex_id = str_replace(complex_id, "CPX-", "")) %>%
  as.data.table()

head(complexHypotheses_minimal)


binary_hyp <- complexHypotheses_minimal %>%
  dplyr::select(complex_id, protein_id) %>%
  mutate(protein_id = gsub("-PRO_.*", "", protein_id))#remove -PRO_

#### plus decoy

pathLength <- calculatePathlength(binary_hyp)

complexTargetsPlusDecoys <- generateComplexDecoys(target_hypotheses=complexHypotheses_minimal,
                                                dist_info=pathLength,
                                                min_distance = 3,
                                                append=TRUE)


```

## - get Complex features

```{r}
#' ## Summarize protein traces across conditions
protein_traces_sum <- integrateTraceIntensities(protein_traces_list,
                                                design_matrix = NULL,
                                                integrate_within = NULL,
                                                aggr_fun = "sum")
```

```{r}

plotSummarizedMScoverage(hypotheses = complexHypotheses_minimal,
                         protTraces = protein_traces_sum,
                         PDF = F)


#' ## Complex feature finding
# Cached and parallelised, matching the Ecoli report. This was running SERIALLY here (parallelized = F,
# n_cores = 1) and with no cache, so it recomputed in full on every render on a single core.
.cf_cores <- parallel::detectCores()
.cf_cores <- if (is.na(.cf_cores) || .cf_cores < 2L) 1L else min(.cf_cores - 1L, 12L)
.cf_file <- file.path(out_dir, "rdata", "complexFeatures.RData")
if (!exists("load_complexFeatures.RData")) load_complexFeatures.RData <- TRUE
if (load_complexFeatures.RData && file.exists(.cf_file)) {
  load(.cf_file)   # restores `complexFeatures` from the previous run
  message("Loaded cached complexFeatures from ", .cf_file, " (skipped findComplexFeatures).")
  .brk("findComplexFeatures SKIPPED (loaded from cache)")
} else {
  .brk(paste0("findComplexFeatures starting (parallelized=T, n_cores=", .cf_cores, ") ..."))
  complexFeatures <-  findComplexFeatures(traces = protein_traces_sum,
                                          complex_hypothesis = complexTargetsPlusDecoys,
                                          corr_cutoff = 0.9,
                                          window_size = 7,
                                          parallelized = T,
                                          n_cores = .cf_cores,
                                          collapse_method = "apex_network",
                                          perturb_cutoff = "5%",
                                          rt_height = 1,
                                          smoothing_length = 7)
  dir.create(dirname(.cf_file), recursive = TRUE, showWarnings = FALSE)
  save(complexFeatures, file = .cf_file)
  .brk("findComplexFeatures DONE and cached")
}
```

## - filter Complex features

```{r, eval=TRUE, message=FALSE, warning=FALSE, cache=FALSE}

complexFeaturesBest <- getBestFeatures(complexFeatures)
```

```{r}

# NOTE: bare `filterFeatures` is MASKED here by another loaded package's S4 generic (signature
# object, filter - e.g. QFeatures/ProtGenerics pulled in via MSnbase), which has no method for a
# data.table -> "unable to find an inherited method for function 'filterFeatures'". Qualify it to
# CCprofiler's function explicitly. (If you hit the same error on another function, CCprofiler:: it too.)
complexFeaturesBestFilterd <- CCprofiler::filterFeatures(complexFeaturesBest,
                                  complex_ids = NULL,
                                  protein_ids = NULL,
                                  min_feature_completeness = NULL,
                                  min_hypothesis_completeness = NULL,
                                  min_subunits = NULL,
                                  min_peak_corr = NULL,
                                  min_monomer_distance_factor = 2)





complexFeaturesBestFilterdScored <- scoreFeatures(complexFeaturesBestFilterd, FDR=0.1, PDF=F) #,  ## at FDR 0.05 anything passes name="qvalueStats_complexFeatures")

scoredDataAll <- appendSecondaryComplexFeatures(scoredPrimaryFeatures = complexFeaturesBestFilterdScored, allFeatures = complexFeatures, peakCorr_cutoff = 0.7)


plotSummarizedComplexes(scoredDataAll, complexHypotheses_minimal, protein_traces_sum, PDF=F) # ,name="complex_completeness_pie")

summarizeFeatures(scoredDataAll,
                plot=TRUE,
                PDF=F ) #,name="feature_summary_scoredDataAll")

summarizeFeatures(complexFeaturesBestFilterdScored,
                  plot=TRUE,
                  PDF= F) #,  name="feature_summary_complexFeaturesBestFiltered")

```

## - extract Complex features from each pep traces

extract complex features from pep traces pepTracesList_filtered_minN

by condition

```{r}


complexFeaturesUnique <- getUniqueFeatureGroups(
                            feature_table = scoredDataAll,
                            rt_height = 0,
                            distance_cutoff = 1.25)

complexFeaturesCollapsed <- callapseByUniqueFeatureGroups(
                              feature_table = complexFeaturesUnique,
                              rm_decoys = TRUE)

################' ## Extract feature values
.complex_featureVals_file <- file.path(out_dir, "rdata", "complex_featureVals.RData")
if(!file.exists(.complex_featureVals_file)){

# THIS IS THE STEP THAT STALLS THE COMPLEX SECTION (unnamed-chunk-31, ~76-77%). extractFeatureVals is
# single-threaded and its cost scales with the TRACES it is handed, not with the number of features -
# ~88 min per call on this data whatever the feature count. Handing it the whole peptide trace list
# therefore pays that cost in full, and nothing was written until it finished, so an interrupted run
# lost everything. Ported from the protein-level fix already in this file (and from Ecoli_PCM):
# restrict the traces to the proteins that actually appear in the complex features (a hypothesis set
# covers far fewer proteins than the whole proteome), and checkpoint in chunks so it survives an interrupt.
.brk("complex extractFeatureVals -- SERIAL CCprofiler step, the usual stall point (chunk 31) ...")
.cfv_chunk_dir <- file.path(out_dir, "rdata", "complex_featureVals_chunks")
dir.create(.cfv_chunk_dir, recursive = TRUE, showWarnings = FALSE)
if (!exists("complex_featureval_chunks")) complex_featureval_chunks <- 10L

.extract_complex_chunked <- function() {
  FT <- complexFeaturesCollapsed
  idcol <- intersect(c("complex_id", "feature_id", "protein_id"), names(FT))[1]
  if (is.na(idcol)) stop("no id column to split on")
  ids <- unique(as.character(FT[[idcol]]))
  if (length(ids) < 2 * complex_featureval_chunks) stop("too few features to be worth chunking")
  .mk <- file.path(.cfv_chunk_dir, "chunk_count.txt")
  if (file.exists(.mk)) {
    prev <- suppressWarnings(as.integer(readLines(.mk, warn = FALSE)[1]))
    if (is.finite(prev) && prev != complex_featureval_chunks) {
      message("complex_featureval_chunks changed - discarding chunks from the old split.")
      unlink(list.files(.cfv_chunk_dir, pattern = "^chunk_.*\\.rds$", full.names = TRUE))
    }
  }
  writeLines(as.character(complex_featureval_chunks), .mk)
  grp <- split(ids, cut(seq_along(ids), breaks = complex_featureval_chunks, labels = FALSE))
  # the proteins a chunk needs: subunit accessions of the complexes in that chunk
  pcol <- intersect(c("protein_id", "subunits_detected", "subunits_annotated"), names(FT))[1]
  # SAFETY: if we cannot get subunit proteins, subsetting the traces is impossible and each chunk would
  # silently fall back to the FULL trace list - i.e. pay the full multi-hour cost N times, far worse than
  # one plain call. Bail out (the outer tryCatch then runs a single plain call) rather than risk that.
  if (is.na(pcol)) stop("no subunit column (protein_id/subunits_detected/subunits_annotated) to subset traces by")
  .allprot   <- unique(unlist(strsplit(as.character(FT[[pcol]]), "[;,]\\s*")))
  .traceprot <- tryCatch(unique(pepTracesList_filtered_minN[[1]]$trace_annotation$protein_id), error = function(e) NULL)
  # if the complex subunits cover most of the trace proteins, subsetting saves little and chunking would
  # be slower than one plain call - so bail (the outer tryCatch runs the single plain call).
  if (!length(.allprot)) stop("could not parse any subunit proteins from ", pcol, " - cannot subset traces")
  if (!is.null(.traceprot) && length(.traceprot) &&
      length(intersect(.allprot, .traceprot)) > 0.6 * length(.traceprot))
    stop("subunit subsetting would keep >60% of the trace list - a plain call is cheaper than chunking")
  out <- vector("list", length(grp)); t0 <- Sys.time(); first_new <- TRUE
  for (k in seq_along(grp)) {
    cf <- file.path(.cfv_chunk_dir, sprintf("chunk_%03d.rds", k))
    if (file.exists(cf)) { out[[k]] <- readRDS(cf); next }
    sub <- FT[as.character(FT[[idcol]]) %in% grp[[k]], ]
    prot <- if (!is.na(pcol))
      unique(unlist(strsplit(as.character(sub[[pcol]]), "[;,]\\s*"))) else NULL
    st <- if (!is.null(prot) && length(prot))
      tryCatch(subset(pepTracesList_filtered_minN, trace_subset_ids = prot,
                      trace_subset_type = "protein_id"),
               error = function(e) pepTracesList_filtered_minN) else pepTracesList_filtered_minN
    out[[k]] <- extractFeatureVals(traces = st, features = sub, design_matrix = design_matrix,
                                   extract = "subunits", imputeZero = T, verbose = F)
    saveRDS(out[[k]], cf)
    .brk(sprintf("  complex extractFeatureVals chunk %d/%d done (%.1f min elapsed)", k, length(grp),
                 as.numeric(difftime(Sys.time(), t0, units = "mins"))))
    if (first_new) {
      .m <- as.numeric(difftime(Sys.time(), t0, units = "mins")); first_new <- FALSE
      .brk(sprintf("  -> first new chunk %.1f min; %d left => about %.1f h to go",
                   .m, length(grp) - k, .m * (length(grp) - k) / 60))
    }
  }
  out <- out[!vapply(out, is.null, logical(1))]
  if (!length(out)) stop("every chunk came back empty")
  if (length(unique(vapply(out, function(z) paste(names(z), collapse = "|"), character(1)))) != 1L)
    stop("chunks disagree on their columns - not safe to combine")
  res <- data.table::rbindlist(out, use.names = TRUE, fill = FALSE)
  if (inherits(out[[1]], "data.table")) res else as.data.frame(res)
}

complex_featureVals <- tryCatch(.extract_complex_chunked(), error = function(e) {
  message("Chunked complex extraction not used (", conditionMessage(e), "); running one plain call.")
  extractFeatureVals(traces = pepTracesList_filtered_minN,
                     features = complexFeaturesCollapsed,
                     design_matrix = design_matrix,
                     extract = "subunits", imputeZero = T, verbose = F)
})
.brk("complex extractFeatureVals DONE")

save(complex_featureVals, file = .complex_featureVals_file)}

if(file.exists(.complex_featureVals_file)){
  load(.complex_featureVals_file)}

######################################################################################

# Cache + breadcrumb the complex fillFeatureVals too, so the last serial step of the complex tail is
# visible in the log and is not repeated on a re-render after the expensive extraction above.
.complex_filled_file <- file.path(out_dir, "rdata", "complex_featureValsFilled.RData")
if (file.exists(.complex_filled_file)) {
  load(.complex_filled_file)
  .brk("complex fillFeatureVals SKIPPED (loaded from cache)")
} else {
  .brk("complex fillFeatureVals -- SERIAL CCprofiler step ...")
  complex_featureValsFilled <- fillFeatureVals(featureVals = complex_featureVals,
                                               tracesList = pepTracesList_filtered_minN,
                                               design_matrix = design_matrix)
  save(complex_featureValsFilled, file = .complex_filled_file)
  .brk("complex fillFeatureVals DONE and cached")
}
```

## - differential

```{r}
# Reordering the vector


# create list to append to
complex_DiffExprPep_list <- list()

.brk("complex differential test (testDifferentialExpression per condition) ...")
# Sample_id_treat / Condition fix - see the "compare local vs global" chunk above for the rationale.
Sample_id_treat <- treatment_conditions   # condition label(s); compare each treatment vs the reference
for(i in Sample_id_treat) {
            
          sample_vector <- c(i, ref_condition) 
          
          #filter for just 0- and condition x
          complex_featureValsFilled_i <- complex_featureValsFilled %>%
            filter(Condition %in% sample_vector) %>%
            mutate(Condition = factor(Condition, levels = sample_vector)) %>%
            arrange(Condition)


          #' ## Perform peptide-level differential expression testing for all features
          complex_DiffExprPep_i <- testDifferentialExpression_beniFix(featureVals = complex_featureValsFilled_i,
                                                            compare_between = "Condition",
                                                            level = "peptide",
                                                            measuredOnly = FALSE)
          
          # Append to the list
                    complex_DiffExprPep_list[[i]] <- complex_DiffExprPep_i
}
.brk("complex differential test DONE")
```

```{r}
for (i in complex_DiffExprPep_list) {
  
#plots some numbers
hist(i$pVal, breaks = 100)
    hist(i$global_pVal, breaks = 100)
    hist(log(i$qint1), breaks = 100)
    hist(log(i$qint2), breaks = 100)
    hist(log(i$global_int2_imp), breaks = 100)
    hist(log(i$global_int1_imp), breaks = 100)
    hist(i$log2FC, breaks = 50)
    hist(i$global_log2FC, breaks = 50)
}
```

```{r}
complex_DiffExprProtein_list <- list()

for (i in seq_along(complex_DiffExprPep_list)) {
  
  # Get the current dataframe
  df <- complex_DiffExprPep_list[[i]] # need to do this, because otherwise the indexing is wrong when I want to append the results to complex_DiffExprProtein_list
  
#' ## Aggregate differential expression results to the protein level
  complex_DiffExprProtein <- aggregatePeptideTests(df) # need to do this, because otherwise the indexing is wrong when I want to append the results to complex_DiffExprProtein_list


# and plot numbers
hist(complex_DiffExprProtein$pVal, breaks = 100)
    hist(complex_DiffExprProtein$global_pVal, breaks = 100)
    hist(complex_DiffExprProtein$medianLog2FC, breaks = 50)
    hist(complex_DiffExprProtein$global_medianLog2FC, breaks = 50)
    
#append to vector
    # Append to the list with name
  name <- names(complex_DiffExprPep_list)[i]
  complex_DiffExprProtein_list[[name]] <- complex_DiffExprProtein
    }
```

```{r}
complex_DiffExprComplex_list <- list()

for (i in seq_along(complex_DiffExprProtein_list)) {
  
  # Get the current dataframe
  df <- complex_DiffExprProtein_list[[i]] # need to do this, because otherwise the indexing is wrong when I want to append the results to complex_DiffExprComplex_list
  
 #' ## Aggregate differential expression results to the complex level
 complex_DiffExprComplex <- aggregateProteinTests(df)

# plot some numbers
 hist(complex_DiffExprComplex$pVal, breaks = 100)
    hist(complex_DiffExprComplex$global_pVal, breaks = 100)
    hist(complex_DiffExprComplex$medianLog2FC, breaks = 50)
    hist(complex_DiffExprComplex$global_medianLog2FC, breaks = 50)

#append to vector keeping names
  
  name <- names(complex_DiffExprProtein_list)[i]
  complex_DiffExprComplex_list[[name]] <- complex_DiffExprComplex
}
    
```

## - general volcano plots

```{r}
for(i in complex_DiffExprPep_list) {
  plotVolcano(i, PDF = F, pBHadj_cutoff = 0.05)}
```

```{r}
for(i in complex_DiffExprProtein_list) {
  plotVolcano(i, PDF = F, pBHadj_cutoff = 0.05)}
```

```{r}
for(i in complex_DiffExprComplex_list) {
  plotVolcano(i, PDF = F, pBHadj_cutoff = 0.05)}
```

## - Get differential complexes

```{r}

diffComplexes_list <- list()

for (j in seq_along(complex_DiffExprComplex_list)) {
  
  # Extract the current dataframe
  i <- complex_DiffExprComplex_list[[j]]
  
  # Identify unique complexes and differential complexes
  allComplexes <- unique(i$complex_id)
  diffComplexes <- unique(i[i$pBHadj < 0.05 & abs(i$medianLog2FC) > 1, "complex_id"])
  
  #   Append diffComplexes to the list with the name from the original list
  name <- names(complex_DiffExprComplex_list)[j]
  diffComplexes_list[[name]] <- diffComplexes
  
  # Volcano plot highlighting differential complexes
  highlight_complex <- diffComplexes
  plotVolcano(i, highlight = highlight_complex, PDF = FALSE, pBHadj_cutoff = 0.05)
}
```

## - plot diffComplexes

```{r}
  #' # Plot some differential protein features
  #' 
for (i in diffComplexes_list) {
  
    for (id in i) {
        plotFeatures(feature_table = scoredDataAll,
                     traces = protein_traces_list,
                     feature_id = id,
                     annotation_label = "protein_id",  # was "input_id" (peptide-level id, absent from protein traces); protein traces are labelled by protein_id
                     design_matrix=design_matrix,
                     calibration=calibration,
                     peak_area = T,
                     legend = F,
                     onlyBest = F,
                     monomer_MW=T)
    }
  }

```

## - plot all_complexes

```{r}

all_complexes <- unique(complex_featureValsFilled$complex_id)

for (id in all_complexes) {
  plotFeatures(feature_table = scoredDataAll,
               traces = protein_traces_list,
               feature_id = id,
               annotation_label = "protein_id",  # was "input_id" (peptide-level id, absent from protein traces); protein traces are labelled by protein_id
               design_matrix=design_matrix,
               calibration=calibration,
               peak_area = F,
               legend = T,
               onlyBest = F,
               monomer_MW=T,
               aggregateReplicates=F) #turn T if n_replicates >1
}
dev.off()
```

## save RData for plotting

```{r}

# FLAG (bug fix): the original referenced protein_DiffExprPep_list / protein_DiffExprProtein_list,
# which are never created - the ORIGINAL report (ORIGINAL_*.Rmd) has the identical broken save. The
# real objects are protein_DiffExprPep / protein_DiffExprProtein (the "_list" is only a FILENAME
# convention; see the per-object saves above). Corrected the two names so the save works; nothing else.
save(pepTracesList_filtered, pepTracesList_filtered_minN,protein_traces_list,protein_DiffExprPep,protein_DiffExprProtein,  file = paste0(RData_for_further_plotting_and_analysis_location,"/",experiment_name, "_for_plotting.RData"))

```

# 7. Cross-correlation differential analysis (alternative / complement)

The statistical test above compares **intensity / assembly state** between
conditions. This section is a complementary, shape-based view: for each protein it
compares the **elution profile shape and position** between the two conditions
using the **cross-correlation function (CCF)**.

For every protein we cross-correlate its mean profile in the reference condition
(`strain31`) against the treatment (`strain83`) over a range of fraction lags and
record:

* `max_ccf` — the highest cross-correlation (how similar the two profiles are in
  shape, scale-independent);
* `best_lag` — the lag (in fractions) at which that maximum occurs, i.e. the
  **peak shift** between conditions (0 = no shift);
* `zero_lag_cor` — the correlation with no shift (similarity "as is").

A non-zero `best_lag` (a peak that moves to a different fraction / apparent MW) or
a low similarity flags proteins whose elution behaviour changes between strains —
e.g. complex (dis)assembly — which an abundance-only test can miss.

## 7a. Parameters

```{r ccf-params}
ccf_lag_max         <- 5                       # max shift (fractions) to scan
ccf_shift_threshold <- 1                       # |best_lag| >= this  -> flagged "shifted"
ccf_min_intensity   <- 0                       # require this summed intensity in BOTH conditions
ccf_ref_condition   <- ref_condition           # strain31 (reference)
ccf_treat_condition <- treatment_conditions[1] # strain83 (treatment)
```

## 7b. Per-condition mean elution profiles

```{r ccf-condition-traces}
# Average replicate protein traces within each condition. CCF is scale-invariant,
# so aggr_fun = "sum" (used elsewhere in this pipeline) and "mean" are equivalent
# for the correlation/shift; "sum" is used here because it is known to work with
# this CCprofiler build.
cond_traces <- integrateTraceIntensities(
  protein_traces_list,
  design_matrix    = design_matrix,
  integrate_within = "Condition",
  aggr_fun         = "sum"
)

# Robust extractor: traces$traces -> numeric matrix (proteins x fractions).
.get_trace_matrix <- function(traces_obj) {
  dt <- traces_obj$traces
  frac_cols <- grep("^[0-9]+$", colnames(dt), value = TRUE)
  frac_cols <- frac_cols[order(as.numeric(frac_cols))]
  if (data.table::is.data.table(dt)) {
    m <- as.matrix(dt[, ..frac_cols])
  } else {
    m <- as.matrix(dt[, frac_cols, drop = FALSE])
  }
  rownames(m) <- dt$id
  m
}

ref_mat    <- .get_trace_matrix(cond_traces[[ccf_ref_condition]])
treat_mat  <- .get_trace_matrix(cond_traces[[ccf_treat_condition]])
common_ids <- intersect(rownames(ref_mat), rownames(treat_mat))
message("Cross-correlation over ", length(common_ids), " proteins present in both conditions.")
```

## 7c. Compute the cross-correlation metrics per protein

```{r ccf-compute}
compute_ccf_metrics <- function(x, y, lag_max = 5) {
  x[is.na(x)] <- 0
  y[is.na(y)] <- 0
  if (stats::sd(x) == 0 || stats::sd(y) == 0) {
    return(c(max_ccf = NA_real_, best_lag = NA_real_, zero_lag_cor = NA_real_))
  }
  cc       <- stats::ccf(x, y, lag.max = lag_max, plot = FALSE)
  acf_vals <- as.numeric(cc$acf)
  lags     <- as.numeric(cc$lag)
  i_max    <- which.max(acf_vals)
  zero_i   <- which(lags == 0)
  c(max_ccf      = acf_vals[i_max],
    best_lag     = lags[i_max],
    zero_lag_cor = if (length(zero_i)) acf_vals[zero_i] else NA_real_)
}

ccf_results <- data.table::rbindlist(lapply(common_ids, function(id) {
  m <- compute_ccf_metrics(ref_mat[id, ], treat_mat[id, ], lag_max = ccf_lag_max)
  data.table::data.table(
    protein_id   = id,
    max_ccf      = m[["max_ccf"]],
    best_lag     = m[["best_lag"]],
    zero_lag_cor = m[["zero_lag_cor"]],
    int_ref      = sum(ref_mat[id, ],   na.rm = TRUE),
    int_treat    = sum(treat_mat[id, ], na.rm = TRUE)
  )
}))

ccf_results[, shifted := (!is.na(best_lag)) &
              (abs(best_lag) >= ccf_shift_threshold) &
              (int_ref   >= ccf_min_intensity) &
              (int_treat >= ccf_min_intensity)]
ccf_results[, log2_int_ratio := log2((int_treat + 1) / (int_ref + 1))]
data.table::setorder(ccf_results, -shifted, -best_lag)

message("Proteins flagged as shifted (|lag| >= ", ccf_shift_threshold, "): ",
        sum(ccf_results$shifted, na.rm = TRUE))
DT::datatable(ccf_results[order(-abs(best_lag))][1:min(.N, 100)],
              caption = "Per-protein cross-correlation metrics (top by |shift|)")
```

## 7d. Diagnostic plots

```{r ccf-plots, fig.width=7, fig.height=4}
# Distribution of optimal lags (peak shifts)
print(
  ggplot(ccf_results[!is.na(best_lag)], aes(x = best_lag)) +
    geom_bar(fill = "#4C72B0") + theme_classic() +
    labs(title = "Optimal cross-correlation lag (peak shift) per protein",
         x = "Optimal lag (fractions)   [0 = no shift]", y = "proteins")
)

# Profile-shape similarity (max CCF)
print(
  ggplot(ccf_results[!is.na(max_ccf)], aes(x = max_ccf)) +
    geom_histogram(bins = 50, fill = "#55A868") + theme_classic() +
    labs(title = "Maximum cross-correlation (profile-shape similarity)",
         x = "max CCF", y = "proteins")
)

# Shift vs zero-lag similarity, hits highlighted
print(
  ggplot(ccf_results[!is.na(best_lag)],
         aes(x = best_lag, y = zero_lag_cor, color = shifted)) +
    geom_jitter(width = 0.12, height = 0, alpha = 0.6) +
    scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick")) +
    theme_classic() +
    labs(title = "Elution shift vs zero-lag similarity",
         x = "Optimal lag (fractions)", y = "zero-lag correlation", color = "shifted")
)
```

## 7e. Inspect top shifted proteins (overlaid profiles)

```{r ccf-profile-plots, fig.width=7, fig.height=4}
plot_ccf_protein <- function(id) {
  df <- rbind(
    data.frame(fraction = seq_len(ncol(ref_mat)),   intensity = ref_mat[id, ],
               condition = ccf_ref_condition),
    data.frame(fraction = seq_len(ncol(treat_mat)), intensity = treat_mat[id, ],
               condition = ccf_treat_condition)
  )
  mrow <- ccf_results[protein_id == id]
  ggplot(df, aes(x = fraction, y = intensity, color = condition)) +
    geom_line(linewidth = 1) + geom_point(size = 1) + theme_classic() +
    labs(title = paste0(id, "   (lag = ", mrow$best_lag,
                        ", maxCCF = ", round(mrow$max_ccf, 2), ")"),
         x = "Fraction", y = "Mean intensity")
}

top_shifted <- head(ccf_results[shifted == TRUE][order(-abs(best_lag))]$protein_id, 12)
for (id in top_shifted) print(plot_ccf_protein(id))
```

## 7f. Complement to the statistical test

```{r ccf-save-and-compare}
# Save the cross-correlation table
write.table(
  ccf_results,
  file      = file.path(out_dir, "tables", "crosscorrelation_results.txt"),
  sep       = "\t", quote = FALSE, row.names = FALSE
)

# If the statistical protein-level differential results exist, cross-tabulate
# which proteins each approach flags (the "complement" view). Adjust the p-value /
# log2FC column names to match your stats table if they differ.
if (exists("protein_DiffExprProtein_list")) {
  stat_dt <- data.table::rbindlist(protein_DiffExprProtein_list, fill = TRUE)
  pcol <- intersect(c("pBHadj", "qVal", "pVal", "global_pBHadj"), names(stat_dt))[1]
  if (!is.na(pcol) && "protein_id" %in% names(stat_dt)) {
    stat_best <- stat_dt[, .(stat_p = min(get(pcol), na.rm = TRUE)), by = protein_id]
    merged    <- merge(ccf_results, stat_best, by = "protein_id", all = TRUE)
    merged[, stat_sig := !is.na(stat_p) & stat_p < 0.05]
    print(table(shifted = merged$shifted, stat_significant = merged$stat_sig, useNA = "ifany"))
    message("Saved comparison-ready table; 'shifted' (cross-correlation) vs ",
            "'stat_significant' (", pcol, " < 0.05) shown above.")
  } else {
    message("Found protein_DiffExprProtein_list but no recognised p-value/protein_id ",
            "column - join ccf_results to it manually on protein_id.")
  }
} else {
  message("Statistical protein-level results not in memory; run section 6's ",
          "differential test, then re-run this chunk to compare the two approaches.")
}
```
