---
title: "QTL 31 vs 83 - experiments & exploratory analyses"
author: "Ilaria"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

```{r experiments-setup, include=TRUE}
# NOTE: chunks here are NOT evaluated by default. This file is a curated, tidied
# *reference* of exploratory / alternative / other-project analyses kept from the
# original .Rmd. Many chunks were written for a different dataset (conditions
# ctrl / aSYN / PEP / GFP / intox) and/or depend on external scripts that were
# NOT retained. Set eval=TRUE on a chunk (or globally) only after adapting it.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# About this file

This file collects everything that is **not** part of the cleaned QTL
`strain31` vs `strain83` pipeline (which lives in
`DiffAnalysis_Ecoli_PCM.Rmd`). Nothing has been deleted - it has been moved
here, grouped, and annotated, per the handover plan.

**External scripts.** Only 5 helper scripts were retained, in
`R/elution_aware/` (see its README). Source them with, e.g.,
`source(here::here("R", "elution_aware", "CFMS_Elution_Pipeline_P3_Height_PROTEIN_LEVEL_FIXED.R"))`.
Any other `source("./ELUTION_AWARE/...R")` / `source("CFMS_*.R")` call below
refers to a script that was **not retained** (no longer needed) - those chunks
are kept for reference only.

**Datasets / paths.** Several chunks use conditions and absolute Windows paths
from other projects; treat them as templates, not runnable as-is.

---

# A. Alternative normalization (trypsin-ratio)

Moved from the main pipeline's normalization section. This is an alternative to
the normalization used there; the main pipeline does not use it.

### - Calculate the ratios between trypsin peptides for each fraction
```{r}
trypsin_data <- as.data.frame(integrateTraceIntensities(subset(pepTraces_raw, trace_subset_ids="CON_P00761", trace_subset_type="protein_id"))$traces[,1:32])

trypsin_peptide_averages <- rowSums(trypsin_data)/32

trypsin_ratios <- trypsin_data

for(i in 1:nrow(trypsin_ratios)){
  trypsin_ratios[i,] <- trypsin_data[i,]/trypsin_peptide_averages[i]
}

trypsin_ratios[trypsin_ratios==0] <- NA

median_trypsin_ratio <- double(length=ncol(trypsin_ratios))
for(j in 1:ncol(trypsin_ratios)){
  median_trypsin_ratio[j] <- median(trypsin_ratios[sort(trypsin_peptide_averages, index.return=T, decreasing=T)$ix[1:5], j], na.rm=T)
}
pepTraces_trypsinNormalized <- pepTracesImp
for(experiment_id in names(pepTraces_trypsinNormalized)){
  for(i in 1:32){
    i_str <- as.character(i)
    pepTraces_trypsinNormalized[[experiment_id]]$traces[,(i_str):=pepTraces_trypsinNormalized[[experiment_id]]$traces[,i:i]/median_trypsin_ratio[i]]
  }
}
```




---

# B. Elution-aware / protein-level exploratory analyses

Originally section 6 of the `.Rmd`. These explore protein- and peptide-level
quantification, proteoform detection, and pipeline comparisons. Retained helper
scripts are in `R/elution_aware/`; chunks that `source()` other `CFMS_*`/
`ELUTION_AWARE/*` files refer to scripts that were not kept.


# 6. Protein quantification

# 6.-1. ELUTION AWARE ANALYSIS (maxLFQ and peptide level)
```{r}
# Source all modules
source("./ELUTION_AWARE/CFMS_Elution_Pipeline_Protein.R")
source("./ELUTION_AWARE/CFMS_Elution_Pipeline_Peptide.R")
source("./ELUTION_AWARE/CFMS_Elution_Pipeline_Comparison.R")
source("./ELUTION_AWARE/CFMS_Elution_Pipeline_QC.R")

# Pipeline 3: Protein-level
results_p3 <- run_protein_elution_pipeline(
  traces_list = pepTracesList_filtered,
  design_matrix = NULL,
  ref_condition = "ctrl",
  treat_condition = "aSYN",
  output_dir = "./ELUTION_AWARE/pipeline3_results",
  gaussian_params = list(
    max_gaussians = 3, #default5
    min_R_squared = 0.3, # default 0.5
    min_consecutive = 5,
    min_iterations = 5, # increasable to 5 for more precise results (slower though)
    max_iterations = 10, # default 50 (very slow!)
    filter_gaussians_center = TRUE,
    filter_gaussians_height = 0.15,
    filter_gaussians_variance_min = 0.5,
    filter_gaussians_variance_max = 50,
    filter_gaussians_min_dist = 1
  )
)

# Pipeline 4: Peptide-level (both approaches)
results_p4 <- run_peptide_elution_pipeline(
  traces_list = pepTracesList_filtered,
  design_matrix = NULL,
  ref_condition = "ctrl",
  treat_condition = "aSYN",
  output_dir = "./ELUTION_AWARE/pipeline4_results",
  gaussian_params = list(
    max_gaussians = 3,
    min_R_squared = 0.3,
    min_consecutive = 5,
    min_iterations = 5,
    max_iterations = 10,
    filter_gaussians_center = TRUE,
    filter_gaussians_height = 0.15,
    filter_gaussians_variance_min = 0.5,
    filter_gaussians_variance_max = 50,
    filter_gaussians_min_dist = 1
  )
)


# Results in: pipeline4_results/adaptive_aggregation/
#             pipeline4_results/proteoform_aware/

# run QC
qc_p3 <- run_comprehensive_qc(results_p3, pipeline_type = "protein",
                               output_dir = "./ELUTION_AWARE/pipeline3_qc")
qc_p4 <- run_comprehensive_qc(results_p4, pipeline_type = "peptide",
                               output_dir = "./ELUTION_AWARE/pipeline4_qc")

# run Comparison
comparison <- compare_elution_pipelines(
  pipeline3_results = results_p3,
  pipeline4_results = results_p4,
  pval_threshold = 0.1,
  output_dir = "./ELUTION_AWARE/pipeline_comparison"
)

##########################################
source("./ELUTION_AWARE/CFMS_Elution_Pipeline_P3_Height_UPGRADED_CUSTOMIDS_ONLY_NAimpute.R") # 4 main steps so it is closer to the gaussian piepline of Aljaz

results_height <- run_p3_height_pipeline(
  traces_list = pepTracesList_filtered,
  ref_condition = "ctrl",
  treat_condition = "aSYN",
  
  # Proteoform clustering (SAME as P3)
  distance_threshold = 1.0,
  min_peptides = 3,
  min_peptides_maxlfq = 2,
  
  # Peak detection (NEW!)
  peak_params = list(
    min_height_fraction = 0.10,
    min_prominence = 0.15,
    min_distance = 3
  ),
   quant_window = 5,                       # Quantification cap: integration window is adaptive now, but never wider than ±5 fractions.
                                          # Think of this as a safety cap against integrating too broadly.

  # --- Peak overlap / ownership handling (NEW) ---
  peak_ownership = "soft",                # "hard" or "soft"
                                          # hard: midpoint boundaries between peak centers (fast; deterministic; no mixing within boundaries)
                                          # soft: proxy shapes → per-fraction ownership; fractions kept only if ownership ≥ min_peak_fraction
  min_peak_fraction = 0.5,                # Only used when peak_ownership="soft".
                                          # 0.5 = keep fractions where this peak contributes at least half (like your Gaussian logic).
                                          # Raise (0.6–0.8) for stricter de-mixing; lower (0.3–0.4) if peaks overlap heavily.

  # --- Peptide support QC in the peak region (NEW) ---
  min_npep = 5,                           # Minimum number of peptides that show signal (>0) in the final included peak fractions.
                                          # Increase to be stricter; decrease if you have low-coverage proteins.
  min_weighted_corr = 0.2,                # Minimum intensity-weighted peptide correlation within the peak region.
                                          # Helps drop “peaks” driven by inconsistent peptides/noise.
                                          # Typical: 0.2 mild, 0.4 moderate, 0.6 strict.

  # --- Peak matching across conditions (shift-aware) ---
  position_tolerance = 3,                 # When quantifying a reference peak in a sample, search ±3 fractions around the reference center
                                          # to find the local max (shift handling).
                                          # Larger = tolerates shifts but risks matching the wrong nearby peak.
                                          # Smaller = safer matching but can miss shifted peaks.

  detect_new_peaks = TRUE,                # If TRUE, also find peaks present in treatment that are NOT near any ctrl reference peak.
                                          # NEW peaks are detected per treatment replicate, then clustered across treatment replicates only
                                          # (stable IDs), then quantified in all samples.

  min_width_estimate = 2,                 # Width QC: minimum estimated halfwidth (FWHM-ish) for a detected peak.
                                          # 1 = basically no filtering (flag only)
                                          # 2 = mild filtering (drops razor-thin spikes)
                                          # 3 = stricter QC (recommended if you see lots of spurious single-fraction peaks)

  
  n_trace_plots = 20,  # Number of significant proteoforms to plot
  custom_ids = c("P06611_A", "P06611_B","P06611_C", "P37028_A", "P06609_A", "P0A6F5_A", "P0A7E5_A", "P0A8G0_A"), #BtuD, BtuF, BtuC (Entabolon shift) and GroEL (sticky contaminant), and CTP synthase and UvrABC (positives controls for ATP dataset)

  return_peak_support_tables = TRUE,       # NEW: writes/returns peak support summaries (e.g., 2/3 replicates detected).
 
  
  output_dir = "p3_height_results_NAimpute"
)


################PROTEIN LEVEL NO CLUSTERING##############
source("./ELUTION_AWARE/CFMS_Elution_Pipeline_P3_Height_PROTEIN_LEVEL_FIXED.R")

# MaxLFQ (default) - more accurate, normalizes across samples
results_median <- run_protein_level_pipeline(
  traces_list = pepTracesList_filtered_label, # or _light or _heavy, (use filtered with SibCorr&CO)
  ref_condition = "GFP",
  treat_condition = "intox",
  aggregation_method = "median", # or "maxlfq", "median", "sum", "top3", "top5", "top10"
  pvalue_adjustment = "qvalue",
  n_permutations = 0,        # 100-1000 (explöoratory: 100, ~10 min running; publication: 1000, ~1h running. with n=3 does not work!)
  adaptive_window = TRUE,      # ← Enable adaptive (FALSE, fixed "quant_window=5", means +/- 5 fractions around peak center)
  width_multiplier = 1.5,      # ← 1.5 × FWHM (Full Width at Half Maximum)
  custom_ids = c("Q04781", "Q05468", "Q12532", "P25694", "P53044", "P33755", "Q04311", "P37840", "P42212"), # ribosome quality control complex and associated proteins and aSYn and GFP at the end.
  output_dir = "./ELUTION_AWARE/protein_results_median_10h"
)
# Source the plotting script
source("./ELUTION_AWARE/plot_protein_traces.R")
# 1. Custom proteins
plot_protein_traces(results_median, protein_ids = c("P42212","P37840",
#                                                         "P10591", "P15108", "P15992", "P25294", "P29311", "P38903", "P04840", "P07251", "P23287", "P25296", "P31109", "P05030", "P35180", "P07170", "Q04432", "P16474", "P36108", "P01123", "P07560", "Q99260", "Q12697"), # claude AI aSYN interactors
"P31109", "P33328", "P32867", "P39926", "P40357", "P20606", "P15303", "P40482", "	
P01123", "Q04491", "P38968", "P40335", "P38759", "P34110", "P49334", "P07251", "P00830","P22141", "P30656", "P23724", "P30657", "P23639", "P38624", "P23638", "P21242", "P40303", "P40302", "P43588", "P38886", "Q06103", "P38764", "P33297", "P40327", "P40016", "Q04062", "P32565", "P33298", "P33299", "Q12250", "Q01939", "Q08723","P06787", "P53968", "P29311", "P34730", "P01123", "P43588"),  # complex disturbance by aSYN claudeAI                 
 output_dir = "./ELUTION_AWARE/protein_results_median_10h/costum_plots/claudeAI_complexes_disturbed")

# 2. Significant proteins
plot_significant_proteins(results, max_plots = 30)

# 3. High fold-change (|log2FC| > 2)
plot_high_fc_proteins(results, fc_threshold = 2, max_plots = 30)

# 4. All interesting proteins (significant + high FC)
plot_interesting_proteins(
  results,
  pval_threshold = 0.05,
  fc_threshold_sig = 1,
  fc_threshold_high = 2,
  max_plots = 50
)

# 5. Top 20 by p-value
plot_top_proteins(results_top10median, criterion = "pvalue", n = 20, output_dir = "./ELUTION_AWARE/protein_results_top10median_10h/plots_top_by_p-val")



##########proteoform traces plotting###############
source("./ELUTION_AWARE/plot_proteoform_peptide_traces.R")

# Plot the best-clustered proteoforms
# quality_metrics <- plot_best_proteoform_peptides(
#   traces_list = pepTracesList_filtered,
#   pipeline_results = results_height,
#   n_top = 10,                                    # Top 10 best-clustered
#   custom_proteoforms = c("P06611_A", "P0A6F5_A", "P0A6T5_A", "	P0A8X2_A"), # Optional: also plot these
#   output_dir = "peptide_trace_plots",
#   file_format = "both"                           # PDF and PNG
# )

# Option A: Best clustering quality (sibling peptide correlation)
plot_by_correlation(pepTracesList_filtered, results_height, n_top = 10)

# Option B: Most significant p-value
plot_by_pvalue(pepTracesList_filtered, results_height, n_top = 10)

# Option C: Largest fold change
plot_by_foldchange(pepTracesList_filtered, results_height, n_top = 10)

# Option D: Custom selection
plot_custom(pepTracesList_filtered, results, 
            custom_ids = c("P75966", "P33634", "P0AFK4", "P0A9X9", "P09323"),
             output_dir = "./ELUTION_AWARE/peptide_trace_plots_median")


# 4. Explore the quality metrics
head(quality_metrics)
high_quality <- quality_metrics %>% filter(mean_sib_corr > 0.8)

#DIAGNOSTIC for protein aggregation method
source("./ELUTION_AWARE/diagnose_aggregation.R")
diagnose_aggregation_sparsity(
  traces_list = pepTracesList_filtered_label,
  protein_map = results_median$protein_map,
  proteins    = results_median$proteins,
  custom_ids  = c("Q04781", "Q05468", "Q12532", "P25694", "P53044",
                   "P33755", "Q04311", "P37840", "P42212")
)



```



```{r}

#' ## Quantify on protein level
# MODIFIED CODE with maxLFQ:
# Your existing code up to Phase 4...
# pepTracesList_filtered is ready

# Load the hybrid module
source("CFMS_Hybrid_Pipeline_Module.R")
source("CFMS_Pipeline_Integration.R")

# Run the complete integrated analysis
results <- run_integrated_analysis(
  pepTracesList_filtered = pepTracesList_filtered,
  design_matrix = design_matrix,
  ref_condition = "ctrl",
  treat_conditions = c("aSYN"),
  output_dir = "hybrid_analysis"
)

# Access results
maxlfq_data <- results$hybrid$maxlfq
proteoform_info <- results$hybrid$piqed
differential <- results$hybrid$differential_combined
predicted_complexes <- results$complexes$complexes$assignments

# Get significant hits (with proteoform info)
significant <- get_significant_hits(differential, padj_threshold = 0.05)


# Source the module (after running hybrid pipeline)
source("CFMS_Comparison_Module.R")

# Run the comparison
comparison <- compare_maxlfq_vs_proteoform(
  combined_results = results$hybrid$differential_combined,
  piqed_results = results$hybrid$piqed,
  concordance_threshold = 0.5,
  output_dir = "comparison_output",
  save_plots = TRUE
)


# Access results
comparison$summary_stats        # Summary statistics
comparison$hidden_hits          # Proteins significant ONLY in proteoform analysis
comparison$discordant_proteins  # All discordant proteins
comparison$recommendations      # Which method to use for each protein

# Create final "best" results table
final_results <- create_recommended_results(
  results$hybrid$differential_combined,
  comparison
)
      
```
# 6a. visualize best candidates and PIQED first analysis v2

```{r}
source("CFMS_PIQED_First_Pipeline.R")

# Visualize top candidates from your current analysis
top_viz <- visualize_top_candidates(
  combined_results = results$hybrid$differential_combined,
  n_top = 50,
  raw_pval_threshold = 0.05,
  output_dir = "top_candidates_viz"
)

# See summary
top_viz$summary

# See top candidates table
head(top_viz$top_candidates, 20)

########## Run the NEW pipeline with PIQED first, MaxLFQ on consistent peptides only###############
results_v2 <- run_piqed_first_pipeline(
  traces_list = pepTracesList_filtered,
  design_matrix = design_matrix,
  ref_condition = "ctrl",
  treat_conditions = c("aSYN"),
  distance_threshold = 1.0,
  min_peptides_clustering = 3,
  use_ibmt = TRUE,
  output_dir = "piqed_first_results"
)

# Compare results
results_v2$differential_combined

# Old results
old_n_sig <- sum(results$hybrid$differential_combined$pBHadj_aSYN < 0.05, na.rm = TRUE)
old_min_p <- min(results$hybrid$differential_combined$pBHadj_aSYN, na.rm = TRUE)

# New results
new_n_sig <- sum(results_v2$differential_combined$pBHadj_aSYN < 0.05, na.rm = TRUE)
new_min_p <- min(results_v2$differential_combined$pBHadj_aSYN, na.rm = TRUE)

message(paste("OLD: Significant proteins:", old_n_sig, "| Min adj.p:", old_min_p))
message(paste("NEW: Significant proteins:", new_n_sig, "| Min adj.p:", new_min_p))

# 1. Visualize PIQED-first results (same function, new data)
top_viz_v2 <- visualize_top_candidates(
  combined_results = results_v2$differential_combined,
  n_top = 50,
  raw_pval_threshold = 0.05,
  output_dir = "top_candidates_piqed_first"
)

# 2. Compare OLD vs PIQED-first pipelines (NEW FUNCTION!)
comparison <- compare_old_vs_piqed_first(
  old_results = results$hybrid$differential_combined,
  new_results = results_v2$differential_combined,
  output_dir = "pipeline_comparison"
)

# 3. View the comparison summary
comparison$old_summary
comparison$new_summary

# 4. See which proteins improved the most
head(comparison$matched_proteins[order(new_pval)], 20)

###### Check improvement specifically for flagged proteins
# Check this output:
comparison$old_summary
comparison$new_summary

flagged_comparison <- comparison$matched_proteins[is_flagged == TRUE]
cat("Flagged proteins - P-value improved:", sum(flagged_comparison$pval_improved, na.rm=TRUE), 
    "out of", nrow(flagged_comparison), "\n")


####################hybrid pipeline
# Source the new module
source("CFMS_Hybrid_Best_Pipeline.R")

# Run the hybrid best-of-both-worlds pipeline
results_hybrid <- run_hybrid_best_pipeline(
  traces_list = pepTracesList_filtered,
  design_matrix = design_matrix,
  ref_condition = "ctrl",
  treat_conditions = c("aSYN"),
  distance_threshold = 1.0,
  min_peptides_clustering = 3,
  use_ibmt = TRUE,
  output_dir = "hybrid_best_results"
)

# Visualize results
top_viz_hybrid <- visualize_top_candidates(
  combined_results = results_hybrid$differential_combined,
  n_top = 50,
  output_dir = "top_candidates_hybrid"
)

```

# 6b. fix BH p-value adjustment
```{r}
# ============================================================
# QUICK FIX: Recalculate BH correction on MAIN analysis only
# ============================================================

# Get main analysis results only (one per protein)
main_only <- results_hybrid$differential_combined[analysis_type == "Hybrid_main"]

# Recalculate BH correction on main analysis only
main_only[, pBHadj_aSYN := p.adjust(pval_aSYN, method = "BH")]

# Check the improvement
cat("=== CORRECTED STATISTICS ===\n")
cat("Total proteins (main analysis):", nrow(main_only), "\n")
cat("Adj.p < 0.05:", sum(main_only$pBHadj_aSYN < 0.05, na.rm = TRUE), "\n")
cat("Adj.p < 0.10:", sum(main_only$pBHadj_aSYN < 0.10, na.rm = TRUE), "\n")
cat("Raw p < 0.05:", sum(main_only$pval_aSYN < 0.05, na.rm = TRUE), "\n")
cat("Min adj.p:", min(main_only$pBHadj_aSYN, na.rm = TRUE), "\n")

# Check for p=1 problem
cat("\n=== DIAGNOSTIC ===\n")
cat("Proteins with p-value = 1.0:", sum(main_only$pval_aSYN == 1, na.rm = TRUE), "\n")
cat("Proteins with p-value = NA:", sum(is.na(main_only$pval_aSYN)), "\n")

# Visualize corrected distribution
hist(main_only$pBHadj_aSYN, breaks = 50, 
     main = "CORRECTED: Adjusted P-value Distribution",
     xlab = "Adjusted P-value (BH on main analysis only)")
abline(v = 0.05, col = "red", lty = 2)

# How many proteins have p-value = 1?
pval_1_proteins <- main_only[pval_aSYN == 1]
cat("Proteins with p=1:", nrow(pval_1_proteins), "\n")

# What do they look like?
head(pval_1_proteins[, .(protein_id, pval_aSYN, log2FC_aSYN, n_peptides_total, is_flagged)])

# Check if they have 0 variance
# This often happens when all replicates have the same value


# Check RAW p-value distribution (this is more informative)
hist(main_only$pval_aSYN, breaks = 50, 
     main = "RAW P-value Distribution", 
     xlab = "Raw P-value")
abline(v = 0.05, col = "red", lty = 2)

# Count adjusted p-values = 1
cat("Adjusted p = 1.0:", sum(main_only$pBHadj_aSYN >= 0.99, na.rm = TRUE), "\n")
cat("Adjusted p < 1.0:", sum(main_only$pBHadj_aSYN < 0.99, na.rm = TRUE), "\n")

```

# 6c. peptide level analysis ALL peptides
```{r}
# Source the module
source("CFMS_Peptide_Level_Analysis.R")

# Run peptide-level analysis (like PIQED)
results_peptide <- run_peptide_level_analysis(
  traces_list = pepTracesList_filtered,
  design_matrix = design_matrix,
  ref_condition = "ctrl",
  treat_conditions = c("aSYN"),
  min_valid_samples = 4,
  use_ibmt = TRUE,
  output_dir = "peptide_level_results"
)

# Compare all FOUR pipelines
comparison_all <- compare_all_four_pipelines(
  old_results = results$hybrid$differential_combined,
  piqed_first_results = results_v2$differential_combined,
  hybrid_results = results_hybrid$differential_combined,
  peptide_results = results_peptide,
  output_dir = "four_pipeline_comparison"
)
```


# 6d. RAW p-value distribution comparisons
```{r}
# ============================================================
# RAW P-VALUE DISTRIBUTION COMPARISON: ALL THREE PIPELINES
# ============================================================

library(ggplot2)
library(data.table)

# --- Extract MAIN analysis results from each pipeline ---

# OLD pipeline (MaxLFQ-first)
old_main <- results$hybrid$differential_combined[analysis_type == "MaxLFQ_main"]
old_pvals <- data.table(
  pval = old_main$pval_aSYN,
  pipeline = "1. OLD (MaxLFQ-first)"
)

# PIQED-first pipeline
# Check what analysis_type is called in your results_v2
piqed_main <- results_v2$differential_combined[grepl("MaxLFQ|main", analysis_type, ignore.case = TRUE)]
# If that doesn't work, try:
# piqed_main <- results_v2$differential_combined[grepl("_A$", proteoform_id)]
piqed_pvals <- data.table(
  pval = piqed_main$pval_aSYN,
  pipeline = "2. PIQED-first"
)

# HYBRID pipeline
hybrid_main <- results_hybrid$differential_combined[analysis_type == "Hybrid_main"]
hybrid_pvals <- data.table(
  pval = hybrid_main$pval_aSYN,
  pipeline = "3. HYBRID"
)

# Combine all
all_pvals <- rbind(old_pvals, piqed_pvals, hybrid_pvals)

# --- Print Summary Statistics ---
cat("\n============================================================\n")
cat("         RAW P-VALUE COMPARISON: ALL THREE PIPELINES        \n")
cat("============================================================\n\n")

summary_stats <- all_pvals[, .(
  n_proteins = .N,
  n_raw_p01 = sum(pval < 0.01, na.rm = TRUE),
  n_raw_p05 = sum(pval < 0.05, na.rm = TRUE),
  min_pval = min(pval, na.rm = TRUE),
  median_pval = median(pval, na.rm = TRUE),
  mean_pval = mean(pval, na.rm = TRUE)
), by = pipeline]

print(summary_stats)

# --- Plot 1: Overlaid Histograms ---
p1 <- ggplot(all_pvals, aes(x = pval, fill = pipeline)) +
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(values = c("1. OLD (MaxLFQ-first)" = "steelblue", 
                                "2. PIQED-first" = "coral",
                                "3. HYBRID" = "forestgreen")) +
  labs(
    title = "Raw P-value Distribution Comparison",
    subtitle = "Overlaid histograms | Red line = p = 0.05",
    x = "Raw P-value",
    y = "Count",
    fill = "Pipeline"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p1)

# --- Plot 2: Faceted Histograms (easier to see) ---
p2 <- ggplot(all_pvals, aes(x = pval, fill = pipeline)) +
  geom_histogram(bins = 50, alpha = 0.7, color = "white") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(values = c("1. OLD (MaxLFQ-first)" = "steelblue", 
                                "2. PIQED-first" = "coral",
                                "3. HYBRID" = "forestgreen")) +
  facet_wrap(~pipeline, ncol = 1, scales = "free_y") +
  labs(
    title = "Raw P-value Distribution by Pipeline",
    subtitle = "Red line = p = 0.05 | Enrichment near 0 indicates true effects",
    x = "Raw P-value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(p2)

# --- Plot 3: Zoom into low p-values (where the action is) ---
p3 <- ggplot(all_pvals[pval < 0.2], aes(x = pval, fill = pipeline)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = 0.01, linetype = "dotted", color = "darkred", linewidth = 1) +
  scale_fill_manual(values = c("1. OLD (MaxLFQ-first)" = "steelblue", 
                                "2. PIQED-first" = "coral",
                                "3. HYBRID" = "forestgreen")) +
  labs(
    title = "Raw P-value Distribution (Zoomed: p < 0.2)",
    subtitle = "Red dashed = 0.05 | Dotted = 0.01",
    x = "Raw P-value",
    y = "Count",
    fill = "Pipeline"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p3)

# --- Plot 4: Bar chart comparing significant counts ---
sig_counts <- all_pvals[, .(
  `p < 0.001` = sum(pval < 0.001, na.rm = TRUE),
  `p < 0.01` = sum(pval < 0.01, na.rm = TRUE),
  `p < 0.05` = sum(pval < 0.05, na.rm = TRUE),
  `p < 0.1` = sum(pval < 0.1, na.rm = TRUE)
), by = pipeline]

sig_counts_long <- melt(sig_counts, id.vars = "pipeline", 
                         variable.name = "threshold", value.name = "count")

p4 <- ggplot(sig_counts_long, aes(x = threshold, y = count, fill = pipeline)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = count), position = position_dodge(width = 0.7), 
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("1. OLD (MaxLFQ-first)" = "steelblue", 
                                "2. PIQED-first" = "coral",
                                "3. HYBRID" = "forestgreen")) +
  labs(
    title = "Number of Proteins by Raw P-value Threshold",
    x = "P-value Threshold",
    y = "Number of Proteins",
    fill = "Pipeline"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0, max(sig_counts_long$count) * 1.15))

print(p4)

# --- Save plots ---
ggsave("pval_comparison_overlay.pdf", p1, width = 10, height = 6)
ggsave("pval_comparison_facet.pdf", p2, width = 10, height = 10)
ggsave("pval_comparison_zoomed.pdf", p3, width = 10, height = 6)
ggsave("pval_comparison_bars.pdf", p4, width = 10, height = 6)

cat("\nPlots saved!\n")
```

# 6e. check peptide level behaviour - opposite directions ect.
```{r}
# Check what columns exist in protein_summary
names(results_peptide$protein_summary)

# Get peptide-level results directly and calculate directions
peptide_results <- results_peptide$peptide_results

# Calculate direction consistency per protein
direction_summary <- peptide_results[, .(
  n_peptides = .N,
  n_positive_fc = sum(log2FC_aSYN > 0, na.rm = TRUE),
  n_negative_fc = sum(log2FC_aSYN < 0, na.rm = TRUE),
  n_sig_p05 = sum(pval_aSYN < 0.05, na.rm = TRUE),
  mean_fc = mean(log2FC_aSYN, na.rm = TRUE),
  fc_sd = sd(log2FC_aSYN, na.rm = TRUE),
  fc_range = max(log2FC_aSYN, na.rm = TRUE) - min(log2FC_aSYN, na.rm = TRUE),
  min_pval = min(pval_aSYN, na.rm = TRUE)
), by = protein_id]

# Calculate percentages and flags
direction_summary[, pct_positive := 100 * n_positive_fc / n_peptides]
direction_summary[, pct_negative := 100 * n_negative_fc / n_peptides]

# Flag proteins with mixed directions (both up and down peptides, at least 3 peptides)
direction_summary[, is_mixed := n_positive_fc > 0 & n_negative_fc > 0 & n_peptides >= 3]

# Flag highly inconsistent (large FC range)
direction_summary[, is_highly_inconsistent := fc_range > 1.5 & n_peptides >= 3]

# Summary
cat("\n=== PEPTIDE DIRECTION CONSISTENCY ===\n")
cat("Total proteins:", nrow(direction_summary), "\n")
cat("Proteins with MIXED peptide directions:", sum(direction_summary$is_mixed), "\n")
cat("Proteins with CONSISTENT directions:", sum(!direction_summary$is_mixed), "\n")
cat("Proteins highly inconsistent (FC range > 1.5):", sum(direction_summary$is_highly_inconsistent, na.rm = TRUE), "\n")

# Among mixed-direction proteins, how many have significant peptides?
mixed_proteins <- direction_summary[is_mixed == TRUE]
cat("\nAmong MIXED proteins:\n")
cat("  With at least 1 sig peptide (p<0.05):", sum(mixed_proteins$n_sig_p05 > 0), "\n")
cat("  With at least 2 sig peptides:", sum(mixed_proteins$n_sig_p05 >= 2), "\n")

# Look at top mixed-direction proteins with significant peptides
cat("\n=== TOP MIXED-DIRECTION PROTEINS WITH SIGNIFICANT PEPTIDES ===\n")
mixed_with_sig <- direction_summary[is_mixed == TRUE & n_sig_p05 > 0]
setorder(mixed_with_sig, -n_sig_p05, min_pval)
print(head(mixed_with_sig[, .(protein_id, n_peptides, n_positive_fc, n_negative_fc, 
                               n_sig_p05, mean_fc, fc_range, min_pval)], 20))

# Visualize
library(ggplot2)

# Plot: Distribution of direction consistency
p1 <- ggplot(direction_summary, aes(x = pct_positive)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(25, 75), linetype = "dashed", color = "red") +
  labs(
    title = "Peptide Direction Consistency per Protein",
    subtitle = "0% = all peptides DOWN | 100% = all peptides UP | 50% = mixed",
    x = "% Peptides with Positive FC",
    y = "Number of Proteins"
  ) +
  theme_minimal()

print(p1)

# Plot: FC range vs significance
p2 <- ggplot(direction_summary[n_peptides >= 3], aes(x = fc_range, y = -log10(min_pval))) +
  geom_point(aes(color = is_mixed), alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  scale_color_manual(values = c("FALSE" = "forestgreen", "TRUE" = "coral"),
                     labels = c("Consistent", "Mixed")) +
  labs(
    title = "Peptide FC Range vs Best P-value",
    subtitle = "Mixed-direction proteins may have opposing proteoform effects",
    x = "FC Range (max - min log2FC)",
    y = "-log10(min p-value)",
    color = "Direction"
  ) +
  theme_minimal()

print(p2)
```

# 6f. peptide.-level wioth proper prptein-aggregation (simes)
```{r}
# Re-source the updated module
source("CFMS_Peptide_Level_Analysis_2.R")

# 1. Aggregate peptide p-values to protein level
protein_from_peptides <- aggregate_peptides_to_protein(
  results_peptide, 
  method = "simes"  # Best for correlated peptides
)

# View top proteins
head(protein_from_peptides[, .(protein_id, n_peptides, pval_aggregated, 
                                pBHadj_aggregated, mean_fc, n_sig_peptides_p05)])

# 2. Compare ALL methods
method_comparison <- compare_aggregation_methods(results_peptide)

# 3. Full comparison including peptide→protein aggregation
comparison_all <- compare_all_four_pipelines(
  old_results = results$hybrid$differential_combined,
  piqed_first_results = results_v2$differential_combined,
  hybrid_results = results_hybrid$differential_combined,
  peptide_results = results_peptide,
  include_peptide_aggregated = TRUE,
  aggregation_method = "simes",
  output_dir = "five_pipeline_comparison"
)
```

# 6g.final proteoform aware peptide-level analysis

 What It Does (5 Steps)

| Step | What | Output |
|------|------|--------|
| **1** | Peptide-level differential analysis | `01_peptide_level_results.csv` |
| **2** | Simes aggregation to protein level | `02_protein_simes_aggregated.csv` |
| **3** | Identify bidirectional proteins | `03_bidirectional_proteins.csv` |
| **4** | Create final summary tables | `05_TOP_DISCOVERIES.csv`, `06_PROTEOFORM_CANDIDATES.csv` |
| **5** | Generate visualizations | `ALL_PLOTS.pdf` |


 Key Output Files
proteoform_aware_results/
├── 01_peptide_level_results.csv      ← All peptide p-values & FCs
├── 02_protein_simes_aggregated.csv   ← Protein p-values (Simes method)
├── 03_bidirectional_proteins.csv     ← Proteins with UP + DOWN peptides
├── 04_protein_direction_summary.csv  ← Direction stats per protein
├── 05_TOP_DISCOVERIES.csv            ← ⭐ Best hits by Simes p-value
├── 06_PROTEOFORM_CANDIDATES.csv      ← ⭐ Bidirectional = proteoform candidates!
├── ALL_PLOTS.pdf                     ← All visualizations
└── plot_*.pdf                        ← Individual plots

```{r}
# Source the FIXED module
source("CFMS_Proteoform_Aware_Final_v2.R")
source("CFMS_Top20_Peptide_Plots.R")

# Run with parallelization
results_final <- run_proteoform_aware_analysis(
  traces_list = pepTracesList_filtered,
  design_matrix = design_matrix,
  ref_condition = "ctrl",
  treat_conditions = c("aSYN"),
  min_valid_samples = 4,
  fc_threshold = 0.5,
  pval_threshold = 0.05,
  #n_cores = 4,  # Adjust based on your CPU (or leave NULL for auto-detect)
  output_dir = "proteoform_aware_results")

# Save PDF plots for top 20 candidates
save_top20_proteoform_plots(results_final, output_dir = "top20_peptide_plots")

# View top proteoform candidates
head(results_final$final_summary$proteoform_candidates, 20)

# View top discoveries
head(results_final$final_summary$top_discoveries, 20)

# See all peptides for a protein of interest
visualize_protein_peptides(results_final$peptide_results, "P0AFL6")

# # Your top proteins - look up each on UniProt
# top_proteins <- c("P0AFL6", "P0A9Q7", "P13029", "P22259", "P0AFI2",
#                   "P0A9M0", "P0ADG7", "P62768", "P52647", "P0A698",
#                   "P15038", "P0A786", "P0A8M6", "P05793", "P15977",
#                   "P33570", "P0A6K6", "P68767", "P0A6F3", "P00370")
# 
# # Quick lookup URL generator
# for (p in top_proteins) {
#   cat(paste0("https://www.uniprot.org/uniprotkb/", p, "/entry\n"))
# }


```


 What the Bidirectional Score Means

Bidirectional Score = (min of n_up_sig, n_down_sig) × 2 
                    + FC_range (capped at 3)
                    + -log10(best_pval)

Higher score = More confident proteoform-specific regulation
Example interpretation:
Score > 5: Strong evidence for distinct proteoforms responding differently
Score 3-5: Moderate evidence
Score < 3: Weak evidence


# 6h. aggregation analysis simes vs stouffer (peptide independencies)
```{r}
source("CFMS_Aggregation_Method_Analysis_v2.R")

agg_results <- run_aggregation_analysis(
  peptide_results = results_final$peptide_results,
  expression_matrix = results_final$expression_matrix,  # optional but recommended
  traces_list = pepTracesList_filtered,                 # for sibling correlation
  output_dir = "aggregation_analysis"
)

# See recommendation
agg_results$recommendation

# Get protein p-values with best method
best_pvals <- agg_results$method_comparison$merged_results
```

# 6i. Consistency analysis btw replicates vs within proteins
```{r}
source("CFMS_Replicate_Consistency_Analysis.R")

rep_results <- run_replicate_consistency_analysis(
  traces_list = pepTracesList_filtered,
  design_matrix = design_matrix,
  ref_condition = "ctrl",
  treat_condition = "aSYN",
  output_dir = "replicate_consistency_analysis"
)

# Key summary metrics
rep_results$summary

# Proteins with reproducible bidirectional behavior
repro_bidir <- rep_results$bidirectional[is_reproducibly_bidirectional == TRUE]
```


# 7. check with Ludovic code for peptide traces along chromatogram
```{r}
source("CFMS_Peptide_Chromatogram_Plots.R")

# Plot control samples with mean ± SD error bars
plots <- generate_chromatogram_plots(
  data = data,                        # Your data already in memory!
  condition_filter = "aSYN",          # Only control samples
  top_n_proteins = 20,                # Top 20 proteins
  top_n_peptides = 20,                # Top 20 peptides per protein
  plot_style = "mean_sd",             # Options: "mean_sd", "replicates", "combined"
  output_dir = "chromatogram_plots_aSYN",
  n_fractions = 32
)


source("CFMS_Peptide_Chromatogram_Plots_treat_vs_ctrl.R")

# ctrl vs aSYN side-by-side, top 100 proteins
plots_2 <- generate_comparison_plots(
  data = data,
  condition1 = "ctrl",
  condition2 = "aSYN",
  top_n_proteins = 100,
  top_n_peptides = 20,
  output_dir = "ctrl_vs_aSYN_plots",
  n_fractions = 32
)

my_proteins <- c("P06612", "P00579", "P22106", "P62399", "P09373", 
                 "P0A6C8", "P0A8V2", "P00864", "P0A8T7", "P21170", 
                 "P42641", "P06959", "P68767", "P23003", "P09831", 
                 "P0A6Z3", "P03018", "P0ABH9", "P0AFP6", "P25665")

plots_3 <- generate_comparison_plots(
  data = data,
  condition1 = "ctrl",
  condition2 = "aSYN",
  priority_proteins = my_proteins,
  top_n_proteins = 200,
  show_protein_sum = FALSE,   # Only colored peptide lines
  output_dir = "ctrl_vs_aSYN_peptides_only"
)

source("CFMS_Peptide_Chromatogram_Plots_zoom.R")

plots_4 <- generate_comparison_plots(
  data = data,
  condition1 = "ctrl",
  condition2 = "aSYN",
  priority_proteins = my_proteins,
  top_n_proteins = 100,
  auto_zoom = TRUE,                   # Default: ON
  intensity_threshold = 5000,         # Default: 5000
  show_protein_sum = FALSE,
  output_dir = "ctrl_vs_PEP_zoomed"
)

```




---

# C. PrInCE, Gaussian/window fitting & one-off tasks

Originally the later sections of the `.Rmd`: PrInCE comparisons, Gaussian/window
fitting, chromatogram-correlation explorations (relevant background for the
planned phase-2 cross-correlation work), and a few one-off data-processing tasks
for other projects (absolute Windows paths - templates only).


# Plot hits from ATP
To compare the quality / noisiness of the chromatograms
```{r}
load("../../../Analysis/PrInCE/Prince_ATP_hits.RData")
prince_hits_ATP <- prince_hits
rm(prince_hits)
for(i in c("P75949")){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  try(plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = i, aggregateReplicates=F))
}
```

# PrInCE analysis
##Prepare data
```{r}

ctrl_treat_traces <- integrateTraceIntensities(protein_traces_list, design_matrix = design_matrix, integrate_within = "Condition", aggr_fun = "sum")
integrated_traces <- integrateTraceIntensities(protein_traces_list, design_matrix = design_matrix, aggr_fun = "sum")

PCM18_treat <- as.matrix(ctrl_treat_traces$PEP$traces[,1:32])
rownames(PCM18_treat) <- ctrl_treat_traces$PEP$traces$id

PCM18_ctrl <- as.matrix(ctrl_treat_traces$ctrl$traces[,1:32])
rownames(PCM18_ctrl) <- ctrl_treat_traces$ctrl$traces$id

PCM18_integrated <- as.matrix(integrated_traces$traces[,1:32])
rownames(PCM18_integrated) <- integrated_traces$traces$id

PCM18_treat[PCM18_treat == 0.0] <- NA
PCM18_ctrl[PCM18_ctrl == 0.0] <- NA
PCM18_integrated[PCM18_integrated == 0.0] <- NA
```

##Functions

```{r bestOfGuessAndRandom}
bestOfGuessAndRandom <- function(gaussians1, gaussians2){
  guess_or_random <- data.frame(protein_id=character(length=length(intersect(names(gaussians1), names(gaussians2)))), better=character(length=length(intersect(names(gaussians1), names(gaussians2)))), random_R2 = double(length=length(intersect(names(gaussians1), names(gaussians2)))), guess_R2 = double(length=length(intersect(names(gaussians1), names(gaussians2)))))
i<-1
for(id in intersect(names(gaussians1), names(gaussians2))){
  guess_or_random[i,1] <- id
  guess_or_random[i,3] <- gaussians1[[id]]$R2
  guess_or_random[i,4] <- gaussians2[[id]]$R2
  if(gaussians1[[id]]$R2 >= gaussians2[[id]]$R2){
    guess_or_random[i,2] <- "random"
  } else {
    guess_or_random[i,2] <- "guess"
  }
  i <- i+1
}

guess_or_random <- guess_or_random %>% mutate(diff=random_R2 - guess_R2)

bestOfRandomAndGuessGaussians <- c(gaussians1[union(filter(guess_or_random, grepl("random", better))$protein_id, setdiff(names(gaussians1),names(gaussians2)))], gaussians2[union(filter(guess_or_random, grepl("guess", better))$protein_id, setdiff(names(gaussians2),names(gaussians1)))])
bestOfRandomAndGuessGaussians
}
```

```{r Join areas function}
#Give a list of replicate areas to join in the order from 1 to n (3 in our case), minimum number of peptides present in the area of the peak to keep it, minimum peptide intensity-weighted average correlation in the area of the peak to keep it and accessions of proteins to remove.
joinAreasAndFoldChanges <- function(replicateAreas=list(), min_npep=5, min_weighted_corr=0.2, exclude_proteins=c()){
  joined_areas <- replicateAreas[[1]] %>% select(protein_id, mu, low_lim, high_lim, excluded_frac, n_frac, n_pep, weighted_corr) %>% filter(n_pep >= min_npep & weighted_corr >= min_weighted_corr)
  i <- 1
  for(replicate in replicateAreas[1:(length(replicateAreas)-1)]){
    replicate_selected_columns <- replicate %>% filter(n_pep >= min_npep & weighted_corr >= min_weighted_corr) %>% select(protein_id, mu, low_lim, high_lim, excluded_frac, n_frac, n_pep, weighted_corr, foldChange, log2area_ctrl, log2area_treat, percent_areactrl, percent_areatreat)
    joined_areas <- joined_areas %>% full_join(replicate_selected_columns, by=join_by(protein_id, mu, low_lim, high_lim, excluded_frac, n_frac, n_pep, weighted_corr), suffix = c("", paste0("_rep", i)))
    i <- i+1
  }
  replicate_selected_columns <- replicateAreas[[length(replicateAreas)]] %>% filter(n_pep >= min_npep & weighted_corr >= min_weighted_corr) %>% select(protein_id, mu, low_lim, high_lim, excluded_frac, foldChange, log2area_ctrl, log2area_treat, percent_areactrl, percent_areatreat)
    joined_areas <- joined_areas %>% full_join(replicate_selected_columns, by=join_by(protein_id, mu, low_lim, high_lim, excluded_frac), suffix = c("_rep1", paste0("_rep", i)))
    
    return(filter(joined_areas, ! protein_id %in% exclude_proteins))
}


```

##Fit gaussians to whole biolRep3 chromatograms
```{r}
#Fit gaussians to biolRep3 chromatograms
ctrl_gaussians <- bestOfGuessAndRandom(build_gaussians_corr_mod(PCM18_ctrl, min_consecutive =3, smooth=F, min_R_squared = 0.9, method = "guess", filter_gaussians_height = 0, filter_gaussians_variance_max = 3, min_iterations = 100, filter_gaussians_min_dist = 1, criterion="AICc", random_seed = 12345), build_gaussians_corr_mod(PCM18_ctrl, min_consecutive =3, smooth=F, min_R_squared = 0.9, method = "random", filter_gaussians_height = 0, filter_gaussians_variance_max = 3, min_iterations = 100, filter_gaussians_min_dist = 1, criterion="AICc", random_seed = 12345))

treat_gaussians <- bestOfGuessAndRandom(build_gaussians_corr_mod(PCM18_treat, min_consecutive =3, smooth=F, min_R_squared = 0.9, method = "guess", filter_gaussians_height = 0, filter_gaussians_variance_max = 3, min_iterations = 100, filter_gaussians_min_dist = 1, criterion="AICc", random_seed = 12345), build_gaussians_corr_mod(PCM18_treat, min_consecutive =3, smooth=F, min_R_squared = 0.9, method = "random", filter_gaussians_height = 0, filter_gaussians_variance_max = 3, min_iterations = 100, filter_gaussians_min_dist = 1, criterion="AICc", random_seed = 12345))

integrated_gaussians <- bestOfGuessAndRandom(build_gaussians_corr_mod(PCM18_integrated, min_consecutive =3, smooth=F, min_R_squared = 0.9, method = "guess", filter_gaussians_height = 0, filter_gaussians_variance_max = 3, min_iterations = 100, filter_gaussians_min_dist = 1, criterion="AICc", random_seed = 12345), build_gaussians_corr_mod(PCM18_integrated, min_consecutive =3, smooth=F, min_R_squared = 0.9, method = "random", filter_gaussians_height = 0, filter_gaussians_variance_max = 3, min_iterations = 100, filter_gaussians_min_dist = 1, criterion="AICc", random_seed = 12345))

save.image()
```


```{r window fitting}
scanFitGaussians <- function(chromatogram_matrix, window_length, increment, min_R2=0.9, max_variance=4) {
  list_of_gaussians=list()
  i=0
  while(ncol(chromatogram_matrix) > window_length + i*increment){
    window_gaussians <- bestOfGuessAndRandom(build_gaussians(chromatogram_matrix[,(1+i*increment):(window_length+i*increment)] ,min_consecutive =3, smooth=F, min_R_squared = min_R2, method = "random", filter_gaussians_height = 0.075, filter_gaussians_variance_max = max_variance), build_gaussians(chromatogram_matrix[,(1+i*increment):(window_length+i*increment)], min_consecutive =3, smooth=F, min_R_squared = min_R2, method = "guess", filter_gaussians_height = 0.075, filter_gaussians_variance_max = max_variance))
    i <- i+1
    list_of_gaussians <- list_of_gaussians %>% list.append(window_gaussians)
  }
  list_of_gaussians
}


#TODO perhaps disregard gaussians with peaks too close to the edge
joinGaussianWindows <- function(gaussianwindows, increment, wholeChromatogramGaussians=list()){
  window<-0
  i <- 1

  all_gaussians <- data.frame(protein_id=character(length=1000), A=double(length=1000), mu=double(length=1000), sigma=double(length=1000))
  
  j<-0
  for(window_iteration in gaussianwindows){
    l<-1
    protein_names <- names(window_iteration)
    for(protein_fit in window_iteration){
      for(k in 1:length(protein_fit$coefs$A)){
        if(i>nrow(all_gaussians)){
          all_gaussians <- all_gaussians %>% rbind(data.frame(protein_id=character(length=1000), A=double(length=1000), mu=double(length=1000), sigma=double(length=1000)))
        }
        all_gaussians[i,1]<-protein_names[l]
        all_gaussians[i,2]<-protein_fit$coefs$A[k]
        all_gaussians[i,3]<-protein_fit$coefs$mu[k]+j*increment
        all_gaussians[i,4]<-protein_fit$coefs$sigma[k]
        i<-i+1
      }
      l<-l+1
    }
    j<-j+1
  }
  print(length(wholeChromatogramGaussians))
  if(length(wholeChromatogramGaussians)>0){
    for(wholeChromatogramGaussians_entry in wholeChromatogramGaussians){
      print(length(wholeChromatogramGaussians_entry))
      protein_names <- names(wholeChromatogramGaussians_entry)
      l<-1
      for(protein_fit in wholeChromatogramGaussians_entry){
        for(k in 1:length(protein_fit$coefs$A)){
          if(i>nrow(all_gaussians)){
            all_gaussians <- all_gaussians %>% rbind(data.frame(protein_id=character(length=1000), A=double(length=1000), mu=double(length=1000), sigma=double(length=1000)))
          }
          all_gaussians[i,1]<-protein_names[l]
          all_gaussians[i,2]<-protein_fit$coefs$A[k]
          all_gaussians[i,3]<-protein_fit$coefs$mu[k]
          all_gaussians[i,4]<-protein_fit$coefs$sigma[k]
          i<-i+1
        }
        l<-l+1
      }
    }
  }
  all_gaussians <- all_gaussians[order(all_gaussians$protein_id),]
  return(all_gaussians[!all_gaussians$mu==0,])
}

gaussianScan <- scanFitGaussians(PCM18_integrated, 12, 4)
joinedGaussians <- joinGaussianWindows(gaussianScan, 4, wholeChromatogramGaussians = list(integrated_gaussians, ctrl_gaussians, treat_gaussians))

joinedGaussians_noWindowFitting <- joinGaussianWindows(list(), 4, wholeChromatogramGaussians = list(integrated_gaussians, ctrl_gaussians, treat_gaussians))

joinedGaussians_OnlyIntegrated <- joinGaussianWindows(list(), 4, wholeChromatogramGaussians = list(integrated_gaussians))

```

```{r}
createConsensusFromWindowFitting <- function(cofracData, joinedgaussians, min_gaussian_dist = 0){
  best_fits <- data.frame(protein_name=character(), A=double(), mu=double(), sigma=double(), model_fold = double(), model_corr=double())
  #Loop over all proteins
  pb_tracker <- 0
  pb <- txtProgressBar(min=0, max=uniqueN(joinedgaussians$protein_id), initial=0)
  for(id in unique(joinedgaussians$protein_id)){
    pb_tracker <- pb_tracker + 1
    setTxtProgressBar(pb, pb_tracker)
    fraction_numbers <- 1:ncol(cofracData) #x-axis values for model calculation
    chromatogram <- cofracData[id,] #experimentally determined data
    chromatogram[is.na(chromatogram)] <- 0 #Change all NAs in the data to zeroes for calculation 
    gaussians <- joinedgaussians %>% filter(grepl(id, protein_id)) #take the gaussians for the current protein id
    #Order them by mu
    gaussians <- gaussians[order(gaussians$mu),]
    #To store the current best candidate
    best_combination <- NA
    best_corr <- -1
    best_square_deviations <- Inf #Metric actually used for selection
    best_model <- NA
    
    #This loops over all possible combinations of the gaussians
    for (i in 1:nrow(gaussians)){
      combination <- utils::combn(1:nrow(gaussians),i)
      
      #This loop selects an individual combination, calculates a model and compares it to the experimental data
      for(j in 1:ncol(combination)){
        model <- double(length=ncol(cofracData))
        
        #If the differences between gaussian peak centers are smaller than mi_gaussian_dist, continue to the next combination
        if(TRUE %in% (diff(gaussians[combination[,j],"mu"]) <= min_gaussian_dist)) next
        
        for(a in combination[,j]){
          model <- model + gaussians$A[[a]]*exp(-(fraction_numbers - gaussians$mu[[a]])**2/(gaussians$sigma[[a]]**2))
        }
        
        #Least squares, if lower than currently best candidate, replace with current model
        square_deviations <- sum((chromatogram-model)**2)
        if(best_square_deviations > square_deviations){
          best_square_deviations <- square_deviations
          best_combination <- combination[,j]
          best_corr <- cor(chromatogram, model, method="pearson")
          best_model <- model
        }
      }
    }
    best_fits <- best_fits %>% rbind(data.frame(gaussians[best_combination,], model_fold=sum(best_model)/sum(chromatogram), model_corr=best_corr))
  }
  close(pb)
  best_fits
}

windowConsensus <- createConsensusFromWindowFitting(PCM18_integrated, joinedGaussians, min_gaussian_dist=1)

windowConsensus_noWindowFitting <- createConsensusFromWindowFitting(PCM18_integrated, joinedGaussians_noWindowFitting, min_gaussian_dist=1)
```


```{r}
calculateGaussianMixture <- function(gaussians, low, high, step=1){
  x_values <- seq(low,high,step)
  model <- double(length=length(x_values))
      for(a in 1:nrow(gaussians)){
        model <- model + gaussians$A[[a]]*exp(-(x_values - gaussians$mu[[a]])**2/(gaussians$sigma[[a]]**2))
      }
  model
}

#Min peak fraction controls what fraction of the signal in a given chromatographic fraction must come from a given peak to be included in the comparison. Zcriterion tells the percent of the peak area that will be included in the comparison (normal distribution)
gaussianAreaComparison <- function(windowConsensus, data_ctrl, data_treat, z_criterion = 1.645, min_peak_fraction = 0.5, peptideTraces){
  nfrac <- 
  if(ncol(data_ctrl) != ncol(data_treat)){
    stop("data_ctrl and data_treat have unequal number of columns!")
  }
  prince_fc <- data.frame(protein_id=windowConsensus$protein_id, mu=windowConsensus$mu, low_lim=double(length=nrow(windowConsensus)), high_lim=double(length=nrow(windowConsensus)), excluded_frac=character(length=nrow(windowConsensus)), n_frac=integer(length=nrow(windowConsensus)), n_pep=integer(length=nrow(windowConsensus)),  weighted_corr=double(length=nrow(windowConsensus)), foldChange=double(length=nrow(windowConsensus)), area_ctrl = double(length=nrow(windowConsensus)), area_treat = double(length=nrow(windowConsensus)), log2area_ctrl= double(length=nrow(windowConsensus)), log2area_treat= double(length=nrow(windowConsensus)), percent_areactrl=double(length=nrow(windowConsensus)), percent_areatreat=double(length=nrow(windowConsensus)))
  for(rowN in 1:nrow(windowConsensus)){
    
    pepMatrix_selectedProt <- as.matrix(subset(peptideTraces, trace_subset_ids=windowConsensus[rowN, "protein_id"], trace_subset_type="protein_id")$traces[,1:ncol(data_ctrl)])
    
    low_lim <- round(windowConsensus[[rowN, 3]] - z_criterion*windowConsensus[[rowN, 4]]/sqrt(2))
    low_lim <- ifelse(low_lim < 1, 1, low_lim)
    high_lim <- round(windowConsensus[[rowN, 3]] + z_criterion*windowConsensus[[rowN, 4]]/sqrt(2))
    high_lim <- ifelse(high_lim > ncol(data_ctrl), ncol(data_ctrl), high_lim)
    prince_fc[rowN,"low_lim"] <- low_lim
    prince_fc[rowN,"high_lim"] <- high_lim
    protein_gaussians <- filter(windowConsensus, protein_id == windowConsensus[rowN, "protein_id"])
    current_gaussian_model <- calculateGaussianMixture(windowConsensus[rowN,],1,ncol(data_ctrl),step=1)
    full_model <- calculateGaussianMixture(protein_gaussians, 1, ncol(data_ctrl), step=1)
    excluded_fractions <- c()
    for(frac in low_lim:high_lim){
      if(current_gaussian_model[[frac]]/full_model[[frac]] < min_peak_fraction){
        excluded_fractions <- excluded_fractions %>% append(frac)
      }
    }
    compared_fractions <- setdiff(low_lim:high_lim, excluded_fractions) # This are the indices of the fractions that will be used for area comparison
    
    prince_fc[rowN, "excluded_frac"] <- paste(excluded_fractions, collapse = "; ")
    prince_fc[rowN, "n_frac"] <- length(compared_fractions)
    n_pep <- 0
    
    for(i in 1:nrow(pepMatrix_selectedProt)){
      if(TRUE %in% (pepMatrix_selectedProt[i, compared_fractions] > 0)) {
        n_pep <- n_pep + 1
      }
    }
    
    prince_fc[rowN, "n_aSYN"] <- n_pep
    
    #suppressWarnings(prince_fc[rowN, "corr"] <- try((sum(cor(t(pepMatrix_selectedProt[,compared_fractions])), na.rm=T)-nrow(pepMatrix_selectedProt))/(n_pep**2 - n_pep), silent=T))
    corr_matrix <- NA
    weight_matrix <- NA
    suppressWarnings(corr_matrix <- try(cor(t(pepMatrix_selectedProt[,compared_fractions])), silent=T))
    try(weight_matrix <- outer(rowSums(pepMatrix_selectedProt[,compared_fractions]), rowSums(pepMatrix_selectedProt[,compared_fractions]), FUN="*"), silent=T)
    try(diag(corr_matrix) <- NA, silent=T)
    try(diag(weight_matrix) <- NA, silent=T)
    
    prince_fc[rowN, "weighted_corr"] <- sum(corr_matrix*weight_matrix, na.rm=T)/sum(weight_matrix, na.rm=T)
    
    if(windowConsensus[[rowN, 1]] %in% rownames(data_ctrl)){
      area_control <- sum(data_ctrl[windowConsensus[[rowN, 1]], compared_fractions])
      prince_fc[rowN,"percent_areactrl"] <- 100*area_control/sum(data_ctrl[windowConsensus[[rowN, 1]],])
    } else {
      area_control <- 0
      prince_fc[rowN,"percent_areactrl"] <- NA
    }
    if(windowConsensus[[rowN, 1]] %in% rownames(data_treat)) {
      area_ATP <- sum(data_treat[windowConsensus[[rowN, 1]], compared_fractions])
      prince_fc[rowN,"percent_areatreat"] <- 100*area_ATP/sum(data_treat[windowConsensus[[rowN, 1]],])
  
    } else {
      area_ATP <- 0
      prince_fc[rowN,"percent_areatreat"] <- NA
  
    }
    prince_fc[rowN,"foldChange"] <- log2(area_ATP/area_control)
    prince_fc[rowN,"area_ctrl"] <- area_control
    prince_fc[rowN,"area_treat"] <- area_ATP
    prince_fc[rowN,"log2area_ctrl"] <- log2(area_control)
    prince_fc[rowN,"log2area_treat"] <- log2(area_ATP)
    
  }
  prince_fc
}

protein_traces_list_30pep <- proteinQuantification_sibPepCorrFix.tracesList(pepTracesList_filtered ,
                                             quantLevel="protein_id",
                                             topN = 30,
                                             keep_less = TRUE, #remove proteins with less then 5 precursors
                                             rm_decoys = TRUE,
                                             use_sibPepCorr = TRUE,
                                             use_repPepCorr = FALSE,
                                             full_intersect_only = FALSE,
                                             verbose = FALSE)

PCM18_biolRep1_ctrl <-  as.matrix(protein_traces_list_30pep$ctrl_1$traces[,1:32])
rownames(PCM18_biolRep1_ctrl) <- protein_traces_list_30pep$ctrl_1$traces$id
PCM18_biolRep1_treat <-  as.matrix(protein_traces_list_30pep$PEP_1$traces[,1:32])
rownames(PCM18_biolRep1_treat) <- protein_traces_list_30pep$PEP_1$traces$id

PCM18_biolRep2_ctrl <-  as.matrix(protein_traces_list_30pep$ctrl_2$traces[,1:32])
rownames(PCM18_biolRep2_ctrl) <- protein_traces_list_30pep$ctrl_2$traces$id
PCM18_biolRep2_treat <-  as.matrix(protein_traces_list_30pep$PEP_2$traces[,1:32])
rownames(PCM18_biolRep2_treat) <- protein_traces_list_30pep$PEP_2$traces$id

PCM18_biolRep3_ctrl <-  as.matrix(protein_traces_list_30pep$ctrl_3$traces[,1:32])
rownames(PCM18_biolRep3_ctrl) <- protein_traces_list_30pep$ctrl_3$traces$id
PCM18_biolRep3_treat <-  as.matrix(protein_traces_list_30pep$PEP_3$traces[,1:32])
rownames(PCM18_biolRep3_treat) <- protein_traces_list_30pep$PEP_3$traces$id

biolRep1_fc <- gaussianAreaComparison(windowConsensus, PCM18_biolRep1_ctrl, PCM18_biolRep1_treat, peptideTraces = integrateTraceIntensities(pepTracesList_filtered))
biolRep2_fc <- gaussianAreaComparison(windowConsensus, PCM18_biolRep2_ctrl, PCM18_biolRep2_treat, peptideTraces = integrateTraceIntensities(pepTracesList_filtered))
biolRep3_fc <- gaussianAreaComparison(windowConsensus, PCM18_biolRep3_ctrl, PCM18_biolRep3_treat, peptideTraces = integrateTraceIntensities(pepTracesList_filtered))

foldChanges <- full_join(biolRep1_fc, biolRep2_fc, by=join_by(protein_id, mu, low_lim, high_lim, excluded_frac), suffix = c("", "_rep2")) %>% full_join(biolRep3_fc, by=join_by(protein_id, mu, low_lim, high_lim, excluded_frac), suffix = c("_rep1", "_rep3"))

foldChanges$ttest_pval <- 1


for(rowN in 1:nrow(foldChanges)){
  foldChanges[rowN, "ttest_pval"] <- try(t.test(foldChanges[rowN, c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3")], foldChanges[rowN, c("log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3")])$p.value, silent = T)
}

foldChanges$limma_pval <- (limma::eBayes(limma::lmFit(foldChanges[c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3", "log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3")], design=model.matrix(~factor(c(0,0,0,1,1,1))))))$p.value[,2] %>% as.numeric()

foldChanges$limma_pBHadj <- foldChanges$limma_pval %>% p.adjust(method="fdr")


foldChanges$meanFCnoInf <- rowMeans(data.frame(ifelse(foldChanges$foldChange_rep1 %>% is.infinite(), NA, foldChanges$foldChange_rep1), ifelse(foldChanges$foldChange_rep2 %>% is.infinite(), NA, foldChanges$foldChange_rep2), ifelse(foldChanges$foldChange_rep3 %>% is.infinite(), NA, foldChanges$foldChange_rep3)), na.rm=T)

foldChanges <- foldChanges %>% left_join(select(uniprot, accession, protein_name, mass, gene_names, go_f, go_p, go_c, ft_act_site, ft_binding, cc_cofactor), by=join_by(protein_id == accession))

save.image()
```


##Traces of changed proteins
```{r}
for(i in unique(filter(foldChanges, limma_pBHadj < 0.05)$protein_id)){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = paste(i, round(filter(foldChanges, limma_pBHadj < 0.05)$mu), collapse=" "))
} #not saved in the nas

```

##No window fitting comparison
```{r}
biolRep1_fc_noWindowFitting <- gaussianAreaComparison(windowConsensus_noWindowFitting, PCM18_biolRep1_ctrl, PCM18_biolRep1_treat, peptideTraces = integrateTraceIntensities(pepTracesList_filtered))
biolRep2_fc_noWindowFitting <- gaussianAreaComparison(windowConsensus_noWindowFitting, PCM18_biolRep2_ctrl, PCM18_biolRep2_treat, peptideTraces = integrateTraceIntensities(pepTracesList_filtered))
biolRep3_fc_noWindowFitting <- gaussianAreaComparison(windowConsensus_noWindowFitting, PCM18_biolRep3_ctrl, PCM18_biolRep3_treat, peptideTraces = integrateTraceIntensities(pepTracesList_filtered))

foldChanges_noWindowFitting <- joinAreasAndFoldChanges(replicateAreas = list(biolRep1_fc_noWindowFitting, biolRep2_fc_noWindowFitting, biolRep3_fc_noWindowFitting), min_npep=5, min_weighted_corr=0.2, exclude_proteins=setdiff(proteins_to_exclude, proteins_to_keep)
)
#setdiff(proteins_to_exclude, proteins_to_keep)

foldChanges_noWindowFitting$ttest_pval <- 1


for(rowN in 1:nrow(foldChanges_noWindowFitting)){
  foldChanges_noWindowFitting[rowN, "ttest_pval"] <- try(t.test(foldChanges_noWindowFitting[rowN, c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3")], foldChanges_noWindowFitting[rowN, c("log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3")])$p.value, silent = T)
}

foldChanges_noWindowFitting$ttest_pval <-as.numeric(foldChanges_noWindowFitting$ttest_pval)

foldChanges_noWindowFitting$ttest_pBHadj <- foldChanges_noWindowFitting$ttest_pval %>% p.adjust(method="fdr")

foldChanges_noWindowFitting$limma_pval <- (limma::eBayes(limma::lmFit(foldChanges_noWindowFitting[c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3", "log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3")], design=model.matrix(~factor(c(0,0,0,1,1,1))))))$p.value[,2] %>% as.numeric()

foldChanges_noWindowFitting$limma_pBHadj <- foldChanges_noWindowFitting$limma_pval %>% p.adjust(method="fdr")

foldChanges_noWindowFitting$ibmt_pval <- (IBMT(limma::eBayes(limma::lmFit(foldChanges_noWindowFitting[,c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3", "log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3")] - log2(foldChanges_noWindowFitting[,"n_frac"]), design=model.matrix(~factor(c(0,0,0,1,1,1))))), 1:2))$IBMT.p[,2] %>% as.numeric()

foldChanges_noWindowFitting$ibmt_pBHadj <- foldChanges_noWindowFitting$ibmt_pval %>% p.adjust(method="fdr")

foldChanges_noWindowFitting$meanFCnoInf <- rowMeans(data.frame(ifelse(foldChanges_noWindowFitting$foldChange_rep1 %>% is.infinite(), NA, foldChanges_noWindowFitting$foldChange_rep1), ifelse(foldChanges_noWindowFitting$foldChange_rep2 %>% is.infinite(), NA, foldChanges_noWindowFitting$foldChange_rep2), ifelse(foldChanges_noWindowFitting$foldChange_rep3 %>% is.infinite(), NA, foldChanges_noWindowFitting$foldChange_rep3)), na.rm=T)

foldChanges_noWindowFitting <- foldChanges_noWindowFitting %>% left_join(select(uniprot, accession, protein_name, mass, gene_names, go_f, go_p, go_c, ft_act_site, ft_binding, cc_cofactor), by=join_by(protein_id == accession))
```

##Traces of changed proteins (no window fitting)
```{r}
for(i in unique(filter(foldChanges_noWindowFitting, ibmt_pBHadj < 0.005 & abs(meanFCnoInf) > 1)$protein_id)){
  pepTest <- subset(pepTracesList_filtered, trace_subset_ids = i, trace_subset_type = "protein_id")
  plot(pepTest, design_matrix = design_matrix, log = FALSE, legend = F, PDF=F, name = paste(unique(filter(foldChanges_noWindowFitting, protein_id==i)$protein_id), paste(round(filter(foldChanges_noWindowFitting, protein_id==i & ibmt_pBHadj < 0.005)$mu), collapse=" ")))
} #not saved in the nas

```



## CCprofiler borders with limma

```{r}
ccprofilerFeatureAreaComparison <- function(proteinFeatures, data_ctrl, data_treat, peptideTraces){
  prince_fc <- data.frame(protein_id=proteinFeatures$protein_id, apex=proteinFeatures$apex, low_lim=double(length=nrow(proteinFeatures)), high_lim=double(length=nrow(proteinFeatures)), excluded_frac=character(length=nrow(proteinFeatures)), n_frac=integer(length=nrow(proteinFeatures)),n_pep=integer(length=nrow(proteinFeatures)), weighted_corr=double(length=nrow(proteinFeatures)), foldChange=double(length=nrow(proteinFeatures)), area_ctrl = double(length=nrow(proteinFeatures)), area_treat = double(length=nrow(proteinFeatures)), log2area_ctrl= double(length=nrow(proteinFeatures)), log2area_treat= double(length=nrow(proteinFeatures)), percent_areactrl=double(length=nrow(proteinFeatures)), percent_areatreat=double(length=nrow(proteinFeatures)))
  
  
  for(rowN in 1:nrow(proteinFeatures)){
    low_lim <- as.integer(proteinFeatures[rowN, "left_pp"])
    high_lim <- as.integer(proteinFeatures[rowN, "right_pp"])
    prince_fc[rowN,"low_lim"] <- low_lim
    prince_fc[rowN,"high_lim"] <- high_lim
    compared_fractions <- low_lim:high_lim # This are the indices of the fractions that will be used for area comparison
    prince_fc[rowN, "excluded_frac"] <- ""
    prince_fc[rowN, "n_frac"] <- length(compared_fractions)
    
    if(proteinFeatures[[rowN, 1]] %in% rownames(data_ctrl)){
      area_control <- sum(data_ctrl[proteinFeatures[[rowN, 1]], compared_fractions])
      prince_fc[rowN,"percent_areactrl"] <- 100*area_control/sum(data_ctrl[proteinFeatures[[rowN, 1]],])
    } else {
      area_control <- 0
      prince_fc[rowN,"percent_areactrl"] <- NA
    }
    if(proteinFeatures[[rowN, 1]] %in% rownames(data_treat)) {
      area_ATP <- sum(data_treat[proteinFeatures[[rowN, 1]], compared_fractions])
      prince_fc[rowN,"percent_areatreat"] <- 100*area_ATP/sum(data_treat[proteinFeatures[[rowN, 1]],])
  
    } else {
      area_ATP <- 0
      prince_fc[rowN,"percent_areatreat"] <- NA
  
    }
    prince_fc[rowN,"foldChange"] <- log2(area_ATP/area_control)
    prince_fc[rowN,"area_ctrl"] <- area_control
    prince_fc[rowN,"area_treat"] <- area_ATP
    prince_fc[rowN,"log2area_ctrl"] <- log2(area_control)
    prince_fc[rowN,"log2area_treat"] <- log2(area_ATP)
    
  }
  prince_fc
}


ccprofiler_rep1_fc <- ccprofilerFeatureAreaComparison(proteinFeaturesFiltered, PCM18_biolRep1_ctrl, PCM18_biolRep1_treat, peptideTraces=integrateTraceIntensities(pepTracesList_filtered))

ccprofiler_rep2_fc <- ccprofilerFeatureAreaComparison(proteinFeaturesFiltered, PCM18_biolRep2_ctrl, PCM18_biolRep2_treat, peptideTraces=integrateTraceIntensities(pepTracesList_filtered))

  ccprofiler_rep3_fc <- ccprofilerFeatureAreaComparison(proteinFeaturesFiltered, PCM18_biolRep3_ctrl, PCM18_biolRep3_treat, peptideTraces=integrateTraceIntensities(pepTracesList_filtered))

ccprofiler_rep1_fc <- ccprofiler_rep1_fc %>% mutate(mu=apex)
ccprofiler_rep2_fc <- ccprofiler_rep2_fc %>% mutate(mu=apex)
ccprofiler_rep3_fc <- ccprofiler_rep3_fc %>% mutate(mu=apex)


ccprofiler_joinedAreas <- joinAreasAndFoldChanges(list(ccprofiler_rep1_fc, ccprofiler_rep2_fc, ccprofiler_rep3_fc), min_npep=0, min_weighted_corr=-1)

ccprofiler_joinedAreas$limma_pval <- (limma::eBayes(limma::lmFit(ccprofiler_joinedAreas[c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3", "log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3")], design=model.matrix(~factor(c(0,0,0,1,1,1))))))$p.value[,2] %>% as.numeric()

ccprofiler_joinedAreas$ttest_pval <- 1

for(i in 1:nrow(ccprofiler_joinedAreas)){
  ccprofiler_joinedAreas[i, "ttest_pval"] <- try(t.test(as.numeric(ccprofiler_joinedAreas[i, c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3", "log2area_ctrl_rep4")]), as.numeric(ccprofiler_joinedAreas[i, c("log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3", "log2area_treat_rep4")]))$p.val, silent=T)
}

ccprofiler_joinedAreas$limma_pBHadj <- ccprofiler_joinedAreas$limma_pval %>% p.adjust(method="fdr")

ccprofiler_joinedAreas$ttest_pBHadj <- ccprofiler_joinedAreas$ttest_pval %>% p.adjust(method="fdr")


ccprofiler_joinedAreas$meanLog2FCnoInf <-rowMeans(data.frame(ifelse(ccprofiler_joinedAreas$foldChange_rep1 %>% is.infinite(), NA, ccprofiler_joinedAreas$foldChange_rep1), ifelse(ccprofiler_joinedAreas$foldChange_rep2 %>% is.infinite(), NA, ccprofiler_joinedAreas$foldChange_rep2), ifelse(ccprofiler_joinedAreas$foldChange_rep3 %>% is.infinite(), NA, ccprofiler_joinedAreas$foldChange_rep3)), na.rm=T)

ccprofiler_joinedAreas$meanLog2FC <-rowMeans(data.frame(ccprofiler_joinedAreas$foldChange_rep1, ccprofiler_joinedAreas$foldChange_rep2, ccprofiler_joinedAreas$foldChange_rep3))

ccprofiler_joinedAreas <- ccprofiler_joinedAreas %>% left_join(select(uniprot, accession, protein_name, mass, go_f, go_p, go_c), by=join_by(protein_id==accession))
```
## QC - correlations between chromatograms

```{r}
calculateChromatogramCorrelations <- function(traces1, traces2, accessions=NULL, start_fraction=1, end_fraction=ncol(traces1)-1){
  correlations <- data.frame(matrix(nrow=uniqueN(accessions), ncol=5))
  names(correlations) <- c("accession", "correlation", "sum", "sum_cond1", "sum_cond2")
  i <- 1
  if(is.null(accessions)){
    used_accessions <- intersect(traces1$id, traces2$id)
  } else {
    used_accessions <- accessions
  }
  for(acc in used_accessions){
    selected1 <- traces1 %>% filter(id == acc) %>% select((start_fraction):(end_fraction))
    selected2 <- traces2 %>% filter(id == acc) %>% select((start_fraction):(end_fraction))
    if(ifelse(is.na(all(as.numeric(selected1)==0) | all(as.numeric(selected2)==0)), T, all(as.numeric(selected1)==0) | all(as.numeric(selected2)==0))) {
      next
    }
    correlations[i,1]<- acc
    correlations[i,2]<- cor(as.numeric(selected1),as.numeric(selected2), method="pearson")
    correlations[i,3] <- sum(as.numeric(selected1)) + sum(as.numeric(selected2))
    correlations[i,4] <- sum(as.numeric(selected1)) 
    correlations[i,5] <- sum(as.numeric(selected2))
    i <- i+1
  }
  correlations
}

corr_ctrl1_ctrl2 <- calculateChromatogramCorrelations(protein_traces_list$ctrl_1$traces, protein_traces_list$ctrl_2$traces)
corr_ctrl1_ctrl3 <- calculateChromatogramCorrelations(protein_traces_list$ctrl_1$traces, protein_traces_list$ctrl_3$traces)
corr_ctrl2_ctrl3 <- calculateChromatogramCorrelations(protein_traces_list$ctrl_2$traces, protein_traces_list$ctrl_3$traces)

corr_aSYN1_aSYN2 <- calculateChromatogramCorrelations(protein_traces_list$PEP_1$traces, protein_traces_list$PEP_2$traces)
corr_aSYN1_aSYN3 <- calculateChromatogramCorrelations(protein_traces_list$PEP_1$traces, protein_traces_list$PEP_3$traces)
corr_aSYN2_aSYN3 <- calculateChromatogramCorrelations(protein_traces_list$PEP_2$traces, protein_traces_list$PEP_3$traces)

corr_ctrl1_aSYN2 <- calculateChromatogramCorrelations(protein_traces_list$ctrl_1$traces, protein_traces_list$PEP_2$traces)
corr_ctrl1_aSYN3 <- calculateChromatogramCorrelations(protein_traces_list$ctrl_1$traces, protein_traces_list$PEP_3$traces)
corr_ctrl2_aSYN3 <- calculateChromatogramCorrelations(protein_traces_list$ctrl_2$traces, protein_traces_list$PEP_3$traces)

corr_aSYN1_ctrl2 <- calculateChromatogramCorrelations(protein_traces_list$PEP_1$traces, protein_traces_list$ctrl_2$traces)
corr_aSYN1_ctrl3 <- calculateChromatogramCorrelations(protein_traces_list$PEP_1$traces, protein_traces_list$ctrl_3$traces)
corr_aSYN2_ctrl3 <- calculateChromatogramCorrelations(protein_traces_list$PEP_2$traces, protein_traces_list$ctrl_3$traces)

correlations_dataframe <- rbind(data.frame(class="ctrl_ctrl", comparison="01_ctrl1_ctrl2", correlations=corr_ctrl1_ctrl2$correlation), data.frame(class="ctrl_ctrl", comparison="02_ctrl1_ctrl3", correlations=corr_ctrl1_ctrl3$correlation)) %>% rbind(data.frame(class="ctrl_ctrl", comparison="03_ctrl2_ctrl3", correlations=corr_ctrl2_ctrl3$correlation)) %>% rbind(data.frame(class="aSYN_aSYN", comparison="04_aSYN1_aSYN2", correlations=corr_aSYN1_aSYN2$correlation)) %>% rbind(data.frame(class="aSYN_aSYN", comparison="05_aSYN1_aSYN3", correlations=corr_aSYN1_aSYN3$correlation)) %>% rbind(data.frame(class="aSYN_aSYN", comparison="06_aSYN2_aSYN3", correlations=corr_aSYN2_aSYN3$correlation)) %>% rbind(data.frame(class="ctrl_aSYN", comparison="07_ctrl1_aSYN2", correlations=corr_ctrl1_aSYN2$correlation)) %>% rbind(data.frame(class="ctrl_aSYN", comparison="08_ctrl1_aSYN3", correlations=corr_ctrl1_aSYN3$correlation)) %>% rbind(data.frame(class="ctrl_aSYN", comparison="10_ctrl2_aSYN3", correlations=corr_ctrl2_aSYN3$correlation)) %>% rbind(data.frame(class="ctrl_aSYN", comparison="09_ctrl2_aSYN1", correlations=corr_aSYN1_ctrl2$correlation)) %>% rbind(data.frame(class="ctrl_aSYN", comparison="11_ctrl3_aSYN1", correlations=corr_aSYN1_ctrl3$correlation)) %>% rbind(data.frame(class="ctrl_aSYN", comparison="12_ctrl3_aSYN2", correlations=corr_aSYN2_ctrl3$correlation)) %>% group_by(class) %>% arrange(factor(class, levels = c("ctrl_ctrl", "aSYN_aSYN", "ctrl_aSYN"))) %>% mutate(class = fct(class))

ggplot(data = correlations_dataframe, mapping = aes(x=comparison, y=correlations, fill=class)) + geom_violin() + geom_boxplot(width=0.07, color="black", alpha=0.6, outliers=F)


ctrl_stdevs <- apply(foldChanges_noWindowFitting[,c("log2area_ctrl_rep1", "log2area_ctrl_rep2", "log2area_ctrl_rep3")], 1, sd, na.rm=T)

PEP_stdevs <- apply(foldChanges_noWindowFitting[,c("log2area_treat_rep1", "log2area_treat_rep2", "log2area_treat_rep3")], 1, sd, na.rm=T)

foldChanges_noWindowFittingQC <- foldChanges_noWindowFitting %>% select(1:29)

foldChanges_noWindowFittingQC$sd_ctrl <- ctrl_stdevs
foldChanges_noWindowFittingQC$sd_treat <- PEP_stdevs

```


```{r}
ggplot(data=protein_DiffExprProtein, mapping = aes(x=pVal)) + geom_histogram(binwidth=0.01)

ggplot(data=ccprofiler_joinedAreas, mapping = aes(x=limma_pval)) + geom_histogram(binwidth=0.01)

ggplot(data=foldChanges_noWindowFitting, mapping = aes(x=ibmt_pval)) + geom_histogram(binwidth=0.01)

ggplot(data=foldChanges_noWindowFitting, mapping = aes(x=limma_pval)) + geom_histogram(binwidth=0.01)

ggplot(data=foldChanges_noWindowFitting, mapping = aes(x=as.numeric(ttest_pval))) + geom_histogram(binwidth=0.01)

```

## Positions of changed peaks
TODO highlight the hits in red on the first plot to simplify

```{r}
ggplot(data=foldChanges_noWindowFitting) + geom_point(mapping=aes(x=mu, y=meanFCnoInf)) + ylim(-7, 7) + labs(title="All PrInCE-detected peaks - aSYN") + xlab("Peak center [fraction nr.]") + ylab("Log2 fold change") + xlim(0,32)
ggplot(data=filter(foldChanges_noWindowFitting, ibmt_pBHadj < 0.05)) + geom_point(mapping=aes(x=mu, y=meanFCnoInf), colour="red") + ylim(-7, 7) + labs(title="PrInCE-detected hits - aSYN") + xlab("Peak center [fraction nr.]") + ylab("Log2 fold change") + xlim(0,32) 

ggplot(data=foldChanges_noWindowFitting) + geom_point(mapping=aes(x=mu, y=meanFCnoInf, colour="Non-hits")) + geom_point(data = filter(foldChanges_noWindowFitting, ibmt_pBHadj < 0.05) , mapping=aes(x=mu, y=meanFCnoInf, colour="Hits (q < 0.05)")) + ylim(-7, 7) + labs(title="All PrInCE-detected peaks - aSYN") + xlab("Peak center [fraction nr.]") + ylab("Log2 fold change") + xlim(0,32)  + scale_colour_manual(values = c("Non-hits"="black", "Hits (q < 0.05)"="red"))
```
##GO enrichment
```{r}
foldChanges_noWindowFitting <- foldChanges_noWindowFitting %>% mutate(isHit = (ibmt_pBHadj < 0.05))

calculate_go_enrichment(foldChanges_noWindowFitting, protein_id = protein_id, is_significant = isHit, go_annotations_uniprot = go_f, plot_cutoff = "pval top10")
```


### Annotate hits as PEP-binding or not
```{r}
IlariaPiazzaLiPData <- read_excel("IlariaPiazza_LiPData.xlsx", sheet="aSYN")
IlariaPiazzaAnnotations <- read_excel("IlariaPiazza_LiPData.xlsx", sheet="Encyc") %>% filter(`LiP-SMap_expID` == "aSYN")

#Get LiP hits, but without those validated only at lower concentration
IlariaPiazzaLiPHits <- IlariaPiazzaLiPData %>% filter((is_C1_validated | is_C2_validated) & !(is_C1_validated & !is_C2_validated)) %>% select(UniProtKB_ID)

foldChanges_noWindowFitting <- foldChanges_noWindowFitting %>% mutate(isAnnotatedMetaboliteInteractor = protein_id %in% IlariaPiazzaAnnotations$`Uniprot ID`) %>% mutate(isLiPHit = protein_id %in% IlariaPiazzaLiPHits$UniProtKB_ID)

ccprofiler_joinedAreas <- ccprofiler_joinedAreas %>% mutate(isAnnotatedMetaboliteInteractor = protein_id %in% IlariaPiazzaAnnotations$`Uniprot ID`) %>% mutate(isLiPHit = protein_id %in% IlariaPiazzaLiPHits$UniProtKB_ID)

overall_annotated_number_gaussian <- uniqueN(filter(foldChanges_noWindowFitting, isAnnotatedMetaboliteInteractor | isLiPHit)$protein_id)

overall_nonannotated_number_gaussian <- uniqueN(filter(foldChanges_noWindowFitting, !(isAnnotatedMetaboliteInteractor | isLiPHit))$protein_id)

hit_annotated_number_gaussian <- uniqueN(filter(foldChanges_noWindowFitting, ibmt_pBHadj < 0.0049 & (isAnnotatedMetaboliteInteractor | isLiPHit))$protein_id)

hit_nonannotated_number_gaussian <- uniqueN(filter(foldChanges_noWindowFitting, ibmt_pBHadj < 0.0049 & !(isAnnotatedMetaboliteInteractor | isLiPHit))$protein_id)
                                                
```


## Overlap with Ilaria Piazza hits
```{r}

foldChanges_noWindowFitting_withLiPData <- foldChanges_noWindowFitting %>% filter(!is.na(ibmt_pBHadj)) %>% select(protein_id, mu, ibmt_pval, ibmt_pBHadj, meanFCnoInf, protein_name, go_f, isAnnotatedMetaboliteInteractor, isLiPHit) %>% left_join(select(IlariaPiazzaLiPData, UniProtKB_ID, is_C1_validated, is_C2_validated, is_known_alloTarget, is_known_reactant, is_known_notMetTarget, N_quantif_peps, protein_coverage), by=join_by(protein_id == UniProtKB_ID), suffix=c("", "_IlariaPiazza"))

foldChanges_noWindowFitting_withLiPData_toSave <- filter(foldChanges_noWindowFitting_withLiPData, ibmt_pBHadj < 0.1 | is_C1_validated | is_C2_validated | isAnnotatedMetaboliteInteractor)

foldChanges_noWindowFitting_withLiPData_toSave <- foldChanges_noWindowFitting_withLiPData_toSave[order(foldChanges_noWindowFitting_withLiPData_toSave$ibmt_pval),]

foldChanges_noWindowFitting_withLiPData_toSave[is.na(foldChanges_noWindowFitting_withLiPData_toSave)] <- "NA"

write_xlsx(foldChanges_noWindowFitting_withLiPData_toSave, path="aSYN_SEChits_withIlariaPiazzaData.xlsx")
```

# Spectronaut report processing for Marc (MaxLFQ and ibaq protein quantity)
```{r read-and-process-data}
library(tidyverse)
# Read the TSV file
data_PCM16 <- read_tsv("C:/Users/ailaria/Projects/PCM/20251215_PCM18_pg_ibaq_Report.tsv")
# Select columns and extract information from r_FileName
processed_data_PCM18 <- data_PCM16 %>%
  select(R.FileName, PG.IBAQ, PG.Quantity, UniprotId = PG.ProteinAccessions) %>%
  mutate(
    # Extract replicate number from "biolrep#"
    R.Replicate = str_extract(R.FileName, "biolrep(\\d+)") %>% 
      str_extract("\\d+") %>% 
      as.integer(),
    
    # Extract fraction number from "f##"
    R.Fraction = str_extract(R.FileName, "_f(\\d+)_") %>% 
      str_extract("\\d+") %>% 
      as.integer(),
    
    # Extract condition (ctrl or PEP) - looks for these words in the filename
    R.Condition = str_extract(R.FileName, "ctrl|ADP")
  )
# Save the processed data to a TSV file
write_tsv(processed_data_PCM18, "C:/Users/ailaria/Projects/PCM/PCM16_pg_ibaq_maxLFQ_essential_columns_report.tsv")
# View the result
head(processed_data_PCM18)
```


# aSYN data processing 
```{r}
library(tidyverse)

# Define the file path
base_path <- "C:/Users/ailaria/Projects/alpha-synuclein/processing_data_Marc/"  # Update this to your actual path

# List of file names
files <- c(
  "20250715_IA003_SEC_SILAC_BR_GFP_BR1_GAL_H10_Report_AP_scheme_qvalue_rawquant (Normal).tsv",
  "20250715_IA003_SEC_SILAC_BR_GFP_BR2_GAL_H10_Report_AP_scheme_qvalue_rawquant (Normal).tsv",
  "20250715_IA003_SEC_SILAC_BR_GFP_BR3_GAL_H10_Report_AP_scheme_qvalue_rawquant (Normal).tsv",
  "20250715_IA003_SEC_SILAC_BR_InTox_BR1_GAL_H10_Report_AP_scheme_qvalue_rawquant (Normal).tsv",
  "20250715_IA003_SEC_SILAC_BR_InTox_BR2_GAL_H10_Report_AP_scheme_qvalue_rawquant (Normal).tsv",
  "20250715_IA003_SEC_SILAC_BR_InTox_BR3_GAL_H10_Report_AP_scheme_qvalue_rawquant (Normal).tsv"
)

# Function to read and process each file
process_file <- function(file_name, base_path) {
  read_tsv(paste0(base_path, file_name), show_col_types = FALSE) %>%
    select(r_FileName = `R.FileName`, pg_Quantity = `PG.Quantity`, UniprotId = `PG.ProteinAccessions`) %>%
    mutate(
      # Extract replicate (BR1, BR2, BR3)
      r_Replicate = str_extract(r_FileName, "BR[1-3]"),
      
      # Extract fraction (f1, f2, ..., f32)
      r_Fraction = str_extract(r_FileName, "f\\d+"),
      
      # Extract condition and rename: GFP -> ctrl, InTox -> aSYN
      r_Condition = case_when(
        str_detect(r_FileName, "GFP") ~ "ctrl",
        str_detect(r_FileName, "InTox") ~ "aSYN",
        TRUE ~ NA_character_
      )
    )
}

# Read, process, and combine all files
processed_data <- files %>%
  map_dfr(~ process_file(.x, base_path))

# Save the merged data
write_tsv(processed_data, paste0(base_path, "aSYN_10h_pg_essential_columns_report.tsv"))

# View the result
head(processed_data)
cat("\nTotal rows:", nrow(processed_data))
colnames(processed_data)

```


# check ctrl correlations among all experimetns
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/ailaria/Projects/PCM/replicates comparison")

# Set working directory to where your data files are
setwd("C:/Users/ailaria/Projects/PCM/replicates comparison")

# Now run the analysis
#source("CFMS_Control_QC_Analysis_Adapted.R")
# Source the module
source("C:/Users/ailaria/Projects/PCM/replicates comparison/CFMS_Control_QC_ElutionAware_finalChecks.R")

files <- c(
  PEP = "C:/Users/ailaria/Projects/PCM/replicates comparison/20251126_114514_20251021_AP_scheme_qvalue_rawquant_Report.tsv",
  aSYN = "C:/Users/ailaria/Projects/PCM/replicates comparison/20251202_PCM18_AP_qvalue_rawquant.tsv",
  NAD = "C:/Users/ailaria/Projects/PCM/replicates comparison/20260108_PCM19_NAD_Anna_rawquant_Report.tsv",
  ATP = "C:/Users/ailaria/Projects/PCM/replicates comparison/20251127_PCM20_directDIA_Anna_SEC_Report.tsv",
  PGP = "C:/Users/ailaria/Projects/PCM/replicates comparison/20251215_PCM21_pg_ibaq_Anna_rawquant_Report.tsv"
)


# ════════════════════════════════════════════════════════════
# FAST RUN (~5-15 minutes) - Use this first!
# ════════════════════════════════════════════════════════════
results_fast <- run_control_qc_elution(
  file_paths = files,
  ctrl_condition = "ctrl",
  output_dir = "C:/Users/ailaria/Projects/PCM/replicates comparison/control_qc_fast_additionalChecks", # CHECK NAMING IST CORRECT AND MATCHES SOURCE
  methods = c("sum", "median"),     # Fewer methods
  run_elution_analysis = TRUE,
  fit_gaussian = FALSE,             # Weighted mean (FAST!)
  sample_frac = 0.2,                # Only 20% of peptides
  n_cores = 8
)

# ════════════════════════════════════════════════════════════
# MEDIUM RUN (~30-60 minutes) - Good balance
# ════════════════════════════════════════════════════════════
results_medium <- run_control_qc_elution(
  file_paths = files,
  ctrl_condition = "ctrl",
  output_dir = "C:/Users/ailaria/Projects/PCM/replicates comparison/control_qc_medium_finalChecks",
  methods = c("top5", "median", "iBAQ", "maxLFQ"), # No maxLFQ = MUCH faster!
  run_elution_analysis = TRUE,
  fit_gaussian = FALSE,             # Weighted mean (fast)
  sample_frac = 1.0,                # All peptides
  n_cores = 8
)

# ════════════════════════════════════════════════════════════
# FULL RUN (hours) - Only if you need Gaussian R² values
# ════════════════════════════════════════════════════════════
results_full <- run_control_qc_elution(
  file_paths = files,
  ctrl_condition = "ctrl",
  output_dir = "C:/Users/ailaria/Projects/PCM/replicates comparison/control_qc_full_finalChecks",
  fit_gaussian = TRUE,              # Full Gaussian (slow!)
  sample_frac = 1.0,
  n_cores = 8
)


# View results
results$comparison                          # Aggregation method comparison
results$elution_analysis$position_correlation  # Cross-experiment position correlations
results$elution_analysis$replicate_cv_summary  # Replicate consistency by experiment
results$elution_analysis$width_anova_pval      # Width batch effect test
```

# BATCH CORRECTION
```{r}

source("C:/Users/ailaria/Projects/aSYN_pipelinePCM/SEC_batch_correction/cofrac_batch_correction_v2.R")
results_batch_correction_10h <- run_cofrac_correction("C:/Users/ailaria/Projects/aSYN_pipelinePCM/20260304_IAA003_aSYN_10h_BRs_GFP_InTox_research_Report_BGS Factory Report (Normal).tsv", reference_replicate = "BR2", output_dir = "C:/Users/ailaria/Projects/asYN_pipelinePCM/SEC_batch_correction/corrected_BR2ref_10h",
  save_tables         = TRUE,
  save_plots          = TRUE)

results_batch_correction_10h$shifts

results_batch_correction_10h$plots$global_profiles



```

